### 2025-01-31 products/manuscript/Manuscript.qmd
Project Part 1

---
title: "Utilizing Google's Mobility Report to predict outbreak dynamics"
author: "Vincent Nguyen"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

-   Vincent Nguyen$^{1}$ (ORCID: 0000-0000-1234-5678)

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some\@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

Understanding the spread of infectious diseases is crucial for public health preparedness and response. Disease infection dynamics are shaped by multiple factors, including population movement, the pathogen itself, and public health interventions. Mobility is considered a key driver of human interactions and thus a primary vehicle for transmission potential. The COVID-19 pandemic provides a great example of how changes in mobility, driven by policies and public health measures, influenced infection rates.

## Description of data and data source

This project explores mobility dynamics, as measured by [Google's Community Mobility Reports](https://www.google.com/covid19/mobility/), in an infectious disease context to understand and forecast disease spread. By integrating time-series infection data (provided by [Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19)), the analysis aims to simulate disease outbreak scenarios and provide insights into optimal intervention strategies for mitigating future outbreaks.\
\
Google's Community Mobility Reports are anonymized insights collected from Google products, such as maps, that detail population movement trends across various categories, such as retail stores, workplaces, residencies, and more.\
\
The COVID-19 Data Repository by the Center for Systems Science and Engineering at Johns Hopkins University is a comprehensive data set that tracks global COVID-19 cases, recoveries, and deaths. It was updated regularly and compiles data from various official sources to support research.\
\
\
\

## Questions/Hypotheses to be addressed

How have mobility dynamics influenced the spread of COVID-19 in the U.S. during 2020-2022?\
\
Can mobility data, in combination with COVID-19 case counts, be used to forecast future infection trends?

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data acquisition

*COVID-19 case data was provided by the Center of Systems and Science and Engineering at JHU and mobility data was provided by Google's Community Mobility Report.*

## Data import and cleaning

Since the pandemic spanned across 3 years, data on mobility was collected from 2020 to 2022. The data on mobility is quite extensive and is stratified by county, resulting in the large amount of mobility. The mobility data measures changes in mobility patterns across different key areas, such as retail, workplace, residential, and more.

The time series data has 3342 observations of 1154 variables. Each row represents a different city. The columns record cases as days progress and locational data like longitude and latitude.

```{r}

# importing data
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))

time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))

```

## Statistical analysis

Possibly utilize the EpiModeling package to understand infectious disease dynamics based on the analysis and then simulate future outbreaks? I am not actually sure if this is a feasible use of the data.

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-02-20 code/eda-code/eda.qmd
Project Work

---
title: "An example exploratory analysis script"
date: "2024-02-07"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
```

Load the data.

```{r}
# importing data
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))
combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)

time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))

chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))

```

# Data exploration through graphs

Showing a bit of code to produce and save a summary table.

```{r}
# graph covid cases overtime

# Select only the date columns (starting from the 12th column)
date_columns_cases <- names(time_series_covid)[12:ncol(time_series_covid)]

# Ensure column names are correctly formatted as dates (removing 'X' from the date string)
date_fixed_cases <- as.Date(gsub("X", "", date_columns_cases), format = "%m.%d.%y")  # Fixed missing parenthesis

# Sum case counts across all locations for each date
total_cases <- colSums(time_series_covid[, date_columns_cases, drop = FALSE], na.rm = TRUE)

# Convert to a tidy data frame
total_cases_df <- data.frame(Date = date_fixed_cases, Cases = total_cases)

# Plot the total case counts over time with formatted y-axis labels without using 'scales'
case_count_graph <- ggplot(total_cases_df, aes(x = Date, y = Cases)) +
  geom_line(color = "blue") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()

print(case_count_graph)
```

```{r}
# Calculate the daily new cases by subtracting previous day's total from the current day's total
new_cases <- c(0, diff(total_cases))  # diff() gives the difference between consecutive days; first day is set to 0

# Convert to a tidy data frame for new cases
new_cases_df <- data.frame(Date = date_fixed_cases, New_Cases = new_cases)

# Plot the new cases (incidence) over time with formatted y-axis labels
incidence_graph <- ggplot(new_cases_df, aes(x = Date, y = New_Cases)) +
  geom_line(color = "red") + 
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "New Cases per Day") +
  theme_minimal()

print(incidence_graph)
```

```{r}
# Filter for U.S. data (adjust based on actual column name for region/country)
us_mobility_data <- combined_mobility_data %>%
  filter(country_region == "United States", sub_region_1 == "")

# Ensure the Date column is in Date format
us_mobility_data$Date <- as.Date(us_mobility_data$date, format = "%Y-%m-%d")  # Adjust if needed

# Calculate the 7-day moving average for each mobility type
us_mobility_data <- us_mobility_data %>%
  mutate(
    grocery_and_pharmacy_avg = rollmean(grocery_and_pharmacy_percent_change_from_baseline, 7, fill = NA, align = "right"),
    transit_stations_avg = rollmean(transit_stations_percent_change_from_baseline, 7, fill = NA, align = "right"),
    workplaces_avg = rollmean(workplaces_percent_change_from_baseline, 7, fill = NA, align = "right"),
    residential_avg = rollmean(residential_percent_change_from_baseline, 7, fill = NA, align = "right")
  )

# Plot the 7-day moving averages for each mobility type
mobility_graph <- ggplot(us_mobility_data, aes(x = Date)) +
  geom_line(aes(y = grocery_and_pharmacy_avg, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = transit_stations_avg, color = "Transit Stations")) +
  geom_line(aes(y = workplaces_avg, color = "Workplaces")) +
  geom_line(aes(y = residential_avg, color = "Residential")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
    x = "Date",
    y = "7-Day Moving Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_minimal()

print(mobility_graph)
```
```{r}

```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

Histogram plots for the continuous outcomes.

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results", "figures", "height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2025-02-20 products/manuscript/Manuscript.qmd
Project Work

@@
 This project explores mobility dynamics, as measured by [Google's Community Mobility Reports](https://www.google.com/covid19/mobility/), in an infectious disease context to understand and forecast disease spread.
-By integrating time-series infection data (provided by [Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19)), the analysis aims to simulate disease outbreak scenarios and provide insights into optimal intervention strategies for mitigating future outbreaks.\
+By integrating time-series infection data (provided by [Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19)), the analysis aims to simulate disease outbreak scenarios and provide insights into optimal intervention strategies for mitigating future outbreaks.
+Additionally, the [County Health Ranking](https://www.countyhealthrankings.org/) is used to assess these dynamics within the context of a county's health care access and performance.\
 \
@@
 It was updated regularly and compiles data from various official sources to support research.\
-\
+
+The County Health Ranking is a data set that provides a comprehensive overview of various health factors and outcomes across U.S.
+counties.
+Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more.\
 \
@@
 Can mobility data, in combination with COVID-19 case counts, be used to forecast future infection trends?
+
+How do socio-economic and health factors influence the relationship between mobility patterns and COVID-19 infection rates across U.S. counties?
+
+Investigate if counties with similar health rankings but different mobility trends (or vice versa) experienced different outbreak severities, potentially revealing the effectiveness of local interventions \[reword\]
+
+Examine the time lag between changes in mobility and subsequent changes in infection rates \[reword\]
 
@@
 time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))
+
+chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))
 

### 2025-02-21 code/eda-code/eda.qmd
Project Part 2
Might be a bit bare, however, I took caution to fleshing out too much if my topic needs to be adjusted. Right now, I have considered adding in the CDC Vaccination Hesitancy survey (possibly to map out and compare with mobility dynamics) or the County Health Ranking (possibly to explore how communities of differing SES dealt with cases in terms of mobility).

@@
 ---
-title: "An example exploratory analysis script"
-date: "2024-02-07"
+title: "Exploratory Data Analysis"
+date: "2025-02-21"
 output: html_document
@@
 This Quarto file loads the cleaned data and does some exploring.
-
-I'm only showing it the way where the code is included in the file. 
-As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.
-
-But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)
-
-Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
-And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.
-
-Also note that while here I split cleaning and exploring, this is iterative.
-You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it.
-In general, as you explore, you'll find things that need cleaning.
-As you clean, you can explore more.
-Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file.
-Or split things in any other logical way.
-
-As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-Figures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data.
-
-Start by exploring one variable at a time.
-Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
-
 # Setup
@@
 ```{r}
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
+#load needed packages. make sure they are installed
+library(here) 
 library(dplyr)
@@
 library(zoo)
+library(tidyr)
+library(tigris)
+library(sf)
 ```
 
-Load the data.
+Load the data from processingfile-v1 file.
 
 ```{r}
-# importing data
-mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
-mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
-mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))
-combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)
 
-time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))
-
-chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))
+# processed data
+processeddata <- readRDS(here("data", "processed-data", "processeddata.rds"))
 
@@
 
-# Data exploration through graphs
+Creation of a heatmap showing total COVID 19 cases on the last observed date, 10/15/2022.
+I also created a heat map showing mobility changes, with hotter indicating more mobility to retail and recreational spots.
+Interestingly, the two maps seem to correlate even though I used total COVID-19 cases rather than new ones.
+```{r}
+# Load the state-level shapefile 
+us_map_states <- tigris::states(cb = TRUE) %>%
+  st_as_sf()  # Convert the shapefile to a Simple Feature object
 
-Showing a bit of code to produce and save a summary table.
+# Clean the state names in the shapefile
+us_map_clean <- us_map_states %>%
+  mutate(NAME = tolower(NAME))  # Convert state abbreviation to lowercase
 
+# Filter your dataset to get data for the last recorded day and group by state
+last_day <- max(processeddata$date)
+data_last_day <- processeddata %>%
+  filter(date == last_day) %>%
+  group_by(state) %>%
+  summarise(total_cases = sum(total_cases, na.rm = TRUE),  # Sum of total cases
+            retail_and_recreation_percent_change_from_baseline = 
+              mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE)) %>%  # Average mobiltiy changes
+  mutate(state = tolower(state))  # Convert state names to lowercase
+
+# Join the shapefile with the COVID-19 data
+us_map_data <- us_map_clean %>%
+  left_join(data_last_day, by = c("NAME" = "state"))
+
+# plot total COVID-19 cases to USA map on 10/15/2022
+cases_map <- ggplot(us_map_data) +
+  geom_sf(aes(fill = total_cases), color = "white") +
+  scale_fill_viridis_c(option = "plasma", trans = "log", labels = NULL) + 
+  labs(title = paste("Heatmap of Total COVID-19 Cases in the U.S. on", last_day),
+       fill = "Hotter = More Cases") +
+  theme_minimal() +
+  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
+
+print(cases_map)
+
+figure_file = here("results","figures","casemap.png")
+ggsave(filename = figure_file, plot=cases_map)
+
+# Create the heatmap for retail and recreation mobility changes
+mobility_map <- ggplot(us_map_data) +
+  geom_sf(aes(fill = retail_and_recreation_percent_change_from_baseline), color = "white") +  
+  scale_fill_viridis_c(option = "inferno", trans = "identity") +
+  labs(title = paste("Heatmap of Retail and Recreation Mobility Changes in the U.S. on", last_day),
+       fill = "Mobility Change") +
+  theme_minimal() +
+  theme(
+        legend.spacing.y = unit(1, "cm"),  
+        legend.box.spacing = unit(1, "cm"),  
+        legend.key.size = unit(1, "cm")) +  
+  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
+
+print(mobility_graph)
+
+# save graph
+figure_file = here("results","figures","mobilitymap.png")
+ggsave(filename = figure_file, plot=mobility_map)
+```
+
+This graph aims to plot total cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all total cases.
 ```{r}
-# graph covid cases overtime
+# Create new data frame to have cases per day instead of state/county
+total_cases_df <- processeddata %>%
+  group_by(date) %>%
+  summarise(Cases = sum(total_cases, na.rm = TRUE)) %>%
+  ungroup()
 
-# Select only the date columns (starting from the 12th column)
-date_columns_cases <- names(time_series_covid)[12:ncol(time_series_covid)]
-
-# Ensure column names are correctly formatted as dates (removing 'X' from the date string)
-date_fixed_cases <- as.Date(gsub("X", "", date_columns_cases), format = "%m.%d.%y")  # Fixed missing parenthesis
-
-# Sum case counts across all locations for each date
-total_cases <- colSums(time_series_covid[, date_columns_cases, drop = FALSE], na.rm = TRUE)
-
-# Convert to a tidy data frame
-total_cases_df <- data.frame(Date = date_fixed_cases, Cases = total_cases)
-
-# Plot the total case counts over time with formatted y-axis labels without using 'scales'
-case_count_graph <- ggplot(total_cases_df, aes(x = Date, y = Cases)) +
+# Plot the total case counts over time
+case_count_graph <- ggplot(total_cases_df, aes(x = date, y = Cases)) +
   geom_line(color = "blue") +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  
   labs(title = "Total COVID-19 Cases Over Time in the U.S.",
@@
 print(case_count_graph)
+
+# save graph
+figure_file = here("results","figures","casecountplot.png")
+ggsave(filename = figure_file, plot=case_count_graph)
 ```
+This graph aims to plot new cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all new cases.
+```{r}
+# data frame to sum all cases on a given day for every location
+new_cases_df <- processeddata %>%
+  group_by(date) %>%
+  summarise(Cases = sum(new_cases, na.rm = TRUE)) %>%
+  ungroup()
 
-```{r}
-# Calculate the daily new cases by subtracting previous day's total from the current day's total
-new_cases <- c(0, diff(total_cases))  # diff() gives the difference between consecutive days; first day is set to 0
-
-# Convert to a tidy data frame for new cases
-new_cases_df <- data.frame(Date = date_fixed_cases, New_Cases = new_cases)
-
-# Plot the new cases (incidence) over time with formatted y-axis labels
-incidence_graph <- ggplot(new_cases_df, aes(x = Date, y = New_Cases)) +
+# Plot the new cases over time
+incidence_graph <- ggplot(new_cases_df, aes(x = date, y = Cases)) +
   geom_line(color = "red") + 
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
+  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
   labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
@@
 print(incidence_graph)
+
+# save graph
+figure_file = here("results","figures","incidenceplot.png")
+ggsave(filename = figure_file, plot=incidence_graph)
+
 ```
+I aim to plot out the changes in mobility for each category as time progresses.
+Originally, it plotted out everyday, however, the line was far too jagged so I applied a weekly rolling average instead.
+I also had to create a dataframe to average values for a given day for every location.
+```{r}
 
-```{r}
-# Filter for U.S. data (adjust based on actual column name for region/country)
-us_mobility_data <- combined_mobility_data %>%
-  filter(country_region == "United States", sub_region_1 == "")
-
-# Ensure the Date column is in Date format
-us_mobility_data$Date <- as.Date(us_mobility_data$date, format = "%Y-%m-%d")  # Adjust if needed
-
-# Calculate the 7-day moving average for each mobility type
-us_mobility_data <- us_mobility_data %>%
+# Create dataframe of average mobility for everywhere in the US per date
+us_average_mobility <- processeddata %>%
+  group_by(date) %>%
+  summarise(
+    total_mobility_changes = mean(average_mobility, na.rm = TRUE),
+    transit_mean = mean(transit_stations_percent_change_from_baseline, na.rm = TRUE),
+    parks_mean = mean(parks_percent_change_from_baseline, na.rm = TRUE),
+    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
+    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
+    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
+    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  arrange(date) %>%  # Ensure dates are sorted
   mutate(
-    grocery_and_pharmacy_avg = rollmean(grocery_and_pharmacy_percent_change_from_baseline, 7, fill = NA, align = "right"),
-    transit_stations_avg = rollmean(transit_stations_percent_change_from_baseline, 7, fill = NA, align = "right"),
-    workplaces_avg = rollmean(workplaces_percent_change_from_baseline, 7, fill = NA, align = "right"),
-    residential_avg = rollmean(residential_percent_change_from_baseline, 7, fill = NA, align = "right")
+    transit_mean_7day = rollmean(transit_mean, k = 7, fill = NA, align = "right"),
+    parks_mean_7day = rollmean(parks_mean, k = 7, fill = NA, align = "right"),
+    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
+    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
+    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
+    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
   )
 
-# Plot the 7-day moving averages for each mobility type
-mobility_graph <- ggplot(us_mobility_data, aes(x = Date)) +
-  geom_line(aes(y = grocery_and_pharmacy_avg, color = "Grocery & Pharmacy")) +
-  geom_line(aes(y = transit_stations_avg, color = "Transit Stations")) +
-  geom_line(aes(y = workplaces_avg, color = "Workplaces")) +
-  geom_line(aes(y = residential_avg, color = "Residential")) +
+# Plot the 7-day averages for each mobility type
+mobility_graph <- ggplot(us_average_mobility, aes(x = date)) +
+  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
+  geom_line(aes(y = transit_mean_7day, color = "Transit Stations")) +
+  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
+  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
+  geom_line(aes(y = parks_mean_7day, color = "Recreational")) +
   labs(
@@
     x = "Date",
-    y = "7-Day Moving Average of Percent Change in Mobility",
+    y = "7-Day Average of Percent Change in Mobility",
     color = "Mobility Type"
@@
 print(mobility_graph)
+
+# save graph
+figure_file = here("results","figures","mobilitygraph.png")
+ggsave(filename = figure_file, plot=mobility_graph)
 ```
-```{r}
-
-```
-
-We are saving the results to the `results` folder.
-Depending on how many tables/figures you have, it might make sense to have separate folders for each.
-And/or you could have separate folders for exploratory tables/figures and for final tables/figures.
-Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.
-
-# Data exploration through figures
-
-Histogram plots for the continuous outcomes.
-
-Height first.
-
-```{r}
-p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
-plot(p1)
-figure_file = here("results", "figures","height-distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-```
-
-Now weights.
-
-```{r}
-p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
-plot(p2)
-figure_file = here("results", "figures","weight-distribution.png")
-ggsave(filename = figure_file, plot=p2) 
-```
-
-Now height as function of weight.
-
-```{r}
-p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
-plot(p3)
-figure_file = here("results", "figures", "height-weight.png")
-ggsave(filename = figure_file, plot=p3) 
-```
-
-Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.
-
-```{r}
-p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
-plot(p4)
-figure_file = here("results", "figures", "height-weight-stratified.png")
-ggsave(filename = figure_file, plot=p4) 
-```
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.

### 2025-02-21 code/processing-code/processingfile-v1.qmd
Project Part 2
Might be a bit bare, however, I took caution to fleshing out too much if my topic needs to be adjusted. Right now, I have considered adding in the CDC Vaccination Hesitancy survey (possibly to map out and compare with mobility dynamics) or the County Health Ranking (possibly to explore how communities of differing SES dealt with cases in terms of mobility).

---
title: "Cleaning Script"
author: "Vincent Nguyen"
date: "2025-02-21"
output: html_document
---

# Setup

Load needed packages. make sure they are installed.

```{r}
# load needed packages
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
```

# Data loading

```{r}
# importing data

# google data is in three parts
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))
combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)

# jhk covid time series cases
time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))

# county health ranking
chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))

```

After visually inspecting the mobility data, I noticed that many mobility changes were missing. The following code seeks to quantify this.
```{r}

# After visually inspecting the data, I noticed that a lot of mobility changes were missing.
sapply(combined_mobility_data[c("transit_stations_percent_change_from_baseline", 
                         "parks_percent_change_from_baseline", 
                         "retail_and_recreation_percent_change_from_baseline", 
                         "grocery_and_pharmacy_percent_change_from_baseline",
                         "workplaces_percent_change_from_baseline", 
                         "residential_percent_change_from_baseline")], function(x) sum(is.na(x)))

```
# Cleaning and Wrangling

By inspecting the data, some problems become apparent. AS mentioned above, the Mobility Report is missing a tremendous amount of values. The Time Series data and Mobility Report are structured in entirely different ways. Joining the two will be difficult without mutating one. I chose to change the Time Series data.

Next, we also can see that the naming scheme of the columns differ so we need to standardize that as well.

Lastly, both datasets do not have the same range in dates

```{r}

# Pivot the time series data and change names around
cases_long <- time_series_covid %>%
  pivot_longer(
    cols = starts_with("X"),  
    names_to = "date",
    values_to = "total_cases"
  ) %>%
  mutate(date = gsub("^X", "", date),  # Remove leading x for dates to join data sets in a bit
         date = gsub("\\.", "/", date),  # Replace dots with slashes
         date = as.Date(date, format = "%m/%d/%y"))  # Use "%y" instead of "%Y"

# Standardize county & state names (convert to lowercase to avoid mismatches)
cases_long <- cases_long %>%
  rename(county = Admin2, state = Province_State) %>%
  mutate(county = tolower(county), state = tolower(state))

mobility_df <- combined_mobility_data %>%
  rename(county = sub_region_2, state = sub_region_1) %>%
  mutate(county = tolower(county), 
         state = tolower(state),
         county = gsub(" county$", "", county),  # Remove "county" from county name
         date = as.Date(date, format = "%Y-%m-%d"))  # Correct format for "YYYY-MM-DD"

# Filter cases_long and mobility_df based on the date range
cases_long_filtered <- cases_long %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

mobility_df_filtered <- mobility_df %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

# Merge the data sets together
merged_df <- cases_long_filtered %>%
  inner_join(mobility_df_filtered, by = c("county", "state", "date"))

# Finally, remove any rows missing mobility changes to ensure consistency. We also have a lot of data values so I deemed it permissible.
merged_df_clean <- merged_df %>%
  drop_na(c(transit_stations_percent_change_from_baseline, parks_percent_change_from_baseline, retail_and_recreation_percent_change_from_baseline, grocery_and_pharmacy_percent_change_from_baseline,workplaces_percent_change_from_baseline, residential_percent_change_from_baseline))

# Additionally, add a column about new cases for further analysis down the line.
merged_df_clean <- merged_df_clean %>%
  arrange(state, county, date) %>%  # Sort (Need to make sure calculations occur between the correct states/counties and in order)
  group_by(state, county) %>%  # Group by state and county to calculate new cases within each county
  mutate(new_cases = c(NA, diff(total_cases))) %>%  # Calculate new cases
  ungroup()  # Ungroup after the calculation
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- merged_df_clean
```

# Save data

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2025-02-21 products/manuscript/Manuscript.qmd
Project Part 2
Might be a bit bare, however, I took caution to fleshing out too much if my topic needs to be adjusted. Right now, I have considered adding in the CDC Vaccination Hesitancy survey (possibly to map out and compare with mobility dynamics) or the County Health Ranking (possibly to explore how communities of differing SES dealt with cases in terms of mobility).

@@
 ---
-title: "Utilizing Google's Mobility Report to predict outbreak dynamics"
+title: "Utilizing Google's Mobility Report to understand Infectious Disease Dynamics"
 author: "Vincent Nguyen"
@@
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
 This uses MS Word as output format.
@@
 
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
-```
+Some text was left from the template in order to adjust to any feedback given. The manuscript will be more fleshed out once the final project idea is approved.
 
@@
 
-$*$ These authors contributed equally to this work.
-
-$\land$ Corresponding author: some\@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
+$\land$ Corresponding author: vln27447\@uga.edu
 
@@
 counties.
-Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more.\
+Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more.
+\[Not sure about the inclusion of this\]
+
+The CDC Vaccination hesitancy survey is a possible data set I could use to make the analysis more complex.\
 \
@@
 
-Investigate if counties with similar health rankings but different mobility trends (or vice versa) experienced different outbreak severities, potentially revealing the effectiveness of local interventions \[reword\]
+Investigate if counties with similar health rankings but different mobility trends experienced different outbreak severities, potentially revealing the effectiveness of local interventions
 
-Examine the time lag between changes in mobility and subsequent changes in infection rates \[reword\]
+Examine ifthere was lag time between changes in mobility and subsequent changes in infection rates
 
@@
 Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
 
@@
 We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
 
@@
 ```{r}
+library(here)
 
@@
 
-Possibly utilize the EpiModeling package to understand infectious disease dynamics based on the analysis and then simulate future outbreaks?
-I am not actually sure if this is a feasible use of the data.
-
 {{< pagebreak >}}
@@
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+To gain a clearer perspective on the overall scale of the pandemic, we visualized the total number of COVID-19 cases over time.
+This plot provides a overview of how case counts evolved.The graph reveals a large spike in cases in the beginning of 2022.
+This does correspond with figures further.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: casecountplot
+#| fig-cap: "Total COVID-19 Cases overtime"
 #| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
+knitr::include_graphics(here("results","figures","casecountplot.png"))
 ```
+
+To gain a clearer perspective on the overall scale of the pandemic, we visualized new caes of COVID-19 over time.
+This plot provides a overview of how case counts and infection evolved.The graph reveals a large spike in cases in the beginning of 2022 which corresponds with the total case count graph.
+
+```{r}
+#| label: incidenceplot
+#| fig-cap: "Daily New COVID-19 Cases overtime in the U.S."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","incidenceplot.png"))
+```
+
+To gain a deeper understanding of how mobiltiy patterns shifted throughout the pandemic, weekly changes in mobility were visualized.
+This plot highlights fluctuations in movement across various sectors.
+Mobility was increased towards residential and recreational areas which warrants further exploration.
+This can help offer insight into how behavior towards COVID-19 restriction interventions evolved over time.
+
+```{r}
+#| label: mobiltiygraph
+#| fig-cap: "Percent Changes in Mobility per week in the U.S."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","mobilitygraph.png"))
+```
+
+As hypothesized, an rise in COVID-19 case count is expected to lead to a decline in mobility to non-residential locations, as individuals alter their behavior in resposne to increasing risks.
+To explore this idea, total case count and changes in mobility were visualzied across the United States.\
+\
+
+```{r}
+#| label: casemap
+#| fig-cap: "Total COVID-19 Cases on 10/15/2022 mapped to the U.S."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","casemap.png"))
+```
+
+```{r}
+#| label: mobilitymap
+#| fig-cap: "Average of mobility changes on 10/15/2022 mapped to the U.S."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","mobilitymap.png"))
+```
+
+Analysis of the mapped data reveals a trend: states with higher case counts generally experienced more significant decreases in mobility.
+This suggests a correlation between rising COVID-19 case count and a greater reduction in movement to non-residential locations.
 
@@
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
 ## Full analysis
@@
 Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
 

### 2025-02-21 code/processing-code/processingfile-v1.qmd
Woops forgot to close Rstudio

@@
   ungroup()  # Ungroup after the calculation
+
+# Create column for mean average mobiltiy change for that given day and location. I omitted mobility changes to residencies because that may be more of an indication of who is at home.
+merged_df_clean <- merged_df_clean %>%
+  arrange(state, county, date) %>%  # Sort to ensure correct calculations
+  group_by(state, county) %>%  # Group by state and county
+  mutate(average_mobility = rowMeans(
+    cbind(transit_stations_percent_change_from_baseline, 
+          parks_percent_change_from_baseline, 
+          retail_and_recreation_percent_change_from_baseline, 
+          grocery_and_pharmacy_percent_change_from_baseline, 
+          workplaces_percent_change_from_baseline), 
+    na.rm = TRUE)) %>%
+  ungroup()
+
 ```

### 2025-03-14 code/processing-code/processingfile-v1.qmd
Part 3 of the project
Done modeling and some work on manuscript, Need to define workflow for analysis and finish on supplementary analyses.

@@
 library(tidyr)
+library(lubridate)
+library(readxl)
 ```
@@
 
+# county population information
+ch_pop_data <- read_excel(here::here("data", "raw-data", "co-est2023-pop.xlsx"), skip = 3)
+```
+
+```{r}
+# Remove the second column
+ch_pop_data <- ch_pop_data[, -2]
+
+ch_pop_data <- ch_pop_data %>%
+  pivot_longer(
+    cols = c(`2020`,`2021`, `2022`, `2023`),    # Specify the year columns
+    names_to = "year",                    # Create a new column for the years
+    values_to = "population_count"              # Create a new column for population counts
+  )
+
+# Fix the first column
+ch_pop_data <- ch_pop_data %>%
+  mutate(
+    county = str_remove(`...1`, "^\\."),            
+    county = str_remove(county, "County"),     
+    county = str_to_lower(county),                  
+    state = str_extract(county, "(?<=,).*"),        
+    county = str_remove(county, ",.*")              
+  ) %>%
+  mutate(
+    state = str_to_lower(str_trim(state))           
+  )
+
+# Change year column to be double for joining later
+ch_pop_data$year <- as.double(ch_pop_data$year)
+
+# Clean names to prepare for joining
+ch_pop_data <- ch_pop_data %>%
+  mutate(
+    county = str_trim(str_to_lower(county)),  
+    state = str_trim(str_to_lower(state))     
+  )
+
 ```
@@
 # Cleaning and Wrangling
+
+This section looks through the CHR data
+
+```{r}
+# First need to filter out the data for what years we are interested in
+unique(chr_data_2024$yearspan)
+
+# Based on this, we are most interested in 2020-2022 because of its overlap with the other data.
+chr_data_filtered <- filter(chr_data_2024, yearspan %in% c("2020", "2021", "2022"))
+
+# Change yearspan to be a double. This is important for joining all the data frames later.
+chr_data_filtered$yearspan <- as.double(chr_data_filtered$yearspan)
+
+# Now we need to decide what meausre interest us the most
+unique(chr_data_filtered$measurename)
+
+# We chose uninsured adults as an indicator of access to healthcare (this is listed in the website)
+chr_data_filtered <- filter(chr_data_filtered, measurename == "Uninsured adults")
+
+# rename the counties to prep for dataframe joining
+chr_data_filtered <- chr_data_filtered %>%
+  mutate(
+    county = tolower(county),  # Convert 'county' to lowercase
+    county = gsub(" county", "", county), # removes the word county
+    uninsured_rate = (numerator / denominator) * 100000) # uninsured rate per 100,000
+
+# Check if counties have same names across different staes
+duplicates <- chr_data_filtered %>%
+  group_by(county, state) %>%
+  tally() %>%
+  filter(n > 1)
+
+# View the duplicates and looks like there is... need to deal with this if I want to join the data with the other data
+print(duplicates)
+
+# Change abbreviated state names into the full name so I can join them easier
+
+# Create a mapping for state abbreviations to full names and convert the full names to lowercase
+state_mapping <- data.frame(
+  state_abbr = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
+  state_full = tolower(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"))
+)
+
+# Convert state abbreviation to full name in chr_data_2024
+chr_data_filtered <- chr_data_filtered %>%
+  left_join(state_mapping, by = c("state" = "state_abbr"))
+
+```
 
@@
 
+merged_df_clean <- merged_df_clean %>%
+  mutate(year = year(date)) %>%
+  left_join(chr_data_filtered %>% select(state_full, county, uninsured_rate, yearspan), 
+            by = c("state" = "state_full", "county" = "county", "year" = "yearspan"))
+
+merged_df_clean <- merged_df_clean %>%
+  left_join(
+    ch_pop_data %>% 
+      select(year, population_count, county, state),
+    by = c("state" = "state", "county" = "county", "year" = "year")
+  )
 ```

### 2025-03-14 products/manuscript/Manuscript.qmd
Part 3 of the project
Done modeling and some work on manuscript, Need to define workflow for analysis and finish on supplementary analyses.

@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
+bibliography: ../../assets/MADA.bib
 csl: ../../assets/american-journal-of-epidemiology.csl
@@
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
-Some text was left from the template in order to adjust to any feedback given. The manuscript will be more fleshed out once the final project idea is approved.
-
 **Authors**
 
--   Vincent Nguyen$^{1}$ (ORCID: 0000-0000-1234-5678)
+-   Vincent Nguyen$^{1}$
 
@@
 
-## General Background Information
+The emergence of the novel corona-virus SARS-CoV-2 caused one of the most significant global health crises in modern history.
+First identified in Wuhan, China, in late 2019, the virus’s transmissibility and globalism led to its rapid spread worldwide, to leading the World Health Organization to declare it a pandemic on March 11, 2020.
+Just two days later, on March 13th, 2020, the Trump Administration declared a nationwide emergency due to the COVID-19 pandemic, marking the biggest virus outbreak since \_\_\_.
+In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lockdown procedures (CDC 2024)[@centersfordiseasecontrolandprevention_2024_covid19].
 
-Understanding the spread of infectious diseases is crucial for public health preparedness and response.
-Disease infection dynamics are shaped by multiple factors, including population movement, the pathogen itself, and public health interventions.
-Mobility is considered a key driver of human interactions and thus a primary vehicle for transmission potential.
-The COVID-19 pandemic provides a great example of how changes in mobility, driven by policies and public health measures, influenced infection rates.
+\[insert stats about pandemic\]
+
+### 2.2 Lockdown and Social Distancing
+
+Nonpharmacological interventions are among the measures public health officials utilize to help curb the transmission of infectious diseases.
+In lieu of medication or vaccinations, these interventions can be a cost-effective and population-based solution.
+These interventions include the use of isolation, the subdivision of an infected individual from the general population, and social distancing, a form of voluntary or mandated physical distancing from others.
+Another common measure is the use of quarantine, which is the restriction of infected or at-risk individuals from traveling to help minimize the risk of healthy individuals; it is intended that the period of transmissibility has passed before lifting the quarantine (Jacobsen and Jacobsen 2020)[@jacobsen_2020_statewide].
+
+Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lockdowns.
+Lockdowns involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
+By April 3rd, 2020, just a few weeks after California’s state-wide mandate, 41 states joined the directive in restricting mobility (Mervosh et al.
+2020)[@mervosh_2020_see].
+Many polictians and citizens questioned the measure citing a lack of evidence for its efficacy; looking at the Wikipedia page shows a lack of prior history implementing lockdowns (Zeleny et al.
+2021)[@zeleny_2021_a].
+
+### 2.3 Mobility Dynamics
+
+According to Paltra et al.
+2021, mobility reductions on COVID-19 can be quadratic in reducing the transmission.
+For example, the authors state a reduction of out-of-home duration to 70% can result in a transmission rate of 0.49.
+\[finish\]
 
@@
 \[Not sure about the inclusion of this\]
-
-The CDC Vaccination hesitancy survey is a possible data set I could use to make the analysis more complex.\
-\
-\
 

### 2025-03-14 code/analysis-code/statistical-analysis.R

@@ -8,13 +8,21 @@
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
+library(glmnet)
+library(MASS)
 
 #path to data
 #note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","processeddata.rds")
 
 #load data. 
-mydata <- readRDS(data_location)
+data <- readRDS(data_location)
+
+# NOTES TO SELF
+# - change to QMD file (personally prefer)
+# - work on other analysis (lag-time & quarantine fatigue? more research)
+# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
+
 
 
 ######################################
@@ -22,35 +30,147 @@ mydata <- readRDS(data_location)
 ######################################
 
 ############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
+#### Initial model fitting
 
-lmfit1 <- lm(Height ~ Weight, mydata)  
+# Some quick data fixing before analysis as some models cannot handle negatives (contextually ok, more interested in NEW cases)
+data_clean <- data
+data_clean$new_cases[data_clean$new_cases < 0] <- 0
 
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
 
-#look at fit results
-print(lmtable1)
+# first Negative Binomial Model
+nb_model_1 <- glm.nb(new_cases ~ population_count + transit_stations_percent_change_from_baseline + 
+                     parks_percent_change_from_baseline + 
+                     retail_and_recreation_percent_change_from_baseline + 
+                     grocery_and_pharmacy_percent_change_from_baseline +
+                     workplaces_percent_change_from_baseline + 
+                     residential_percent_change_from_baseline +
+                     population_count, 
+                   data = data_clean)
 
-# save fit results table  
-table_file1 = here("results", "tables", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
+# View summary
+summary(nb_model_1)
 
-############################
-#### Second model fit
-# fit linear model using height as outcome, weight and gender as predictor
+# Poisson regression model based on the same formula
+poisson_model_1 <- glm(new_cases ~ transit_stations_percent_change_from_baseline + 
+                       parks_percent_change_from_baseline + 
[…]

### 2025-04-04 code/analysis-code/statisitical-analysis.qmd
Part 4
Hefty changes to my project. First, narrowing the project's scope to just Georgia and GA counties. More modeling work; lot of exploration resulting in unsuccessful models and deletions unfortunately. Also, redid some exploratory analysis but working on the figures.

---
title: "Analysis Script"
author: "Vincent Nguyen"
date: "2025-02-21"
output: html_document
---

# Setup

Load needed packages. make sure they are installed.

```{r}
library(ggplot2) 
library(broom) 
library(here) 
library(glmnet)
library(MASS)
library(tidymodels)
library(dplyr)
library(rsample)
library(poissonreg)
library(parsnip)
library(future)
```

```{r}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

#load data. 
data <- readRDS(data_location)

# NOTES TO SELF
# - change to QMD file (personally prefer)
# - work on other analysis (lag-time & quarantine fatigue? more research)
# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
class(data$new_cases)
```

```{r}
# Some quick data fixing before analysis as some models cannot handle negatives (contextually ok, more interested in NEW cases)
data_clean <- data
data_clean$new_cases[data_clean$new_cases < 0] <- 0
data_clean$new_cases[is.na(data_clean$new_cases)] <- 0
data_clean$incidence_rate[is.na(data_clean$incidence_rate)] <- 0

data_clean <- na.omit(data_clean)

# In my intial analyses, there were an absurd amount of zeroes; this was most likely due to the data sets capturing days prior to the pandemic.
last_case_date <- as.Date("2022-01-01")
first_case_date <- as.Date("2020-03-02")

df_clean <- data_clean %>%
  filter(date >= first_case_date) %>%
  filter(date <= last_case_date) %>%
  group_by(county)

df_clean <- df_clean %>%
  as_tibble() %>%
  select(-geometry)

df_clean <- df_clean %>% 
  mutate(new_cases = as.integer(df_clean$new_cases))

class(df_clean$new_cases)
```

```{r}
# Intial Recipe setting
recipe <- recipe(new_cases ~ pop_density + transit_stations_percent_change_from_baseline + 
                     parks_percent_change_from_baseline + 
                     retail_and_recreation_percent_change_from_baseline + 
                     grocery_and_pharmacy_percent_change_from_baseline +
                     workplaces_percent_change_from_baseline + 
                     residential_percent_change_from_baseline +
                     health_factor_rank, data = data)

```

First, I created a linear regression model as the baseline standard. This model performs very poorly as it has a low R-squared, indicating low predictive power. The model does mark all variables of the model to be significant.Additionally, the AIC is very extremely high.

```{r}
# Create linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_wf <- workflow() %>%
  add_recipe(recipe)

lm_fit <- lm_wf %>%
  add_model(lm_spec) %>%
  fit(data = df_clean)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lm_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lm_preds<- predict(lm_fit, new_data = df_clean)

lm_metrics <- tibble(truth = df_clean$new_cases, predicted = lm_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lm_metrics
```

I utilized Lasso here to hopefully bring down the RMSE and bring up the rsquared. It seems Lasso has no effect on model performance as the data lacks multicolinearity and does not have irrelevant predictors. 
```{r}
# Set method for LASSO
lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet")

# Insert recipe
lasso_wf <- workflow() %>%
  add_recipe(recipe)

lasso_fit <- lasso_wf %>%
  add_model(lasso_spec) %>%
  fit(data = df_clean)

lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lasso_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lasso_preds<- predict(lasso_fit, new_data = df_clean)

lasso_metrics <- tibble(truth = df_clean$new_cases, predicted = lasso_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lasso_metrics
```

After the last two models performed poorly, the use of random forest seemed rational as this model can capture more complex relationships. 
```{r}

rngseed = 1234

forest_spec <- rand_forest(mode = "regression") %>%
  set_engine("ranger", seed = rngseed)

# Insert recipe
forest_wf <- workflow() %>%
  add_recipe(recipe)

# Create model with all predictors and random forest
forest_fit <- forest_wf %>%
 add_model(forest_spec) %>%
  fit(data = data)

forest_preds <- predict(forest_fit, new_data = data)
forest_preds <- tibble(truth = data$new_cases, predicted = forest_preds$.pred)
forest_rmse <- forest_preds %>%
  metrics(truth = truth, estimate = predicted)
print(forest_rmse)
```

```{r}
# Tuning forest
forest_spec <- rand_forest(
  mode = "regression", 
  trees = 300, 
  mtry = tune(), 
  min_n = tune()
) %>%
  set_engine("ranger")

# Create grid with set parameters
forest_grid <- grid_regular(
  mtry(range = c(1, 7)),  
  min_n(range = c(1, 21)),  
  levels = 7  
)

# Set seed
set.seed(1234)

# 5 fold crossvalidation repeated 5 times
# Note: Limitation of computing power so doing rolling window would not work
cv_folds <- vfold_cv(data, v = 5, repeats = 5)
plan(multisession, workers = 4)
# Tune the forest
forest_tune_res <- tune_grid(
  forest_wf %>% add_model(forest_spec),
  resamples = cv_folds,  
  grid = forest_grid,
  metrics = metric_set(rmse, rsq),
  control = control_grid(parallel_over = "everything")
)

# Plot results
autoplot(forest_tune_res)

forest_tune_res %>%
  collect_metrics()

best_params <- select_best(forest_tune_res, metric = "rmse")  # Or "rsq"
best_params

best_params$trees <- 500
```

```{r}
forest_spec <- rand_forest(
  mode = "regression",
  mtry = best_params$mtry,
  trees = best_params$trees,
  min_n = best_params$min_n
) %>%
  set_engine("ranger", seed = 1234)

forest_fit_final <- forest_wf %>%
  add_model(forest_spec) %>%
  fit(data = data)

# Get final predictions
forest_preds_final <- predict(forest_fit_final, new_data = data)
forest_preds_final <- tibble(truth = data$new_cases, predicted = forest_preds_final$.pred)

# Calculate performance metrics (RMSE, R², etc.)
forest_rmse <- forest_preds_final %>%
  metrics(truth = truth, estimate = predicted)

print(forest_rmse)
```

### 2025-04-04 code/eda-code/eda.qmd
Part 4
Hefty changes to my project. First, narrowing the project's scope to just Georgia and GA counties. More modeling work; lot of exploration resulting in unsuccessful models and deletions unfortunately. Also, redid some exploratory analysis but working on the figures.

@@
 
+processeddata <- processeddata %>%
+   mutate(incidence_rate = if_else(incidence_rate < 0, 0, incidence_rate))
 ```
@@
 Interestingly, the two maps seem to correlate even though I used total COVID-19 cases rather than new ones.
+
 ```{r}
-# Load the state-level shapefile 
-us_map_states <- tigris::states(cb = TRUE) %>%
-  st_as_sf()  # Convert the shapefile to a Simple Feature object
+# load GA shapefile
+ga_counties <- counties(state = "GA", cb = TRUE, class = "sf")
 
-# Clean the state names in the shapefile
-us_map_clean <- us_map_states %>%
-  mutate(NAME = tolower(NAME))  # Convert state abbreviation to lowercase
+ga_counties <- ga_counties %>%
+  mutate(NAME = tolower(NAME))
 
-# Filter your dataset to get data for the last recorded day and group by state
-last_day <- max(processeddata$date)
-data_last_day <- processeddata %>%
-  filter(date == last_day) %>%
-  group_by(state) %>%
-  summarise(total_cases = sum(total_cases, na.rm = TRUE),  # Sum of total cases
-            retail_and_recreation_percent_change_from_baseline = 
-              mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE)) %>%  # Average mobiltiy changes
-  mutate(state = tolower(state))  # Convert state names to lowercase
+ga_map_data <- ga_counties %>%
+  left_join(processeddata, by = c("NAME" = "county"))
 
-# Join the shapefile with the COVID-19 data
-us_map_data <- us_map_clean %>%
-  left_join(data_last_day, by = c("NAME" = "state"))
+ga_health <- processeddata %>%
+  group_by(county) %>%
+  summarize(
+    health_outcome = mean(health_outcome_rank, na.rm = TRUE), 
+    population = first(population_count)  
+  ) %>%
+  ungroup()
 
-# plot total COVID-19 cases to USA map on 10/15/2022
-cases_map <- ggplot(us_map_data) +
-  geom_sf(aes(fill = total_cases), color = "white") +
-  scale_fill_viridis_c(option = "plasma", trans = "log", labels = NULL) + 
-  labs(title = paste("Heatmap of Total COVID-19 Cases in the U.S. on", last_day),
-       fill = "Hotter = More Cases") +
+ga_health_map_data <- ga_counties %>%
+  left_join(ga_health, by = c("NAME" = "county"))
+
+ranking_map <- ggplot(ga_health_map_data) +
+  geom_sf(aes(fill = health_outcome), color = "black") +  
+  scale_fill_viridis_c(option = "inferno", na.value = "grey") +  
   theme_minimal() + 
-  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
+  labs(title = 
+         "Map of Georgia Counties Colored by Average Health Outcome Rankings (2020-2022) ",
+       fill = "Health Outcome Rank (Lower is better)") +
+  theme(legend.position = "right")
 
-print(cases_map)
-
-figure_file = here("results","figures","casemap.png")
-ggsave(filename = figure_file, plot=cases_map)
-
-# Create the heatmap for retail and recreation mobility changes
-mobility_map <- ggplot(us_map_data) +
-  geom_sf(aes(fill = retail_and_recreation_percent_change_from_baseline), color = "white") +  
-  scale_fill_viridis_c(option = "inferno", trans = "identity") +
-  labs(title = paste("Heatmap of Retail and Recreation Mobility Changes in the U.S. on", last_day),
-       fill = "Mobility Change") +
-  theme_minimal() +
-  theme(
-        legend.spacing.y = unit(1, "cm"),  
-        legend.box.spacing = unit(1, "cm"),  
-        legend.key.size = unit(1, "cm")) +  
-  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
-
-print(mobility_graph)
-
-# save graph
-figure_file = here("results","figures","mobilitymap.png")
-ggsave(filename = figure_file, plot=mobility_map)
+print(ranking_map)
 ```
 
-This graph aims to plot total cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all total cases.
 ```{r}
-# Create new data frame to have cases per day instead of state/county
-total_cases_df <- processeddata %>%
+
+top_10_counties <- processeddata %>%
+  group_by(county) %>%
+  filter(year == 2022) %>%
+  summarise(cases = max(new_cases, na.rm = TRUE)) %>%
+  arrange(desc(new_cases)) %>%
+  slice_head(n = 10) %>%
+  pull(county)
+
+# Filter the original data to only include the top 10 counties
+filtered_data <- processeddata %>%
+  filter(county %in% top_10_counties)
+
+ggplot(processeddata, aes(x = week(date), y = county, fill = incidence_rate)) +
+  geom_tile(color = "white") +  # Grid-like separation
+  scale_fill_viridis_c(option = "plasma", na.value = "grey", 
+                       breaks = c(1, 3, 5, 10), 
+                       labels = c("<1", "<3", "<5", "≥10")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()) +
+  labs(title = "Average Weekly Incidence per 100,000",
+       x = "Year", y = "State", fill = "Incidence Rate")
+```
+
+This graph charts prevalence of COVID-19 in 2022 against a County's Health Outcome Rank.
+One severe outlier was removed for the graph's clarity.
+Regardless, from this scatter plot, it seems a county's health outcome ranking is not indicative of their prevention of COVID-19 cases.
+This could be the result of population density in counties of greater ranking.
+After graphing population count against a
+
+```{r}
+health_ranking_cases <- processeddata %>%
+  filter(lubridate::year(date) == 2021) %>%
+  group_by(county) %>%
+  summarize(
+    total_cases_2022 = sum(new_cases, na.rm = TRUE),
+    health_factor_rank = first(health_factor_rank),
+    density = first(pop_density),
+    population_count = first(population_count)
+  ) %>%
+  ungroup() %>%
+  mutate(prevalence = total_cases_2022 / population_count)
+
+health_incidence_scatter <- ggplot(health_ranking_cases, aes(x = health_factor_rank, y = prevalence )) +
+  geom_point(alpha = 0.5, color = "blue") + theme_minimal() +
+  scale_x_reverse() +
+  coord_cartesian(ylim = c(0, 0.2)) +
+  labs(title = "Prevalence of COVID-19 in 2022 vs Health Factor Rank", x = "Health Factor Rank",
+       y= "Prevalence Rate")
+
+print(health_incidence_scatter)
+
+# Include larger values
+ranking_population_scatter_1 <- ggplot(health_ranking_cases, aes(x = health_factor_rank, y = density)) +
+  geom_point(alpha = 0.5, color = "red") + theme_minimal()+
+  scale_x_reverse() +
+  scale_y_continuous(labels = comma) +
+  labs(title = "Health Factor Rank vs Population Density (per km2) in 2022", x = "Health Factor Rank", 
+       y = "Density")
+
+print(ranking_population_scatter_1)
+
+prevalence_density_scatter <- ggplot(health_ranking_cases, aes(x = density, y = prevalence)) +
+  geom_point(alpha = 0.5, color = "red") + theme_minimal()+
+  scale_y_continuous(labels = comma) +
+  labs(title = "Population Density (per km2) vs COVID-19 Prevalence in 2022", 
+       x = "Density (Persons per km2)", 
+       y = "Prevalence")
+
+print(prevalence_density_scatter)
+```
+
+```{r}
+ga_case_count <- processeddata %>%
   group_by(date) %>%
-  summarise(Cases = sum(total_cases, na.rm = TRUE)) %>%
-  ungroup()
+  summarise(ga_total_cases = sum(new_cases, na.rm = TRUE))
 
-# Plot the total case counts over time
-case_count_graph <- ggplot(total_cases_df, aes(x = date, y = Cases)) +
-  geom_line(color = "blue") +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  
-  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
-       x = "Date",
-       y = "Total Cases") +
-  theme_minimal()
+ga_cases_graph <- ggplot(ga_case_count, aes(x = date)) +
+  geom_line(aes(y = ga_total_cases), color = "blue") +
+  labs(title = "Case Count per Day in Georgia",
+       x= "Date",
+       y = "Case Count") + theme_minimal()
 
-print(case_count_graph)
-
-# save graph
-figure_file = here("results","figures","casecountplot.png")
-ggsave(filename = figure_file, plot=case_count_graph)
-```
-This graph aims to plot new cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all new cases.
-```{r}
-# data frame to sum all cases on a given day for every location
-new_cases_df <- processeddata %>%
-  group_by(date) %>%
-  summarise(Cases = sum(new_cases, na.rm = TRUE)) %>%
-  ungroup()
-
-# Plot the new cases over time
-incidence_graph <- ggplot(new_cases_df, aes(x = date, y = Cases)) +
-  geom_line(color = "red") + 
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
-  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
-       x = "Date",
-       y = "New Cases per Day") +
-  theme_minimal()
-
-print(incidence_graph)
-
-# save graph
-figure_file = here("results","figures","incidenceplot.png")
-ggsave(filename = figure_file, plot=incidence_graph)
+print(ga_cases_graph)
 
@@
 # Create dataframe of average mobility for everywhere in the US per date
-us_average_mobility <- processeddata %>%
+ga_average_mobility <- processeddata %>%
   group_by(date) %>%
@@
 # Plot the 7-day averages for each mobility type
-mobility_graph <- ggplot(us_average_mobility, aes(x = date)) +
+mobility_graph <- ggplot(ga_average_mobility, aes(x = date)) +
   geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
@@
   labs(
-    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
+    title = "7-Day Moving Average of Percent Change in Mobility in Georgia Over Time",
     x = "Date",

### 2025-04-04 code/processing-code/processingfile.qmd
Part 4
Hefty changes to my project. First, narrowing the project's scope to just Georgia and GA counties. More modeling work; lot of exploration resulting in unsuccessful models and deletions unfortunately. Also, redid some exploratory analysis but working on the figures.

---
title: "Cleaning Script"
author: "Vincent Nguyen"
date: "2025-02-21"
output: html_document
---

# Setup

Load needed packages. make sure they are installed.

```{r}
# load needed packages
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
library(lubridate)
library(readxl)
library(stringr)
```

# Data loading

```{r}
# google data is in three parts
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv")) %>%
  filter(sub_region_1 == "Georgia")
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))%>%
  filter(sub_region_1 == "Georgia")
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))%>%
  filter(sub_region_1 == "Georgia")

combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)

# jhk covid time series cases
time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv")) %>%
  filter(Province_State == "Georgia")

# county health ranking
#hr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv")) %>%
#  filter(state == "GA")

# county population information
ch_pop_data <- read_excel(here::here("data", "raw-data", "co-est2023-pop.xlsx"), skip = 3)

# Load in CHR Data from 2020 to 2022, remove the first column, remove first two rows, add a year column, and also rename the columns

chr_2020 <- read_excel(here::here("data", "raw-data", "2020 County Health Rankings Georgia Data - v1_0.xlsx"), sheet = 2, skip = 2) %>% select(-1) %>% mutate(year = 2020)

chr_2020 <- chr_2020 %>%
  rename(
    state = `Georgia`,
    county = `...3`,
    health_outcome_zscore = `...4`,
    health_outcome_rank = `...5`,
    health_factor_zscore = `...6`,
    health_factor_rank = `...7`
  )

chr_2021 <- read_excel(here::here("data", "raw-data", "2021 County Health Rankings Georgia Data - v1.xlsx"), sheet = 2, skip = 2)%>% select(-1) %>% mutate(year = 2021)

chr_2021 <- chr_2021 %>%
  rename(
    state = `Georgia`,
    county = `...3`,
    health_outcome_zscore = `...4`,
    health_outcome_rank = `...5`,
    health_factor_zscore = `...6`,
    health_factor_rank = `...7`
  )

chr_2022 <- read_excel(here::here("data", "raw-data", "2022 County Health Rankings Georgia Data - v2.xlsx"), sheet = 2, skip = 2)%>% select(-1) %>% mutate(year = 2022)

chr_2022 <- chr_2022 %>%
  rename(
    state = `Georgia`,
    county = `...3`,
    health_outcome_zscore = `...4`,
    health_outcome_rank = `...5`,
    health_factor_zscore = `...6`,
    health_factor_rank = `...7`
  )

# Combine all CHR data sets into one
combined_chr <- bind_rows(chr_2020, chr_2021, chr_2022)
```

```{r}
# Remove the second column
ch_pop_data <- ch_pop_data[, -2]

ch_pop_data <- ch_pop_data %>%
  pivot_longer(
    cols = c(`2020`,`2021`, `2022`, `2023`),    
    names_to = "year",        
    values_to = "population_count"              
  )

# Fix the first column
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_remove(`...1`, "^\\."),            
    county = str_remove(county, "County"),     
    county = str_to_lower(county),                  
    state = str_extract(county, "(?<=,).*"),        
    county = str_remove(county, ",.*")              
  ) %>%
  mutate(
    state = str_to_lower(str_trim(state))           
  )

# Change year column to be double for joining later
ch_pop_data$year <- as.double(ch_pop_data$year)

# Clean names to prepare for joining
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_trim(str_to_lower(county)),  
    state = str_trim(str_to_lower(state))     
  )

# Only include Georgia counties
ch_pop_data <- ch_pop_data %>%
  filter(state == "georgia")

```

# Cleaning and Wrangling

By inspecting the data, some problems become apparent. AS mentioned above, the Mobility Report is missing a tremendous amount of values. The Time Series data and Mobility Report are structured in entirely different ways. Joining the two will be difficult without mutating one. I chose to change the Time Series data.

Next, we also can see that the naming scheme of the columns differ so we need to standardize that as well.

Lastly, both datasets do not have the same range in dates

```{r}

# Pivot the time series data and change names around
cases_long <- time_series_covid %>%
  pivot_longer(
    cols = starts_with("X"),  
    names_to = "date",
    values_to = "total_cases"
  ) %>%
  mutate(date = gsub("^X", "", date),  # Remove leading x for dates to join data sets in a bit
         date = gsub("\\.", "/", date),  # Replace dots with slashes
         date = as.Date(date, format = "%m/%d/%y"))  # Use "%y" instead of "%Y"

# Standardize county & state names (convert to lowercase to avoid mismatches)
cases_long <- cases_long %>%
  rename(county = Admin2, state = Province_State) %>%
  mutate(county = tolower(county), state = tolower(state))

mobility_df <- combined_mobility_data %>%
  rename(county = sub_region_2, state = sub_region_1) %>%
  mutate(county = tolower(county), 
         state = tolower(state),
         county = gsub(" county$", "", county),  # Remove "county" from county name
         date = as.Date(date, format = "%Y-%m-%d"))  # Correct format for "YYYY-MM-DD"

# Filter cases_long and mobility_df based on the date range
cases_long_filtered <- cases_long %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

mobility_df_filtered <- mobility_df %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

# Merge the data sets together
merged_df <- cases_long_filtered %>%
  inner_join(mobility_df_filtered, by = c("county", "state", "date"))

# Finally, remove any rows missing mobility changes to ensure consistency. We also have a lot of data values so I deemed it permissible.
#merged_df_clean <- merged_df %>%
#  drop_na(c(transit_stations_percent_change_from_baseline, parks_percent_change_from_baseline, retail_and_recreation_percent_change_from_baseline, grocery_and_pharmacy_percent_change_from_baseline,workplaces_percent_change_from_baseline, residential_percent_change_from_baseline))
merged_df_clean <- merged_df
# Additionally, add a column about new cases for further analysis down the line.
merged_df_clean <- merged_df_clean %>%
  arrange(state, county, date) %>%  
  group_by(state, county) %>%  
  mutate(new_cases = c(NA, diff(total_cases))) %>%  
  ungroup()

# Create column for mean average mobiltiy change for that given day and location. I omitted mobility changes to residencies because that may be more of an indication of who is at home.
merged_df_clean <- merged_df_clean %>%
  arrange(state, county, date) %>%  # Sort to ensure correct calculations
  group_by(state, county) %>%  # Group by state and county
  mutate(average_mobility = rowMeans(
    cbind(transit_stations_percent_change_from_baseline, 
          parks_percent_change_from_baseline, 
          retail_and_recreation_percent_change_from_baseline, 
          grocery_and_pharmacy_percent_change_from_baseline, 
          workplaces_percent_change_from_baseline), 
    na.rm = TRUE)) %>%
  ungroup()

merged_df_clean <- merged_df_clean %>%
  mutate(year = year(date)) %>%
  left_join(
    combined_chr %>%
      mutate(across(c(state, county), str_to_lower)),  # Convert state and county to lowercase
    by = c("state", "county", "year")
  )

merged_df_clean <- merged_df_clean %>%
  left_join(
    ch_pop_data %>% 
      select(year, population_count, county, state),
    by = c("state" = "state", "county" = "county", "year" = "year")
  )

# Create column for incidence rate per 10,000 (10,000 because numbers tend to be small when per 100k)
merged_df_clean <- merged_df_clean %>%
  mutate(incidence_rate = (new_cases / population_count) * 10000)

# Clean up and Removal of unnecessary columns
merged_df_clean <- merged_df_clean %>% select(-UID, -iso2, -iso3, -code3, -country_region_code, -country_region, -metro_area, -iso_3166_2_code, -census_fips_code, -place_id)
```

This code chunk entails the creation of a population density column in the data.

```{r}
library(tigris)
library(sf)

# Load county shapefile for GA
ga_counties <- counties(state = "GA", cb = TRUE, class = "sf")

# Make county column lowercase for merging
ga_counties <- ga_counties %>%
  mutate(NAME = tolower(NAME))

# Merge with processed population data (assumed already cleaned)
ga_counties <- ga_counties %>%
  left_join(ch_pop_data, by = c("NAME" = "county"))

# Calculate county area in square kilometers
ga_counties <- ga_counties %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6)  # Convert m² to km²

# Compute population density (people per km²) and remove unnecessary data
ga_counties <- ga_counties %>%
  mutate(pop_density = population_count / area_km2) %>%
  select(pop_density, year, NAME)

# Add in population density column
merged_df_clean <- merged_df_clean %>%
  left_join(ga_counties, by = c("county" = "NAME", "year" = "year"))
```
```{r}
# Clean negative and NA values
merged_df_clean$new_cases[merged_df_clean$new_cases < 0] <- 0
merged_df_clean$new_cases[is.na(merged_df_clean$new_cases)] <- 0
merged_df_clean$incidence_rate[is.na(merged_df_clean$incidence_rate)] <- 0

# Remove any remaining NA values
merged_df_clean <- na.omit(merged_df_clean)

# Filter for pandemic period
last_case_date <- as.Date("2022-01-01")
first_case_date <- as.Date("2020-03-02")

merged_df_clean <- merged_df_clean %>%
  filter(date >= first_case_date) %>%
  filter(date <= last_case_date) %>%
  group_by(county)

# Convert to tibble and remove geometry
merged_df_clean <- merged_df_clean %>%
  as_tibble() %>%
  select(-geometry)

# Convert new_cases to integer
merged_df_clean <- merged_df_clean %>% 
  mutate(new_cases = as.integer(new_cases))
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- merged_df_clean
```

# Save data

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2025-04-04 code/analysis-code/statistical-analysis.R

@@ -1,176 +1,0 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
-
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
-library(glmnet)
-library(MASS)
-
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
-
-#load data. 
-data <- readRDS(data_location)
-
-# NOTES TO SELF
-# - change to QMD file (personally prefer)
-# - work on other analysis (lag-time & quarantine fatigue? more research)
-# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
-
-
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-############################
-#### Initial model fitting
-
-# Some quick data fixing before analysis as some models cannot handle negatives (contextually ok, more interested in NEW cases)
-data_clean <- data
-data_clean$new_cases[data_clean$new_cases < 0] <- 0
-
-
-# first Negative Binomial Model
-nb_model_1 <- glm.nb(new_cases ~ population_count + transit_stations_percent_change_from_baseline + 
-                     parks_percent_change_from_baseline + 
-                     retail_and_recreation_percent_change_from_baseline + 
-                     grocery_and_pharmacy_percent_change_from_baseline +
-                     workplaces_percent_change_from_baseline + 
-                     residential_percent_change_from_baseline +
-                     population_count, 
-                   data = data_clean)
-
-# View summary
-summary(nb_model_1)
-
-# Poisson regression model based on the same formula
-poisson_model_1 <- glm(new_cases ~ transit_stations_percent_change_from_baseline + 
-                       parks_percent_change_from_baseline + 
-                       retail_and_recreation_percent_change_from_baseline + 
-                       grocery_and_pharmacy_percent_change_from_baseline +
-                       workplaces_percent_change_from_baseline + 
-                       residential_percent_change_from_baseline +
-                       population_count, 
-                     data = data_clean, family = poisson(link = "log"))
-
-# View the summary of the Poisson model
-summary(poisson_model_1)
[…]

### 2025-04-04 products/manuscript/Manuscript.qmd
Fixes

@@
 In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lockdown procedures (CDC 2024)[@centersfordiseasecontrolandprevention_2024_covid19].
+
+On March 2nd, 2020, Governor Brian Kemp announces the first two confirmed COVID-19 cases in the state, a Father and Son returning from Italy.
+At the same time as the rest of the nation, March 11th, Governor Kemp announced a state of emergency and closed all public schools on March 16th indefinitely.
+In the span of the next two weeks, Georgia saw its first and hundredth death to the virus.
+Weeks later, on April 2nd, Governor Kemp announced a shelter in place order for all Georgia residents.
+After only a month, on April 30th, the shelter-in-place order expired allowing Georgians to go to gyms, salons, barberships, and shortly after, restaurants.
 

### 2025-04-04 code/analysis-code/statisitical-analysis.qmd
Fixes

@@
 ```{r}
-# Some quick data fixing before analysis as some models cannot handle negatives (contextually ok, more interested in NEW cases)
-data_clean <- data
-data_clean$new_cases[data_clean$new_cases < 0] <- 0
-data_clean$new_cases[is.na(data_clean$new_cases)] <- 0
-data_clean$incidence_rate[is.na(data_clean$incidence_rate)] <- 0
-
-data_clean <- na.omit(data_clean)
-
-# In my intial analyses, there were an absurd amount of zeroes; this was most likely due to the data sets capturing days prior to the pandemic.
-last_case_date <- as.Date("2022-01-01")
-first_case_date <- as.Date("2020-03-02")
-
-df_clean <- data_clean %>%
-  filter(date >= first_case_date) %>%
-  filter(date <= last_case_date) %>%
-  group_by(county)
-
-df_clean <- df_clean %>%
-  as_tibble() %>%
-  select(-geometry)
-
-df_clean <- df_clean %>% 
-  mutate(new_cases = as.integer(df_clean$new_cases))
-
-class(df_clean$new_cases)
-```
-
-```{r}
 # Intial Recipe setting
@@
   add_model(lm_spec) %>%
-  fit(data = df_clean)
+  fit(data = data)
 
@@
 
-lm_preds<- predict(lm_fit, new_data = df_clean)
+lm_preds<- predict(lm_fit, new_data = data)
 
-lm_metrics <- tibble(truth = df_clean$new_cases, predicted = lm_preds$.pred) %>%
+lm_metrics <- tibble(truth = data$new_cases, predicted = lm_preds$.pred) %>%
   metrics(truth = truth, estimate = predicted)
@@
   add_model(lasso_spec) %>%
-  fit(data = df_clean)
+  fit(data = data)
 
@@
 
-lasso_preds<- predict(lasso_fit, new_data = df_clean)
+lasso_preds<- predict(lasso_fit, new_data = data)
 
-lasso_metrics <- tibble(truth = df_clean$new_cases, predicted = lasso_preds$.pred) %>%
+lasso_metrics <- tibble(truth = data$new_cases, predicted = lasso_preds$.pred) %>%
   metrics(truth = truth, estimate = predicted)
@@
 plan(multisession, workers = 4)
+
 # Tune the forest
+# This step took 20-ish minutes
 forest_tune_res <- tune_grid(
@@
 
-best_params <- select_best(forest_tune_res, metric = "rmse")  # Or "rsq"
+best_params <- select_best(forest_tune_res, metric = "rmse")
 best_params
@@
 
-# Calculate performance metrics (RMSE, R², etc.)
+# Calculate performance metrics
 forest_rmse <- forest_preds_final %>%

### 2025-04-18 code/analysis-code/statisitical-analysis.qmd
Work

@@
 
+Note to self:
+Tried lagging mobility (doesnt improve models)
+Tried filtering data to pre-vaccination (doesnt improve models)
+
 ```{r}
@@
 library(rsample)
-library(poissonreg)
+library(tibble)
 library(parsnip)
 library(future)
+library(vip)
 ```
@@
 
-# NOTES TO SELF
-# - change to QMD file (personally prefer)
-# - work on other analysis (lag-time & quarantine fatigue? more research)
-# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
-class(data$new_cases)
+```
+
+```{r}
+# Creation of lagged cases; commonly used in epi modeling
+# Note: lagged mobility does not significantly improve results from what I can tell
+data <- data %>%
+  group_by(county) %>%
+  arrange(date) %>%
+  mutate(
+    lag_1 = lag(new_cases, 1),
+    lag_7 = lag(new_cases, 7),
+    lag_14 = lag(new_cases, 14),
+  ) %>%
+  ungroup()
+
+data <- data %>% drop_na()
+```
+
+Ok, With help from ChatGPT, I implemented time blocked CV folds.
+The idea is that the data needs to respect the time-series nature and rolling window cannot be implemented because of the missing days in some counties.
+
+This code splits the data into 8 folds.
+For each fold, the model trains on all the data up to and incuding the last fold before the test set.
+For example, the model will train on fold 1 and test on fold 2.
+Then, the model will train on folds 1 + 2 and test on fold 3.
+```{r}
+# Set seed
+seed <- 1234
+set.seed(seed)
+
+library(dplyr)
+library(lubridate)
+library(rsample)
+
+# Step 1: Ensure the data is ordered by date
+train_data <- data %>%
+  arrange(date)
+
+# Step 2: Create a new variable that indicates the month or quarter
+train_data <- train_data %>%
+  mutate(month = floor_date(date, "month"))
+
+# Step 3: Define the fold boundaries manually based on your specific date range
+train_data <- train_data %>%
+  mutate(fold_id = case_when(
+    month >= as.Date("2020-03-01") & month <= as.Date("2020-06-13") ~ 1,
+    month >= as.Date("2020-06-14") & month <= as.Date("2020-09-13") ~ 2,
+    month >= as.Date("2020-09-14") & month <= as.Date("2020-12-13") ~ 3,
+    month >= as.Date("2020-12-14") & month <= as.Date("2021-03-14") ~ 4,
+    month >= as.Date("2021-03-15") & month <= as.Date("2021-06-14") ~ 5,
+    month >= as.Date("2021-06-15") & month <= as.Date("2021-09-14") ~ 6,
+    month >= as.Date("2021-09-15") & month <= as.Date("2021-12-14") ~ 7,
+    month >= as.Date("2021-12-15") & month <= as.Date("2022-03-14") ~ 8
+  ))
+
+# Step 4: Check the number of observations per fold and the date ranges
+folds_summary <- train_data %>%
+  group_by(fold_id) %>%
+  summarise(
+    min_date = min(date),
+    max_date = max(date),
+    n_obs = n()
+  ) %>%
+  print()
+
+# Step 5: Create the blocked CV folds using the fold_id
+folds <- group_vfold_cv(train_data, group = "fold_id")
+
+# Set aside last fold for testing
+test_data <- train_data %>% filter(fold_id == 8)
+train_data_final <- train_data %>% filter(fold_id < 8)
+
+```
+
+```{r}
+library(rsample)
+library(dplyr)
+library(lubridate)
+library(rsample)
+
+# Step 0: Make sure your data is sorted
+data <- data %>%
+  arrange(date, county)
+
+# Step 1: Define test set as final 30 days
+test_days <- 30
+last_date <- max(data$date)
+test_start_date <- last_date - days(test_days - 1)
+
+train_data <- data %>% filter(date < test_start_date)
+test_data  <- data %>% filter(date >= test_start_date)
+
+rolling_folds <- rolling_origin(
+  data = train_data,
+  initial = 120 * 44,  # ~4 months of training
+  assess  = 30 * 44,   # ~1 month of validation
+  skip    = 15 * 44,   # roll forward 1 month
+  cumulative = FALSE
+)
+
+# Check date ranges for each fold
+train_ranges <- purrr::map(rolling_folds$splits, ~range(pull(training(.x), date)))
+
+# Print the ranges for each fold
+train_ranges
+
+# Step 3: (Optional) View fold summary
+fold_sizes <- purrr::map_dfr(rolling_folds$splits, function(split) {
+  tibble(
+    train_n = nrow(training(split)),
+    test_n  = nrow(testing(split))
+  )
+})
+
+print(fold_sizes)
+
+# Check date ranges too (optional)
+range(train_data$date)
+range(test_data$date)
+
 ```
@@
 # Intial Recipe setting
-recipe <- recipe(new_cases ~ pop_density + transit_stations_percent_change_from_baseline + 
-                     parks_percent_change_from_baseline + 
+recipe <- recipe(new_cases ~ pop_density +
                   retail_and_recreation_percent_change_from_baseline +
@@
                    residential_percent_change_from_baseline +
-                     health_factor_rank, data = data)
+                     lag_1 + lag_7 + lag_14, data = data)
 
+recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data)
 ```
 
-First, I created a linear regression model as the baseline standard.
-This model performs very poorly as it has a low R-squared, indicating low predictive power.
-The model does mark all variables of the model to be significant.Additionally, the AIC is very extremely high.
+```{r}
+# Define your model specification (linear regression)
+base_lm_spec <- linear_reg() %>%
+  set_engine("lm")
+
+# Create the workflow with the baseline recipe
+base_lm_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_lm_spec)
+
+# Resampling with cross-validation
+base_lm_res <- fit_resamples(
+  base_lm_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# Collect and print metrics from resampling
+base_lm_metrics <- collect_metrics(base_lm_res)
+print(base_lm_metrics)
+
+# Collect and evaluate predictions
+base_lm_res %>%
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+# Store metrics
+print(base_lm_metrics)
+
+# Fitting the final model on the training data
+base_lm_fit <- base_lm_wf %>%
+  fit(data = train_data)
+
+# Plot variable importance
+vip(base_lm_fit$fit$fit)
+```
 
@@
 lm_wf <- workflow() %>%
-  add_recipe(recipe)
+  add_recipe(recipe) %>%
+  add_model(lm_spec)
 
+# Resampling with CV
+lm_res <- fit_resamples(
+  lm_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# Collect and print metrics from resampling
+collect_metrics(lm_res)
+
+lm_res %>% 
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+# Store metrics
+lm_metrics <- collect_metrics(lm_res)
+print(lm_metrics)
+
+# Fitting final model
 lm_fit <- lm_wf %>%
-  add_model(lm_spec) %>%
-  fit(data = data)
+  fit(data = train_data)
 
-lm_fit %>%
-  extract_fit_parsnip() %>%
-  tidy()
+# Plot variable importance
+vip(lm_fit$fit$fit)
 
-lm_fit %>%
-  extract_fit_parsnip() %>%
-  broom::glance()
-
-lm_preds<- predict(lm_fit, new_data = data)
-
-lm_metrics <- tibble(truth = data$new_cases, predicted = lm_preds$.pred) %>%
-  metrics(truth = truth, estimate = predicted)
-lm_metrics
 ```
 
-I utilized Lasso here to hopefully bring down the RMSE and bring up the rsquared.
-It seems Lasso has no effect on model performance as the data lacks multicolinearity and does not have irrelevant predictors.
+After the linear regression model performed poorly, the use of random forest seemed rational as this model can capture more complex relationships and handle the quirks of the data.
+With an r-squared of 0.7824, this model does capture more variance than the linear regression model.
+Additionally, the MAE has improvements as well.
+The use of RMSE may be difficult due to the presence of outliers, or in other words, days where huge outbreaks occured.
 ```{r}
-# Set method for LASSO
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
+rngseed = 1234
 
-# Insert recipe
-lasso_wf <- workflow() %>%
-  add_recipe(recipe)
+# base_1: Forest specification (Random Forest)
+base_forest_spec <- rand_forest(mode = "regression") %>%
+  set_engine("ranger", seed = rngseed, importance = "impurity")
 
-lasso_fit <- lasso_wf %>%
-  add_model(lasso_spec) %>%
-  fit(data = data)
+# base_2: Insert baseline recipe into workflow
+base_forest_wf <- workflow() %>%
+  add_recipe(recipe_baseline)
 
-lasso_fit %>%
-  extract_fit_parsnip() %>%
-  tidy()
+# base_3: Time-blocked CV (folds already created earlier)
+# base_folds <- group_vfold_cv(train_data, group = fold_id)
 
-lasso_fit %>%
-  extract_fit_parsnip() %>%
-  broom::glance()
+# base_4: Perform cross-validation with resampling
+base_forest_res <- fit_resamples(
+  base_forest_wf %>% add_model(base_forest_spec),
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
 
-lasso_preds<- predict(lasso_fit, new_data = data)
+# base_5: View fold performance
+base_forest_res %>%
+  collect_metrics() %>%
+  print()
 
-lasso_metrics <- tibble(truth = data$new_cases, predicted = lasso_preds$.pred) %>%
-  metrics(truth = truth, estimate = predicted)
-lasso_metrics
+# base_6: Per-fold metrics
+base_forest_res %>% 
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+# base_7: Store metrics
+base_forest_metrics <- collect_metrics(base_forest_res)
+
+# base_8: Residuals
+base_forest_preds <- base_forest_res %>%
+  collect_predictions() %>%
+  mutate(residual = new_cases - .pred)
+
+# Plot residuals
+ggplot(base_forest_preds, aes(x = residual)) +
+  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Histogram of Residuals (Random Forest)",
+    x = "Residual",
+    y = "Count"
+  ) +
+  theme_minimal()
+
+# base_9: Fit final model on full train_data
+base_forest_fit <- base_forest_wf %>%
+  add_model(base_forest_spec) %>%
+  fit(data = train_data)
+
+# base_10: Variable importance plot
+vip(base_forest_fit$fit$fit)
 ```
 
-After the last two models performed poorly, the use of random forest seemed rational as this model can capture more complex relationships. 
 ```{r}
@@
 
+# Forest specification (Random Forest)
 forest_spec <- rand_forest(mode = "regression") %>%
+  set_engine("ranger", seed = rngseed, importance = "impurity")
+
+# Insert recipe
+forest_wf <- workflow() %>%
+  add_recipe(recipe)
+
+# Step 1: Resampling with time-blocked CV (folds from previous code)
+# folds <- group_vfold_cv(train_data, group = fold_id)
+
+# Step 2: Perform cross-validation with resampling
+forest_res <- fit_resamples(
+  forest_wf %>% add_model(forest_spec),    # Add forest model to workflow
+  resamples = rolling_folds,                       # Use the time-blocked folds
+  metrics = metric_set(rmse, rsq, mae),    # Choose evaluation metrics
+  control = control_resamples(save_pred = TRUE)   # Save predictions for each fold
+)
+
+# Step 3: View fold performance
+forest_res %>%
+  collect_metrics() %>%
+  print()
+
+forest_res %>% 
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+# Store metrics
+forest_metrics <- collect_metrics(forest_res)
+
+forest_preds <- forest_res %>%
+  collect_predictions() %>%
+  mutate(residual = new_cases - .pred)
+
+ggplot(forest_preds, aes(x = residual)) +
+  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Histogram of Residuals (XGBoost)",
+    x = "Residual",
+    y = "Count"
+  ) +
+  theme_minimal()
+
+forest_fit <- forest_wf %>%
+  add_model(forest_spec) %>%
+  fit(data = train_data)
+
+vip(forest_fit$fit$fit)
+```
+
+Next, I explored the use of an XGBoost model as the Random Forest went well.
+These models tend to perform better.
+With slightly improved r-squared and similar MAE, both models are promising and are tuned in code chunks ahead.
+```{r}
+# base_1: Define the XGBoost model specification
+base_boost_spec <- boost_tree(
+  trees = 1000,
+  tree_depth = 6,
+  learn_rate = 0.01,
+  loss_reduction = 0.01,
+  sample_size = 0.8,
+  mtry = 0.8
+) %>%
+  set_engine("xgboost", counts = FALSE) %>%
+  set_mode("regression")
+
+# base_2: Build the workflow with baseline recipe and XGBoost model
+base_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_boost_spec)
+
+# base_3: Fit and evaluate the model using cross-validation
+base_boost_res <- fit_resamples(
+  base_boost_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# base_4: View fold performance
+base_boost_res %>%
+  collect_metrics() %>%
+  print()
+
+# base_5: Fold-wise prediction metrics
+base_boost_res %>% 
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+# base_6: Collect metrics for storage
+base_boost_metrics <- collect_metrics(base_boost_res)
+
+# base_7: Collect predictions and compute residuals
+base_boost_preds <- base_boost_res %>%
+  collect_predictions() %>%
+  mutate(residual = new_cases - .pred)
+
+base_boost_fit <- base_boost_wf %>%
+  fit(data = train_data)
+
+vip(base_boost_fit$fit$fit)
+```
+
+```{r}
+# Define the XGBoost model specification
+boost_spec <- boost_tree(
+  trees = 1000,
+  tree_depth = 6,
+  learn_rate = 0.01,
+  loss_reduction = 0.01,
+  sample_size = 0.8,
+  mtry = 0.8
+) %>%
+  set_engine("xgboost", counts = FALSE) %>%
+  set_mode("regression")
+
+# Build the workflow with recipe and model
+boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(boost_spec)
+
+# Fit and evaluate the model using cross-validation
+boost_res <- fit_resamples(
+  boost_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# Step 3: View fold performance
+boost_res %>%
+  collect_metrics() %>%
+  print()
+
+boost_res %>% 
+  collect_predictions() %>%
+  group_by(id) %>%
+  metrics(truth = new_cases, estimate = .pred)
+
+boost_metrics <- collect_metrics(boost_res)
+
+boost_fit <- boost_wf %>%
+  fit(data = train_data)
+
+vip(boost_fit$fit$fit)
+```
+
+```{r}
+# Define the Random Forest model with tuning parameters (baseline version)
+base_forest_spec <- rand_forest(
+  mode = "regression", 
+  mtry = tune(),           
+  min_n = tune(),          
+  trees = 500              
+) %>%
+  set_engine("ranger", seed = rngseed)
+
+# Insert baseline recipe
+base_forest_wf <- workflow() %>%
+  add_recipe(recipe_baseline)
+
+# Create grid of hyperparameters for tuning
+base_forest_grid <- grid_regular(
+  mtry(range = c(1, 4)),  
+  min_n(range = c(1, 50)), 
+  levels = 10
+)
+
+# Set seed for reproducibility
+set.seed(1234)
+
+# Tune the Random Forest model using time-blocked CV (folds)
+base_forest_tune_res <- tune_grid(
+  base_forest_wf %>% add_model(base_forest_spec),
+  resamples = rolling_folds,                     
+  grid = base_forest_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# Plot tuning results
+autoplot(base_forest_tune_res)
+
+# Collect and print the performance metrics from tuning
+base_forest_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+# Get best hyperparameters based on R-squared
+base_best_params <- select_best(base_forest_tune_res, metric = "rsq")
+print(base_best_params)
+
+# Final baseline RF model with best parameters
+base_final_forest_spec <- rand_forest(
+  mode = "regression",
+  mtry = base_best_params$mtry,
+  min_n = base_best_params$min_n,
+  trees = 500
+) %>%
+  set_engine("ranger", seed = rngseed)
+
+# Final baseline workflow
+base_final_forest_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_final_forest_spec)
+
+# Fit final baseline model on training data
+base_final_forest_fit <- base_final_forest_wf %>%
+  fit(data = train_data)
+
+base_final_forest_res <- fit_resamples(
+  base_final_forest_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+print(base_final_forest_res)
+
+#test_data <- train_data %>% filter(fold_id == 8)
+
+# Predict on test data
+base_final_forest_preds <- predict(base_final_forest_fit, new_data = test_data)
+
+# Prepare prediction results
+base_final_forest_preds <- tibble(
+  truth = test_data$new_cases, 
+  predicted = base_final_forest_preds$.pred
+)
+
+# Evaluate performance
+base_final_forest_metrics <- base_final_forest_preds %>%
+  metrics(truth = truth, estimate = predicted)
+
+print(base_final_forest_metrics)
+```
+
+```{r}
+# Define the Random Forest model with tuning parameters
+forest_spec <- rand_forest(
+  mode = "regression", 
+  mtry = tune(),           
+  min_n = tune(),          
+  trees = 500              
+) %>%
   set_engine("ranger", seed = rngseed)
@@
 
-# Create model with all predictors and random forest
-forest_fit <- forest_wf %>%
- add_model(forest_spec) %>%
-  fit(data = data)
+# Create grid of hyperparameters for tuning
+forest_grid <- grid_regular(
+  mtry(range = c(1, 8)),  
+  min_n(range = c(1, 50)), 
+  levels = 10
+)
 
-forest_preds <- predict(forest_fit, new_data = data)
-forest_preds <- tibble(truth = data$new_cases, predicted = forest_preds$.pred)
-forest_rmse <- forest_preds %>%
+# Set seed for reproducibility
+set.seed(1234)
+
+# Tune the Random Forest model using time-blocked CV (folds)
+forest_tune_res <- tune_grid(
+  forest_wf %>% add_model(forest_spec),
+  resamples = rolling_folds,                     
+  grid = forest_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# Plot tuning results to visualize performance across hyperparameters
+autoplot(forest_tune_res)
+
+# Collect and print the performance metrics from the tuning process
+forest_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+# Get the best hyperparameters based on the chosen metric (e.g., R-squared)
+best_params <- select_best(forest_tune_res, metric = "rsq")
+print(best_params)
+
+# Train the final model with the best parameters on the full training set
+final_forest_spec <- rand_forest(
+  mode = "regression",
+  mtry = best_params$mtry,
+  min_n = best_params$min_n,
+  trees = 500
+) %>%
+  set_engine("ranger", seed = rngseed)
+
+# Final model workflow with the best hyperparameters
+final_forest_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(final_forest_spec)
+
+# Fit the final model on the entire training data
+final_forest_fit <- final_forest_wf %>%
+  fit(data = train_data)
+
+# Make predictions on the test data
+final_forest_preds <- predict(final_forest_fit, new_data = test_data)
+
+# Prepare prediction results
+final_forest_preds <- tibble(truth = test_data$new_cases, predicted = final_forest_preds$.pred)
+
+# Calculate RMSE for the final model
+final_forest_metrics <- final_forest_preds %>%
   metrics(truth = truth, estimate = predicted)
-print(forest_rmse)
+
+# Print final performance metrics
+print(final_forest_metrics)
+
+final_forest_preds <- final_forest_preds %>%
+  mutate(residual = truth - predicted)
+
+ggplot(final_forest_preds, aes(x = residual)) +
+  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Histogram of Residuals (XGBoost)",
+    x = "Residual",
+    y = "Count"
+  ) +
+  theme_minimal()
+
 ```
@@
 ```{r}
-# Tuning forest
-forest_spec <- rand_forest(
-  mode = "regression", 
-  trees = 300, 
-  mtry = tune(), 
-  min_n = tune()
+# Define the XGBoost model specification (baseline version)
+base_boost_spec <- boost_tree(
+  trees = 500,             # Trees to tune
+  tree_depth = tune(),     # Depth of trees to tune
+  learn_rate = tune(),
+  loss_reduction = tune()
 ) %>%
-  set_engine("ranger")
+  set_engine("xgboost") %>%
+  set_mode("regression")
 
-# Create grid with set parameters
-forest_grid <- grid_regular(
-  mtry(range = c(1, 7)),  
-  min_n(range = c(1, 21)),  
-  levels = 7  
-)
+# Build the workflow with the baseline recipe and model
+base_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_boost_spec)
 
-# Set seed
-set.seed(1234)
-
-# 5 fold crossvalidation repeated 5 times
-# Note: Limitation of computing power so doing rolling window would not work
-cv_folds <- vfold_cv(data, v = 5, repeats = 5)
-plan(multisession, workers = 4)
-
-# Tune the forest
-# This step took 20-ish minutes
-forest_tune_res <- tune_grid(
-  forest_wf %>% add_model(forest_spec),
-  resamples = cv_folds,  
-  grid = forest_grid,
-  metrics = metric_set(rmse, rsq),
+# Tune the XGBoost model using time-blocked CV (folds)
+base_boost_tune_res <- tune_grid(
+  base_boost_wf,
+  resamples = rolling_folds,
+  grid = 30,
+  metrics = metric_set(rmse, rsq, mae),
   control = control_grid(parallel_over = "everything")
@@
 
-# Plot results
-autoplot(forest_tune_res)
+# View tuning results
+base_boost_tune_res %>%
+  collect_metrics() %>%
+  print()
 
-forest_tune_res %>%
-  collect_metrics()
+# Get the best hyperparameters based on R-squared
+base_best_params <- select_best(base_boost_tune_res, metric = "rsq")
+print(base_best_params)
 
-best_params <- select_best(forest_tune_res, metric = "rmse")
-best_params
+# Collect the metrics of the best model based on R-squared
+base_best_metrics <- base_boost_tune_res %>%
+  collect_metrics() %>%
+  filter(.config == base_best_params$.config)
 
-best_params$trees <- 500
+# Print best model metrics
+print(base_best_metrics)
+
+# Plot tuning performance
+autoplot(base_boost_tune_res)
+
+# Train the final baseline XGBoost model with best parameters
+base_final_boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = base_best_params$tree_depth,
+  learn_rate = base_best_params$learn_rate,
+  loss_reduction = base_best_params$loss_reduction
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+# Final model workflow using the baseline recipe
+base_final_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_final_boost_spec)
+
+# Fit the final XGBoost baseline model
+base_final_boost_fit <- base_final_boost_wf %>%
+  fit(data = train_data)
+
+# Predict on the test data
+base_final_boost_preds <- predict(base_final_boost_fit, new_data = test_data)
+
+# Prepare results
+base_final_boost_preds <- tibble(
+  truth = test_data$new_cases, 
+  predicted = base_final_boost_preds$.pred
+)
+
+# Calculate final performance metrics
+base_final_boost_metrics <- base_final_boost_preds %>%
+  metrics(truth = truth, estimate = predicted)
+
+# Print performance metrics
+print(base_final_boost_metrics)
 ```
@@
 ```{r}
-forest_spec <- rand_forest(
-  mode = "regression",
-  mtry = best_params$mtry,
-  trees = best_params$trees,
-  min_n = best_params$min_n
+# Define the XGBoost model specification
+boost_spec <- boost_tree(
+  trees = 500,             # Trees to tune
+  tree_depth = tune(),     # Depth of trees to tune
+  learn_rate = tune(),
+  loss_reduction = tune()
 ) %>%
-  set_engine("ranger", seed = 1234)
+  set_engine("xgboost") %>%
+  set_mode("regression")
 
-forest_fit_final <- forest_wf %>%
-  add_model(forest_spec) %>%
-  fit(data = data)
+# Build the workflow with recipe and model
+boost_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(boost_spec)
 
-# Get final predictions
-forest_preds_final <- predict(forest_fit_final, new_data = data)
-forest_preds_final <- tibble(truth = data$new_cases, predicted = forest_preds_final$.pred)
+# Tune the XGBoost model using time-blocked CV (folds)
+boost_tune_res <- tune_grid(
+  boost_wf,
+  resamples = rolling_folds,
+  grid = 30,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
 
-# Calculate performance metrics
-forest_rmse <- forest_preds_final %>%
+# View tuning results
+boost_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+# Get the best hyperparameters based on R-squared
+best_params <- select_best(boost_tune_res, metric = "rsq")
+print(best_params)
+
+# Collect the metrics of the best model based on R-squared
+best_metrics <- boost_tune_res %>%
+  collect_metrics() %>%
+  filter(.config == best_params$.config)
+
+# Print the metrics (R-squared, RMSE, MAE) for the best model
+print(best_metrics)
+
+# Plot the results of the tuning process
+autoplot(boost_tune_res)
+
+# Train the final model with the best parameters on the full training set
+final_boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = best_params$tree_depth,
+  learn_rate = best_params$learn_rate
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+# Final model workflow with the best hyperparameters
+final_boost_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(final_boost_spec)
+
+# Fit the final XGBoost model on the entire training data
+final_boost_fit <- final_boost_wf %>%
+  fit(data = train_data)
+
+# Make predictions on the test data
+final_boost_preds <- predict(final_boost_fit, new_data = test_data)
+
+# Prepare prediction results (truth and predicted values)
+final_boost_preds <- tibble(truth = test_data$new_cases, predicted = final_boost_preds$.pred)
+
+# Calculate RMSE, MAE, and R-squared for the final model
+final_boost_metrics <- final_boost_preds %>%
   metrics(truth = truth, estimate = predicted)
 
-print(forest_rmse)
+# Print final performance metrics (RMSE, MAE, and R-squared)
+print(final_boost_metrics)
+
+final_boost_preds <- final_boost_preds %>%
+  mutate(residual = truth - predicted)
+
+ggplot(final_boost_preds, aes(x = residual)) +
+  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Histogram of Residuals (XGBoost)",
+    x = "Residual",
+    y = "Count"
+  ) +
+  theme_minimal()
+
 ```
+
+```{r}
+base_lm_metrics <- base_lm_metrics %>% 
+  mutate(model = "Linear Regression (Baseline)") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+lm_metrics <- lm_metrics %>% 
+  mutate(model = "Linear Regression") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+base_forest_metrics <- base_forest_metrics %>% 
+  mutate(model = "Random Forest (Baseline)") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+forest_metrics <- forest_metrics %>% 
+  mutate(model = "Random Forest") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+base_boost_metrics <- base_boost_metrics %>% 
+  mutate(model = "XGBoost (Baseline)") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+boost_metrics <- boost_metrics %>% 
+  mutate(model = "XGBoost") %>%
+  select(-n, -std_err, -.config, -.estimator)
+
+final_boost_metrics <- final_boost_metrics %>%
+  mutate(model = "Tuned XGBoost") %>%
+  rename(mean = .estimate)
+
+final_forest_metrics <- final_forest_metrics %>%
+  mutate(model = "Tuned Random Forest") %>%
+  rename(mean = .estimate)
+
+base_final_boost_metrics <- base_final_boost_metrics %>%
+  mutate(model = "Tuned XGBoost (Baseline)") %>%
+  rename(mean = .estimate)
+
+base_final_forest_metrics <- base_final_forest_metrics %>%
+  mutate(model = "Tuned Random Forest (Baseline)") %>%
+  rename(mean = .estimate)
+
+model_metrics <- bind_rows(base_lm_metrics, lm_metrics, base_forest_metrics, forest_metrics, base_boost_metrics, boost_metrics, base_final_boost_metrics, final_boost_metrics, base_final_forest_metrics, final_forest_metrics)
+
+model_metrics <- model_metrics %>%
+  select(-.estimator) %>%
+  pivot_wider(names_from = '.metric', values_from = mean)
+```
+
+```{r}
+library(gt)
+
+model_table <- model_metrics %>%
+  gt() %>%
+  tab_header(
+    title = "Model Performance Metrics",
+    subtitle = "Best values are highlighted per metric"
+  ) %>%
+  fmt_number(columns = c(rmse, mae), decimals = 1) %>%
+  fmt_number(columns = rsq, decimals = 3) %>%
+  cols_label(
+    model = "Model",
+    rmse = "RMSE",
+    rsq = "R²",
+    mae = "MAE"
+  ) %>%
+
+  # Highlight lowest RMSE
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rmse,
+      rows = rmse == min(rmse, na.rm = TRUE)
+    )
+  ) %>%
+
+  # Highlight lowest MAE
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = mae,
+      rows = mae == min(mae, na.rm = TRUE)
+    )
+  ) %>%
+
+  # Highlight highest R²
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rsq,
+      rows = rsq == max(rsq, na.rm = TRUE)
+    )
+  ) %>%
+
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_options(
+    table.width = pct(100),
+    table.align = "left",
+    column_labels.padding = px(10),
+    data_row.padding = px(6)
+  )
+
+library(gtExtras)
+library(webshot2)
+
+print(model_table)
+
+gtsave(model_table, here("results", "figures", "models.png"))
+```

### 2025-04-18 code/eda-code/eda.qmd
Work

--- 2025-02-21 code/eda-code/eda.qmd
@@
 
-This Quarto file loads the cleaned data and does some exploring.
-# Setup
+This Quarto file loads the cleaned data and does some exploring. \# Setup
 
@@
 library(dplyr)
-library(skimr)
 library(ggplot2)
 library(zoo)
-library(tidyr)
-library(tigris)
-library(sf)
+library(patchwork)
+library(e1071)
+library(scales)
+library(RColorBrewer)
+library(corrplot)
+library(reshape2)
+
 ```
@@
 # processed data
-processeddata <- readRDS(here("data", "processed-data", "processeddata.rds"))
+data <- readRDS(here("data", "processed-data", "processeddata.rds"))
 
+case_data <- readRDS(here("data", "processed-data", "processedcasedata.rds"))
+
+data<- data %>%
+   mutate(incidence_rate = if_else(incidence_rate < 0, 0, incidence_rate))
 ```
 
-Creation of a heatmap showing total COVID 19 cases on the last observed date, 10/15/2022.
-I also created a heat map showing mobility changes, with hotter indicating more mobility to retail and recreational spots.
-Interestingly, the two maps seem to correlate even though I used total COVID-19 cases rather than new ones.
+DataVisualization
+
+Since new case incidence is the variable of interest, it was visually inspected to understand the distribution.
+
+From the graphs, kurtosis test, and skewness test, it is apparent that daily new case incidence is not normally distribued and heavily skewed towards 0 and other lower values.
+The second graph, new cases over 500, was graphed to show how certain days do reach high case incidence.
 ```{r}
-# Load the state-level shapefile 
-us_map_states <- tigris::states(cb = TRUE) %>%
-  st_as_sf()  # Convert the shapefile to a Simple Feature object
 
-# Clean the state names in the shapefile
-us_map_clean <- us_map_states %>%
-  mutate(NAME = tolower(NAME))  # Convert state abbreviation to lowercase
+density_cases <- data %>%
+  ggplot( aes(x=new_cases)) +
+    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
+  labs(title = "Distribution of Daily New Case Incidence", 
+       x = "Daily New Cases",
+       y = "Density") +
+  theme_bw()
 
-# Filter your dataset to get data for the last recorded day and group by state
-last_day <- max(processeddata$date)
-data_last_day <- processeddata %>%
-  filter(date == last_day) %>%
-  group_by(state) %>%
-  summarise(total_cases = sum(total_cases, na.rm = TRUE),  # Sum of total cases
-            retail_and_recreation_percent_change_from_baseline = 
-              mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE)) %>%  # Average mobiltiy changes
-  mutate(state = tolower(state))  # Convert state names to lowercase
+print(density_cases)
 
-# Join the shapefile with the COVID-19 data
-us_map_data <- us_map_clean %>%
-  left_join(data_last_day, by = c("NAME" = "state"))
+figure_file = here("results","figures","density_cases.png")
+ggsave(filename = figure_file, plot=density_cases)
 
-# plot total COVID-19 cases to USA map on 10/15/2022
-cases_map <- ggplot(us_map_data) +
-  geom_sf(aes(fill = total_cases), color = "white") +
-  scale_fill_viridis_c(option = "plasma", trans = "log", labels = NULL) + 
-  labs(title = paste("Heatmap of Total COVID-19 Cases in the U.S. on", last_day),
-       fill = "Hotter = More Cases") +
-  theme_minimal() +
-  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
+density_cases_500 <- data %>%
+  filter(new_cases > 500) %>%
+  ggplot( aes(x=new_cases)) +
+    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
+  labs(title = "Distribution of Daily New Case Incidence (Over 500)", 
+       x = "Daily New Cases",
+       y = "Density") +
+  theme_bw()
 
-print(cases_map)
+print(density_cases_500)
 
-figure_file = here("results","figures","casemap.png")
-ggsave(filename = figure_file, plot=cases_map)
+figure_file = here("results","figures","density_cases_500.png")
+ggsave(filename = figure_file, plot=density_cases_500)
 
-# Create the heatmap for retail and recreation mobility changes
-mobility_map <- ggplot(us_map_data) +
-  geom_sf(aes(fill = retail_and_recreation_percent_change_from_baseline), color = "white") +  
-  scale_fill_viridis_c(option = "inferno", trans = "identity") +
-  labs(title = paste("Heatmap of Retail and Recreation Mobility Changes in the U.S. on", last_day),
-       fill = "Mobility Change") +
-  theme_minimal() +
-  theme(
-        legend.spacing.y = unit(1, "cm"),  
-        legend.box.spacing = unit(1, "cm"),  
-        legend.key.size = unit(1, "cm")) +  
-  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))
+print(kurtosis(data$new_cases))
+print(skewness(data$new_cases))
+```
+A histogram of population density was created.
+I hypothesize that population density plays a large role in case incidence.
+Population density is also skewed, however, there are some notable counties that do have large populationd densities.
+```{r}
+data_summary <- data %>%
+  group_by(county) %>%
+  summarize(pop_density = mean(pop_density, na.rm = TRUE))
 
-print(mobility_graph)
+histo_pop_density <- ggplot(data_summary, aes(x = pop_density)) + 
+  geom_histogram(bins = 30, fill = "#69b3a2", color = "black", alpha = 0.7) +
+  labs(title = "Histogram of Population Density (2-Year Average per County)", 
+       x = "Population Density (Calculated as Population Count / km squared)", y = "Frequency") + theme_bw()
 
-# save graph
-figure_file = here("results","figures","mobilitymap.png")
-ggsave(filename = figure_file, plot=mobility_map)
+print(histo_pop_density)
+
+figure_file = here("results","figures","histo_pop_density.png")
+ggsave(filename = figure_file, plot=histo_pop_density)
 ```
 
-This graph aims to plot total cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all total cases.
-```{r}
-# Create new data frame to have cases per day instead of state/county
-total_cases_df <- processeddata %>%
-  group_by(date) %>%
-  summarise(Cases = sum(total_cases, na.rm = TRUE)) %>%
-  ungroup()
+A 7-day rolling average was applied to the mobility data as the original iteration was too jagged. This code entails the creation of a data frame for 7 day averages.
 
-# Plot the total case counts over time
-case_count_graph <- ggplot(total_cases_df, aes(x = date, y = Cases)) +
-  geom_line(color = "blue") +
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  
-  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
-       x = "Date",
-       y = "Total Cases") +
-  theme_minimal()
-
-print(case_count_graph)
-
-# save graph
-figure_file = here("results","figures","casecountplot.png")
-ggsave(filename = figure_file, plot=case_count_graph)
-```
-This graph aims to plot new cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all new cases.
-```{r}
-# data frame to sum all cases on a given day for every location
-new_cases_df <- processeddata %>%
-  group_by(date) %>%
-  summarise(Cases = sum(new_cases, na.rm = TRUE)) %>%
-  ungroup()
-
-# Plot the new cases over time
-incidence_graph <- ggplot(new_cases_df, aes(x = date, y = Cases)) +
-  geom_line(color = "red") + 
-  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
-  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
-       x = "Date",
-       y = "New Cases per Day") +
-  theme_minimal()
-
-print(incidence_graph)
-
-# save graph
-figure_file = here("results","figures","incidenceplot.png")
-ggsave(filename = figure_file, plot=incidence_graph)
-
-```
-I aim to plot out the changes in mobility for each category as time progresses.
-Originally, it plotted out everyday, however, the line was far too jagged so I applied a weekly rolling average instead.
-I also had to create a dataframe to average values for a given day for every location.
 ```{r}
 
-# Create dataframe of average mobility for everywhere in the US per date
-us_average_mobility <- processeddata %>%
+# Create dataframe of average mobility per county per date
+ga_average_mobility <- data %>%
   group_by(date) %>%
   summarise(
-    total_mobility_changes = mean(average_mobility, na.rm = TRUE),
-    transit_mean = mean(transit_stations_percent_change_from_baseline, na.rm = TRUE),
-    parks_mean = mean(parks_percent_change_from_baseline, na.rm = TRUE),
     retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
@@
   mutate(
-    transit_mean_7day = rollmean(transit_mean, k = 7, fill = NA, align = "right"),
-    parks_mean_7day = rollmean(parks_mean, k = 7, fill = NA, align = "right"),
     retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
@@
 # Plot the 7-day averages for each mobility type
-mobility_graph <- ggplot(us_average_mobility, aes(x = date)) +
+mobility_graph <- ggplot(ga_average_mobility, aes(x = date)) +
   geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
-  geom_line(aes(y = transit_mean_7day, color = "Transit Stations")) +
   geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
   geom_line(aes(y = residential_mean_7day, color = "Residential")) +
-  geom_line(aes(y = parks_mean_7day, color = "Recreational")) +
+  geom_line(aes(y = retail_mean_7day, color = "Retail")) +
   labs(
-    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
+    title = "7-Day Moving Average of Percent Change in Mobility in Georgia Over Time",
     x = "Date",
@@
   ) +
-  theme_minimal()
+  theme_bw()
 
@@
 ```
+
+For better legiiblity, all 4 mobiltiy types were graphed separately.
+
+```{r}
+# Create one plot per mobility type
+p1 <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
+  geom_line(color = "steelblue") +
+  labs(title = "Grocery & Pharmacy", x = "Date", y = "7-Day Avg % Change") +
+  theme_bw() +
+  coord_cartesian(ylim = c(-50, 20))
+
+p2 <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
+  geom_line(color = "darkred") +
+  labs(title = "Workplaces", x = "Date", y = "7-Day Avg % Change") +
+  theme_bw() +
+  coord_cartesian(ylim = c(-50, 20))
+
+p3 <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
+  geom_line(color = "forestgreen") +
+  labs(title = "Residential", x = "Date", y = "7-Day Avg % Change") +
+  theme_bw() +
+  coord_cartesian(ylim = c(-50, 20))
+
+p4 <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
+  geom_line(color = "purple") +
+  labs(title = "Retail", x = "Date", y = "7-Day Avg % Change") +
+  theme_bw() +
+  coord_cartesian(ylim = c(-50, 20))
+
+# Arrange in a 2x2 grid
+mobility_grid <- (p1 | p2) / (p3 | p4)
+
+# Show it
+print(mobility_grid)
+
+# Save it
+figure_file = here("results", "figures", "mobility_grid.png")
+ggsave(filename = figure_file, plot = mobility_grid)
+
+```
+
+After visually inspecting the new cases over time graph, heteroscedastity and non-stationairty are of concern.
+Large outbreaks can skew these metrics and even when tested, could be misleading.
+Additionally, assumptions of homoscedastity are required for some tests of stationairty and some assumptions of normality are required for some tests of homoscdastity.
+Because of these concerns, I chose to focus on models like XGBoost and Random Forest which do not require these assumptions.
+
+```{r}
+library(scales)
+
+first_date <- min(data$date)
+last_date <- max(data$date)
+
+cases_graph <- case_data %>%
+  group_by(date) %>%
+  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
+  arrange(date) %>%
+  ggplot(aes(x = date, y = total_cases)) +
+    geom_area(fill = "#69b3a2", alpha = 0.5) +
+    geom_line(color = "#69b3a2") +
+    ylab("Total Cases (in GA)") +
+    scale_y_continuous(labels = label_number()) +
+   scale_x_date(
+      breaks = seq(from = first_date, to = last_date, by = "6 months"),  # Breaks every 3 months
+      labels = date_format("%b %Y")  # Format labels as month year (e.g., Jan 2020)
+    ) +
+    theme_bw() +
+  labs(title = "Total Cases (in GA) Over Time",
+       x = "Date (2020 - 2022)",
+       y = "Total Cases")
+
+print(cases_graph)
+
+figure_file = here("results","figures","cases_graph.png")
+ggsave(filename = figure_file, plot=cases_graph)
+
+incidence_graph <- data %>%
+  group_by(date) %>%
+  summarise(new_cases = sum(new_cases, na.rm = TRUE)) %>%
+  arrange(date) %>%
+  ggplot(aes(x = date, y = new_cases)) +
+    geom_area(fill = "#69b3a2", alpha = 0.5) +
+    geom_line(color = "#69b3a2") +
+    ylab("Daily New Cases") +
+     scale_x_date(
+      date_breaks = "3 months",  # Automatically set the breaks to every 3 months
+      date_labels = "%b %Y"      # Format labels as month-year (e.g., Jan 2020)
+    ) +
+    theme_bw() +
+  labs(title = "Daily New Cases (in GA) Over Time",
+       x = "Date (2020 - 2022)",
+       y = "Total Cases")
+
+print(incidence_graph)
+
+figure_file = here("results","figures","incidence_graph.png")
+ggsave(filename = figure_file, plot=incidence_graph)
+```
+
+This code chunk plots a scatter plot of a county's population count against its final total cases. 
+```{r}
+
+year_end_cases <- data %>%
+  mutate(year = year(date)) %>%
+  group_by(county, year) %>%
+  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
+  ungroup() %>%
+  select(county, year, total_cases)
+
+pop_density_info <- data %>%
+  mutate(year = year(date)) %>%
+  select(county, year, pop_density) %>%
+  distinct()
+
+plot_df <- left_join(year_end_cases, pop_density_info, by = c("county", "year"))
+
+year_end_cases_graph <- ggplot(plot_df, aes(x = pop_density, y = total_cases)) +
+  geom_point(color = "#69b3a2", alpha = 0.6) +
+  facet_wrap(~year) +
+  labs(
+    x = "Population Density",
+    y = "Total COVID Cases (End of Year)",
+    title = "Year-End COVID Cases vs. Population Density by County"
+  ) +
+  theme_bw()
+
+print(year_end_cases_graph)
+
+figure_file = here("results","figures","year_end_cases_graph.png")
+ggsave(filename = figure_file, plot=year_end_cases_graph)
+```
+Correlation Analysis
+
+This code chunk entails correlation testing through Spearman's and the creation of a matrix. First, lagged variables for case count was created for the data.
+```{r}
+
+data <- data %>%
+  group_by(county) %>%
+  arrange(date) %>%
+  mutate(
+    lag_1 = lag(new_cases, 1),
+    lag_7 = lag(new_cases, 7),
+    lag_14 = lag(new_cases, 14)
+  ) %>%
+  ungroup()
+
+data <- data %>% drop_na()
+
+# Create a data frame for matrix creation
+cor_df <- data %>%
+  select("New Cases" = new_cases, "1 Day Lag" = lag_1, "7 Day Lag" = lag_7, "14 Day Lag" = lag_14, "Population Density" = pop_density, "Retail Mobility" = retail_and_recreation_percent_change_from_baseline, "Workplace Mobility" = workplaces_percent_change_from_baseline, "Residential Mobility" = residential_percent_change_from_baseline, "Grocery Mobility" = grocery_and_pharmacy_percent_change_from_baseline,)
+
+# Calculate correlations using Spearman's rank correlation because data is non parametric
+cor_matrix <- cor(cor_df, method = "spearman")
+
+# Melt for visualization
+cor_matrix_melted <- melt(cor_matrix)
+
+# Create plot for matrix
+matrix <- ggplot(cor_matrix_melted, aes(x = Var1, y = Var2)) +
+  geom_point(aes(size = abs(value), color = value)) +
+  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
+  scale_color_gradient2(low = "#BFA2DB", high = "#92C7CF", mid = "white",
+                        midpoint = 0, limit = c(-1, 1), name = "Correlation") +
+  scale_size(range = c(2, 10), guide = "none") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(title = "Correlation Heatmap (Spearman's)", x = "Variables", y = "Variables")
+
+print(matrix)
+
+figure_file = here("results","figures","matrix.png")
+ggsave(filename = figure_file, plot=matrix)
+```

### 2025-04-18 products/manuscript/Manuscript.qmd
Work

--- 2025-03-14 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Utilizing Google's Mobility Report to understand Infectious Disease Dynamics"
+title: "Assessing the Added Predictive Power of Google's Community Mobility Report for COVID-19 Forecasting on a County-Level"
 author: "Vincent Nguyen"
@@
 
-{{< pagebreak >}}
+{{< Examining the relationship between human mobility and infectious disease spread is essential for forecasting disease incidence and informing public health interventions.
+This analysis investigates community mobility patterns and their association with daily new COVID-19 case incidence in Georgia.
+Using county-level case counts and mobility trends from Google’s Community Mobility reports, machine learning models were constructed and evaluated; models included linear regression, Random Forest, and XGBoost.
+To assess the predictive power of mobility change metrics, two feature sets were utilized.
+The first set, referred to as “baseline”, included features known to affect disease incidence, such as lagged cases and population density.
+The second set included both the previous set’s features and metrics measuring changes in mobility (recorded as percent change from baseline).
+Models were trained and tuned using time-series blocked cross-validation and performance was evaluated R-squared, RMSE, and MAE metrics.
+Results indicated that models with mobility features could achieve modest improvements in RMSE and MAE, however, did not substantially increase R-squared, suggesting limited explanatory power.
+These findings suggest that while mobility data may slightly enhance short-term predictive accuracy, their overall predictive power in explaining COVID-19 incidence may be limited in this context.
+>}}
 
-# Introduction
+# 2 Introduction
+
+## 2.1 COVID-19 Pandemic
 
@@
 
+## 2.2 Mobility Dynamics
+
+Human mobility, a key driver[of respiratory disease transmi](https://www.nature.com/articles/s41598-024-64230-1)ssion, shifted significantly during the pandemic in response to public health policies and disease risk perception.
+Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lockdowns.
+Lockdowns involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
+Looking at the Wikipedia page shows a lack of prior history implementing lockdowns.
+Beyond these measures which lasted only a few weeks in Georgia, individual risk perception played into the compliance of other preventative behaviors (masking, social distancing, etc.).
+The[variability of individual risk perception](https://pmc.ncbi.nlm.nih.gov/articles/PMC9028425/#:~:text=The%20results%20indicate%20that%20a,%2C%20geographical%20factors%2C%20and%20timing.)has led to complex mobility patterns during the pandemic, for example, surges in mobility amidst large case outbreaks.
+In order to analyze and quantify this relationship, machine learning models were adapted.
+
 \[insert stats about pandemic\]
 
-### 2.2 Lockdown and Social Distancing
+## 2.3 Machine Learning
 
-Nonpharmacological interventions are among the measures public health officials utilize to help curb the transmission of infectious diseases.
-In lieu of medication or vaccinations, these interventions can be a cost-effective and population-based solution.
-These interventions include the use of isolation, the subdivision of an infected individual from the general population, and social distancing, a form of voluntary or mandated physical distancing from others.
-Another common measure is the use of quarantine, which is the restriction of infected or at-risk individuals from traveling to help minimize the risk of healthy individuals; it is intended that the period of transmissibility has passed before lifting the quarantine (Jacobsen and Jacobsen 2020)[@jacobsen_2020_statewide].
+[Machine Learning](https://pmc.ncbi.nlm.nih.gov/articles/PMC9949554/) refers to a class of data-driven algorithms that aim to analyze associations and relationships found in data.
+These techniques can be supervised or unsupervised; in supervised ML, models are given labeled inputs/features to derive and predict an output.
+These models aim to also approximate the relationship between outputs and inputs, quantify predictions, or approximate classification tasks.
+Beyond regular statistical analysis, machine learning methods offer powerful tools for forecasting infectious disease trends by identifying complex, nonlinear relationships between predictors and outcomes.
+S[ome common m](https://pmc.ncbi.nlm.nih.gov/articles/PMC8219638/)odels in disease forecasting include ARMA, ARIMA, LASSO, XGBoost, and various neural network techniques.
+Several autho[rs](https://pmc.ncbi.nlm.nih.gov/articles/PMC8731233/)have utilized case counts, estimations, demographics, and more to forecast disease trends.
+However, the predictive value of real-time mobility data, particularly in the context of an evolving pandemic, has not been assessed through modeling.
+This study aims to explore whether mobility dynamics, form Google’s Community Mobility Reports, can enhance predictive performance when modeling COVID-19 transmission at the county level.
 
-Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lockdowns.
-Lockdowns involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
-By April 3rd, 2020, just a few weeks after California’s state-wide mandate, 41 states joined the directive in restricting mobility (Mervosh et al.
-2020)[@mervosh_2020_see].
-Many polictians and citizens questioned the measure citing a lack of evidence for its efficacy; looking at the Wikipedia page shows a lack of prior history implementing lockdowns (Zeleny et al.
-2021)[@zeleny_2021_a].
+# 3 Data
 
-### 2.3 Mobility Dynamics
+## 3.1 Google’s Community Mobility Report
 
-According to Paltra et al.
-2021, mobility reductions on COVID-19 can be quadratic in reducing the transmission.
-For example, the authors state a reduction of out-of-home duration to 70% can result in a transmission rate of 0.49.
-\[finish\]
+During the pandemic, Google began to collect aggregated, anonymized data from users utilizing Google products (apps, phones, etc.) to track changes in mobility.
+Within the data, Google measures mobility as a percent change difference from baseline measurements; for example, a -45% change in retail and recreation mobility indicates a 45% reduction in movement to those categorical locations.
+These measurements are stratified by county, however, with the implication of technology use, are limited to counties with enough Google users.
+Additionally, two mobility metrics, transit stations and recreational parks, were omitted in this analysis due to incompleteness.
 
-## Description of data and data source
+## 3.2 John Hopkins University COVID-19 Case Data
 
-This project explores mobility dynamics, as measured by [Google's Community Mobility Reports](https://www.google.com/covid19/mobility/), in an infectious disease context to understand and forecast disease spread.
-By integrating time-series infection data (provided by [Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19)), the analysis aims to simulate disease outbreak scenarios and provide insights into optimal intervention strategies for mitigating future outbreaks.
-Additionally, the [County Health Ranking](https://www.countyhealthrankings.org/) is used to assess these dynamics within the context of a county's health care access and performance.\
-\
-Google's Community Mobility Reports are anonymized insights collected from Google products, such as maps, that detail population movement trends across various categories, such as retail stores, workplaces, residencies, and more.\
-\
 The COVID-19 Data Repository by the Center for Systems Science and Engineering at Johns Hopkins University is a comprehensive data set that tracks global COVID-19 cases, recoveries, and deaths.
-It was updated regularly and compiles data from various official sources to support research.\
+This data was recorded daily for several years across every county in the US.
 
-The County Health Ranking is a data set that provides a comprehensive overview of various health factors and outcomes across U.S.
-counties.
-Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more.
-\[Not sure about the inclusion of this\]
+## 3.3 Final Data-set
 
-## Questions/Hypotheses to be addressed
-
-How have mobility dynamics influenced the spread of COVID-19 in the U.S. during 2020-2022?\
-\
-Can mobility data, in combination with COVID-19 case counts, be used to forecast future infection trends?
-
-How do socio-economic and health factors influence the relationship between mobility patterns and COVID-19 infection rates across U.S. counties?
-
-Investigate if counties with similar health rankings but different mobility trends experienced different outbreak severities, potentially revealing the effectiveness of local interventions
-
-Examine ifthere was lag time between changes in mobility and subsequent changes in infection rates
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
+After data wrangling and cleaning for completeness, 44 counties were included in the analysis.
+An 80% completeness of data per county was required for inclusion into analysis, resulting in the removal of 115 counties.
+Many of Georgia’s counties are rural and as such, lacked many mobility metrics.
+The data was filtered from March 14t 2020 to March 14 2022.
+This was selected specifically as March 14th marks the date of Governor Brian Kemp’s announcement of Georgia's Public Health State of Emergency.
 
@@
 
-# Methods
+# 4 Methods
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+## 4.1 Model Selection
 
-## Schematic of workflow
+As informed by Alfred and Obit 2021, machine learning models have wide application in forecasting outbreaks and disease incidence.
+In their review, they outline several common applications of regression and classification models.
+Their review, along with class content and data structure, informed model selection.
+Previous iterations of the analysis indicated that the data was non-normal and non-stationary which violated assumptions in common time series models like ARIMA.
+Linear regression was chosen as a standard model while Random Forest and XGBoost were chosen for their lack of assumptions required for analysis.
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+### Linear Regression
 
-## Data acquisition
+A statistical modeling method used to quantify the relationship between a continuous outcome variable and one or more predictors.
+In its simplest form, linear regression estimates, or fits, a linear equation explaining the relationship by minimizing the sum of squared differences between residuals.
+Linear regression assumes
 
-*COVID-19 case data was provided by the Center of Systems and Science and Engineering at JHU and mobility data was provided by Google's Community Mobility Report.*
+### Random Forest
 
-## Data import and cleaning
+(https://www.stat.berkeley.edu/\~breiman/randomforest2001.pdf)\
 
-Since the pandemic spanned across 3 years, data on mobility was collected from 2020 to 2022.
-The data on mobility is quite extensive and is stratified by county, resulting in the large amount of mobility.
-The mobility data measures changes in mobility patterns across different key areas, such as retail, workplace, residential, and more.
+A ensemble machine learning method that builds multiple decision trees and aggregates their predictions to improve accuracy and reduce overfitting.
+A decision tree is a modeling technique that partitions  data into subsets based on the input features; it aims to minimize variance of the outcome variable in each subgroup.
+These splits form a tree-like structure whe -.
+Decision trees are powerful and can capture nonlinear complex relationships, however, can overfit training data.  Random forests are an average of these trees to produce a final prediction and can produce more stable/accurate models as a result.
 
-The time series data has 3342 observations of 1154 variables. Each row represents a different city. The columns record cases as days progress and locational data like longitude and latitude.
+### XGBoost
 
-```{r}
-library(here)
+(https://arxiv.org/abs/1603.02754): Also known as Extreme Gradient Boosting, XGBoost is a machine learning algorithm based on decision tress, however, instead of averaging trees, it tries to correct errors made by previous ones.
+This technique is known as boosting.
+In simple terms, XGBoost makes predictions with a single tree and looks at the errors and creates a second tree focused on minimizing the errors.
 
-# importing data
-mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
-mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
-mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))
+Two sets of predictors were chosen to assess the effect mobility changes have on case incidence.
+The first set, the baseline predictors, included population density and 3 spaced out lagged variables for case counts (1 day, 7 days, and 14 days).
+These were chosen because of  their [known influence](https://pmc.ncbi.nlm.nih.gov/articles/PMC8983058/#:~:text=This%20study%20has%20shown%20that,data%20for%20COVID%E2%80%9019%20cases.) on disease transmission.
+The second set, the mobility predictors, included population density, the 3 spaced out lagged variables, and mobility indicators from Google’s report (e.g., workplace, grocery, retail, transit).
 
-time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))
+## 4.2 Cross Validation
 
-chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))
+To evaluate model performance while accounting for the temporal structure of the data, a time-blocked cross-validation strategy was employed.
+The dataset was partitioned into eight non-overlapping temporal folds, each representing a three-month interval spanning March 2020 to March 2022.
+Each observation (a day for a corresponding county) was assigned a fold based on its calendar month.
+Grouped k-fold cross-validation using these time-based identifiers were used to ensure data from the future could not leak into the training set.
 
-```
+## 4.3 Model Evaluation and Metrics
 
-## Statistical analysis
+### R-squared
+
+A statistical measure that explains the proportion of the data variation that can be explained by the model.
+A higher R-squared value generally is indicative of a better-fitting model, although could be misleading when over fitting occurs.
+
+### RMSE
+
+Root Mean Square Deviation is a statistical measure that averages the magnitude of the errors in the model’s predictions.
+
+Arithmetically, it is the square root of the average squared differences between the predicted and actual values. A lower RMSE indicates a better predictive performance.
+
+### MAE
+
+Mean Absolute Error is a statistical measure that also measures prediction error, however, utilizes the absolute difference between predicted and actual values instead of the squaring the errors.
+It is calculated by taking the average absolute difference between each predicted and actual value.
+Because of its computation, MAE is less sensitive to large errors, like outliers, than RMSE.
+MAE may be a more reliable metric for this data set due to the presence of large outbreaks that can inflate the errors.
+
+Linear regression, Random Forest, and XGBoost models were trained using the baseline set of features to predict COVID-19 case incidence.
+These baseline models served as reference points to evaluate the added predictive value of mobility-based features and to assess the potential association between changes in population mobility and COVID-19 case incidence.
+
+After the initial RF and XGBoost models were made, their hyperparameters were tuned to improve predictive performance.
+
+## 4.4 Software
+
+The analysis was conducted on RStudio 4.3.2 on Windows 11.
+The following R packages were used: [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html), [broom](https://cran.r-project.org/web/packages/broom/index.html), [here](https://cran.r-project.org/web/packages/here/index.html), [glmnet](https://cran.r-project.org/web/packages/glmnet/index.html), [MASS](https://cran.r-project.org/web/packages/MASS/index.html), [tidymodels](https://cran.r-project.org/web/packages/tidymodels/index.html), [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html), [rsample](https://cran.r-project.org/web/packages/rsample/index.html), [parsnip](https://cran.r-project.org/web/packages/parsnip/index.html), [future](https://cran.r-project.org/web/packages/future/index.html), [vip](https://cran.r-project.org/web/packages/vip/index.html), [zoo](https://cran.r-project.org/web/packages/zoo/index.html), [patchwork](https://cran.r-project.org/web/packages/patchwork/index.html), [e1071](https://cran.r-project.org/web/packages/e1071/index.html), [scales](https://cran.r-project.org/web/packages/scales/index.html), [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/index.html), [corrplot](https://cran.r-project.org/web/packages/corrplot/index.html), [reshape2](https://cran.r-project.org/web/packages/reshape2/index.html), [tidyr](https://cran.r-project.org/web/packages/tidyr/index.html), [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html), [readxl](https://cran.r-project.org/web/packages/readxl/index.html), [stringr](https://cran.r-project.org/web/packages/stringr/index.html), [skimr](https://cran.r-project.org/web/packages/skimr/index.html), [tigris](https://cran.r-project.org/web/packages/tigris/index.html), and [sf](https://cran.r-project.org/web/packages/sf/index.html).
 
@@
 
-# Results
+# 5 Results
 
-## Exploratory/Descriptive analysis
-
-To gain a clearer perspective on the overall scale of the pandemic, we visualized the total number of COVID-19 cases over time.
-This plot provides a overview of how case counts evolved.The graph reveals a large spike in cases in the beginning of 2022.
-This does correspond with figures further.
+## 5.1 Exploratory/Descriptive analysis
 
 ```{r}
-#| label: casecountplot
-#| fig-cap: "Total COVID-19 Cases overtime"
+#| label: cases_graph
+#| fig-cap: "Total COVID-19 Cases in Georgia over time"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","casecountplot.png"))
+knitr::include_graphics(here("results","figures","cases_graph.png"))
 ```
-
-To gain a clearer perspective on the overall scale of the pandemic, we visualized new caes of COVID-19 over time.
-This plot provides a overview of how case counts and infection evolved.The graph reveals a large spike in cases in the beginning of 2022 which corresponds with the total case count graph.
 
@@
 #| label: incidenceplot
-#| fig-cap: "Daily New COVID-19 Cases overtime in the U.S."
+#| fig-cap: "Daily New COVID-19 Cases overtime in the GA (2020-2022)"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","incidenceplot.png"))
+knitr::include_graphics(here("results","figures","incidence_graph.png"))
 ```
-
-To gain a deeper understanding of how mobiltiy patterns shifted throughout the pandemic, weekly changes in mobility were visualized.
-This plot highlights fluctuations in movement across various sectors.
-Mobility was increased towards residential and recreational areas which warrants further exploration.
-This can help offer insight into how behavior towards COVID-19 restriction interventions evolved over time.
 
@@
 #| label: mobiltiygraph
-#| fig-cap: "Percent Changes in Mobility per week in the U.S."
+#| fig-cap: "Percent Changes in Mobility per week in GA"
 #| echo: FALSE
@@
 
-As hypothesized, an rise in COVID-19 case count is expected to lead to a decline in mobility to non-residential locations, as individuals alter their behavior in resposne to increasing risks.
-To explore this idea, total case count and changes in mobility were visualzied across the United States.\
-\
-
 ```{r}
-#| label: casemap
-#| fig-cap: "Total COVID-19 Cases on 10/15/2022 mapped to the U.S."
+#| label: mobilitygrid
+#| fig-cap: "Percent Changes in Mobility per week in GA"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","casemap.png"))
+knitr::include_graphics(here("results","figures","mobility_grid.png"))
 ```
@@
 
-Analysis of the mapped data reveals a trend: states with higher case counts generally experienced more significant decreases in mobility.
-This suggests a correlation between rising COVID-19 case count and a greater reduction in movement to non-residential locations.
+## Correlation Analysis
 
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
+```{r}
+#| label: correlation
+#| fig-cap: "Correlation Matrix of Predictors using Spearman's"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","matrix.png"))
+```
+ In addition to visual analysis, correlation analysis was done.
+The non-parametric method, Spearman’s rank correlation coefficient, was used to assess correlation among predictors.
+Analysis reveals that lagged case counts were strongly correlated with daily incidence while in contrast mobility variables demonstrated weak relationships.
+This is suggestive of lagged variables's strong predictive power and mobility metric’s weak power.
 

### 2025-04-21 code/analysis-code/statisitical-analysis.qmd
Revert "Work"
This reverts commit 63ab86524d9b8d1ab4357a823b03d789a625d851.

--- 2025-04-04 code/analysis-code/statisitical-analysis.qmd
@@
 
-Load needed packages. make sure they are installed.
-
+Load needed packages
 ```{r}
@@
 library(rsample)
-library(poissonreg)
+library(tibble)
 library(parsnip)
 library(future)
+library(vip)
+library(lubridate)
+library(patchwork)
 ```
 
+Load in processed data and set eed
 ```{r}
-#path to data
-#note the use of the here() package and not absolute paths
+# path to data
+# note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","processeddata.rds")
 
-#load data. 
+# load data 
 data <- readRDS(data_location)
 
-# NOTES TO SELF
-# - change to QMD file (personally prefer)
-# - work on other analysis (lag-time & quarantine fatigue? more research)
-# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
-class(data$new_cases)
+# set seed
+rngseed = 1234
+set.seed(rngseed)
 ```
 
+Creation of lagged predictors
 ```{r}
-# Intial Recipe setting
-recipe <- recipe(new_cases ~ pop_density + transit_stations_percent_change_from_baseline + 
-                     parks_percent_change_from_baseline + 
+# Creation of lagged cases and lagged mobility data
+data <- data %>%
+  group_by(county) %>%
+  arrange(date) %>%
+  mutate(
+    lag_1 = lag(new_cases, 1),
+    lag_7 = lag(new_cases, 7),
+    lag_14 = lag(new_cases, 14),
+    retail_lag7 = lag(retail_and_recreation_percent_change_from_baseline, 7),
+    grocery_lag7 = lag(grocery_and_pharmacy_percent_change_from_baseline, 7),
+    workplace_lag7 = lag(workplaces_percent_change_from_baseline, 7),
+    residential_lag7 = lag(residential_percent_change_from_baseline, 7)
+  ) %>%
+  ungroup()
+
+# remove NAs as result of lagging
+data <- data %>% drop_na()
+```
+
+Implementation of rolling window cross-validation
+```{r}
+# Arrange by date
+data <- data %>%
+  arrange(date, county)
+
+# Define testing set as last 30 days for 44 counties
+test_days <- 30
+last_date <- max(data$date)
+test_start_date <- last_date - days(test_days - 1)
+
+# Define train and test split through dates
+train_data <- data %>% filter(date < test_start_date)
+test_data  <- data %>% filter(date >= test_start_date)
+
+# Create rolling window folds for cross-validation 
+rolling_folds <- rolling_origin(
+  data = train_data,
+  initial = 120 * 44,  # ~4 months of training
+  assess  = 30 * 44,   # ~1 month of validation
+  skip    = 15 * 44,   # roll forward 15 days
+  cumulative = FALSE
+)
+
+# Check date ranges for each fold
+train_ranges <- purrr::map(rolling_folds$splits, ~range(pull(training(.x), date)))
+
+# Print the ranges for each fold
+train_ranges
+
+# View fold summary
+fold_sizes <- purrr::map_dfr(rolling_folds$splits, function(split) {
+  tibble(
+    train_n = nrow(training(split)),
+    test_n  = nrow(testing(split))
+  )
+})
+
+print(fold_sizes)
+
+# Check date ranges
+range(train_data$date)
+range(test_data$date)
+
+```
+
+Recipe setting for modeling
+```{r}
+# Recipe for baseline model
+recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data)
+
+# Recipe with all predictors
+recipe <- recipe(new_cases ~ pop_density +
                   retail_and_recreation_percent_change_from_baseline +
@@
                    residential_percent_change_from_baseline +
-                     health_factor_rank, data = data)
+                     lag_1 + lag_7 + lag_14, data = data)
+
+# Recipe with lagged versions of mobility
+recipe_lag <- recipe(new_cases ~ pop_density + retail_lag7 + grocery_lag7 + residential_lag7 + workplace_lag7 + lag_1 + lag_7 + lag_14, data = data)
+```
+
+Modeling
+
+The basic workflow is to fit a baseline model, fit the mobility model, and then fit the lagged model.
+
+First model are linear regression models.
+This code chunk fits a baseline linear regression model.
+```{r}
+# Define the Lasso model with tuning parameters
+lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
+  set_engine("glmnet")
+
+# Create the Lasso workflow with baseline recipe
+base_tuned_lasso_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(lasso_spec)
+
+# Define a grid for tuning the penalty
+lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)
+
+# Tune the Lasso model using rolling origin cv
+base_tuned_lasso_res <- tune_grid(
+  base_tuned_lasso_wf,
+  resamples = rolling_folds,
+  grid = lasso_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything", save_pred = TRUE)
+)
+
+# View tuning results
+autoplot(base_tuned_lasso_res)
+base_tuned_lasso_metrics <- collect_metrics(base_tuned_lasso_res)
+print(base_tuned_lasso_metrics)
+
+# Select best penalty based on R-squared
+base_tuned_lasso_best <- select_best(base_tuned_lasso_res, metric = "rsq")
+print(base_tuned_lasso_best)
+
+# Finalize the Lasso model with the selected penalty
+base_final_lasso_spec <- linear_reg(
+  penalty = base_tuned_lasso_best$penalty,
+  mixture = 1
+) %>%
+  set_engine("glmnet")
+
+# Final workflow with tuned Lasso model
+base_final_lasso_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_final_lasso_spec)
+
+# Fit the final Lasso model on training data
+base_final_lasso_fit <- base_final_lasso_wf %>%
+  fit(data = train_data)
+
+# Evaluate the final Lasso model with rolling CV
+base_final_lasso_res <- fit_resamples(
+  base_final_lasso_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# View resampling performance for the final Lasso model
+collect_metrics(base_final_lasso_res)
+
+# Predict on test data using the final Lasso model
+base_final_lasso_test_preds <- predict(base_final_lasso_fit, new_data = test_data)
+
+# Combine predictions with actual values
+base_final_lasso_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = base_final_lasso_test_preds$.pred)
+
+# Compute test metrics for the final Lasso model
+base_final_lasso_test_metrics <- base_final_lasso_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
+
+# View the test set performance for the final Lasso model
+print(base_final_lasso_test_metrics)
+
+# View variable importance for the Lasso model
+vip(base_final_lasso_fit$fit$fit)
 
@@
 
-First, I created a linear regression model as the baseline standard.
-This model performs very poorly as it has a low R-squared, indicating low predictive power.
-The model does mark all variables of the model to be significant.Additionally, the AIC is very extremely high.
+This code chunk fits a linear regression model with mobility predictors.
+```{r}
+# Create the Lasso workflow
+lasso_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(lasso_spec)
 
-```{r}
-# Create linear regression model
-lm_spec <- linear_reg() %>%
-  set_engine("lm")
+# Tune the Lasso model using time-blocked CV
+lasso_res <- tune_grid(
+  lasso_wf,
+  resamples = rolling_folds,
+  grid = lasso_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(save_pred = TRUE)
+)
 
-lm_wf <- workflow() %>%
-  add_recipe(recipe)
+# View tuning results
+autoplot(lasso_res)
+lasso_metrics <- collect_metrics(lasso_res)
+print(lasso_metrics)
 
-lm_fit <- lm_wf %>%
-  add_model(lm_spec) %>%
-  fit(data = data)
+# Select best penalty by R-squared
+lasso_best <- select_best(lasso_res, metric = "rsq")
+print(lasso_best)
 
-lm_fit %>%
-  extract_fit_parsnip() %>%
-  tidy()
-
-lm_fit %>%
-  extract_fit_parsnip() %>%
-  broom::glance()
-
-lm_preds<- predict(lm_fit, new_data = data)
-
-lm_metrics <- tibble(truth = data$new_cases, predicted = lm_preds$.pred) %>%
-  metrics(truth = truth, estimate = predicted)
-lm_metrics
-```
-
-I utilized Lasso here to hopefully bring down the RMSE and bring up the rsquared.
-It seems Lasso has no effect on model performance as the data lacks multicolinearity and does not have irrelevant predictors.
-```{r}
-# Set method for LASSO
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
+# Finalize the Lasso model with selected lambda
+final_lasso_spec <- linear_reg(
+  penalty = lasso_best$penalty,
+  mixture = 1
+) %>%
   set_engine("glmnet")
 
-# Insert recipe
-lasso_wf <- workflow() %>%
-  add_recipe(recipe)
+# Final workflow with tuned Lasso model
+final_lasso_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(final_lasso_spec)
 
-lasso_fit <- lasso_wf %>%
-  add_model(lasso_spec) %>%
-  fit(data = data)
+# Fit final model on training data
+final_lasso_fit <- final_lasso_wf %>%
+  fit(data = train_data)
 
-lasso_fit %>%
-  extract_fit_parsnip() %>%
-  tidy()
+# Evaluate final model with rolling CV
+final_lasso_res <- fit_resamples(
+  final_lasso_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
 
-lasso_fit %>%
-  extract_fit_parsnip() %>%
-  broom::glance()
+# View resampling performance
+collect_metrics(final_lasso_res)
 
-lasso_preds<- predict(lasso_fit, new_data = data)
+# Predict on test data
+final_lasso_test_preds <- predict(final_lasso_fit, new_data = test_data)
 
-lasso_metrics <- tibble(truth = data$new_cases, predicted = lasso_preds$.pred) %>%
-  metrics(truth = truth, estimate = predicted)
-lasso_metrics
-```
+# Combine predictions with actuals
+final_lasso_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = final_lasso_test_preds$.pred)
 
-After the last two models performed poorly, the use of random forest seemed rational as this model can capture more complex relationships. 
-```{r}
+# Compute test metrics
+final_lasso_test_metrics <- final_lasso_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
 
-rngseed = 1234
+# View test set performance
+print(final_lasso_test_metrics)
 
-forest_spec <- rand_forest(mode = "regression") %>%
-  set_engine("ranger", seed = rngseed)
-
-# Insert recipe
-forest_wf <- workflow() %>%
-  add_recipe(recipe)
-
-# Create model with all predictors and random forest
-forest_fit <- forest_wf %>%
- add_model(forest_spec) %>%
-  fit(data = data)
-
-forest_preds <- predict(forest_fit, new_data = data)
-forest_preds <- tibble(truth = data$new_cases, predicted = forest_preds$.pred)
-forest_rmse <- forest_preds %>%
-  metrics(truth = truth, estimate = predicted)
-print(forest_rmse)
+# View variable importance
+vip(final_lasso_fit$fit$fit)
 ```
@@
 ```{r}
-# Tuning forest
+# Create the Lasso workflow
+lag_lasso_wf <- workflow() %>%
+  add_recipe(recipe_lag) %>%
+  add_model(lasso_spec)
+
+# Tune the Lasso model using time-blocked CV
+lag_lasso_res <- tune_grid(
+  lag_lasso_wf,
+  resamples = rolling_folds,
+  grid = lasso_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(save_pred = TRUE)
+)
+
+# View tuning results
+autoplot(lag_lasso_res)
+lag_lasso_metrics <- collect_metrics(lag_lasso_res)
+print(lag_lasso_metrics)
+
+# Select best penalty by R-squared
+lag_lasso_best <- select_best(lag_lasso_res, metric = "rsq")
+print(lag_lasso_best)
+
+# Finalize the Lasso model with selected lambda
+lag_final_lasso_spec <- linear_reg(
+  penalty = lag_lasso_best$penalty,
+  mixture = 1
+) %>%
+  set_engine("glmnet")
+
+# Final workflow with tuned Lasso model
+lag_final_lasso_wf <- workflow() %>%
+  add_recipe(recipe_lag) %>%
+  add_model(lag_final_lasso_spec)
+
+# Fit final model on training data
+lag_final_lasso_fit <- lag_final_lasso_wf %>%
+  fit(data = train_data)
+
+# Evaluate final model with rolling CV
+lag_final_lasso_res <- fit_resamples(
+  lag_final_lasso_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# View resampling performance
+collect_metrics(lag_final_lasso_res)
+
+# Predict on test data
+lag_final_lasso_test_preds <- predict(lag_final_lasso_fit, new_data = test_data)
+
+# Combine predictions with actuals
+lag_final_lasso_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = lag_final_lasso_test_preds$.pred)
+
+# Compute test metrics
+lag_final_lasso_test_metrics <- lag_final_lasso_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
+
+# View test set performance
+print(lag_final_lasso_test_metrics)
+
+# View variable importance
+vip(lag_final_lasso_fit$fit$fit)
+
+```
+```{r}
+# plot baseline lasso model prediction vs actual
+plot_base_lasso <- base_final_lasso_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "darkgreen") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "Base Lasso",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot mobility lasso model prediction vs actual
+plot_lasso <- final_lasso_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "steelblue") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "Lasso",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot lagged lasso model prediction vs actual
+plot_lag_lasso <- lag_final_lasso_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "purple") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Lag Lasso",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+lasso_graphs <- plot_base_lasso + plot_lasso + plot_lag_lasso +
+  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
+  theme(
+    plot.margin = margin(10, 10, 10, 10),
+    plot.title = element_text(size = 14, hjust = 0.5),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+
+print(lasso_graphs)
+
+figure_file = here("results", "figures", "lasso_graphs.png")
+ggsave(
+  filename = figure_file,
+  plot = lasso_graphs,
+  width = 16, height = 5, dpi = 300
+)
+
+```
+
+Second, random forest models are tuned, trained, and evaluated. Best models are chosen through R-squared.
+
+This code chunk fits and tunes a baseline RF model with no mobility predictors.
+```{r}
+# Define the Random Forest model with tuning parameters
 forest_spec <- rand_forest(
   mode = "regression", 
-  trees = 300, 
   mtry = tune(),           
-  min_n = tune()
+  min_n = tune(),          
+  trees = 500              
 ) %>%
-  set_engine("ranger")
+  set_engine("ranger", seed = rngseed)
 
-# Create grid with set parameters
+# Create grid of hyperparameters for tuning
 forest_grid <- grid_regular(
-  mtry(range = c(1, 7)),  
-  min_n(range = c(1, 21)),  
-  levels = 7  
+  mtry(range = c(1, 4)),  
+  min_n(range = c(1, 50)), 
+  levels = 10
 )
 
-# Set seed
-set.seed(1234)
+# Create baseline workflow with baseline recipe
+base_tuned_rf_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(forest_spec)
 
-# 5 fold crossvalidation repeated 5 times
-# Note: Limitation of computing power so doing rolling window would not work
-cv_folds <- vfold_cv(data, v = 5, repeats = 5)
-plan(multisession, workers = 4)
-
-# Tune the forest
-# This step took 20-ish minutes
-forest_tune_res <- tune_grid(
-  forest_wf %>% add_model(forest_spec),
-  resamples = cv_folds,  
+# Tune the Random Forest model using time-blocked CV (folds)
+base_tuned_rf_res <- tune_grid(
+  base_tuned_rf_wf,
+  resamples = rolling_folds,                     
   grid = forest_grid,
-  metrics = metric_set(rmse, rsq),
+  metrics = metric_set(rmse, rsq, mae),
   control = control_grid(parallel_over = "everything")
@@
 
-# Plot results
-autoplot(forest_tune_res)
+# View tuning results
+autoplot(base_tuned_rf_res)
+base_tune_rf_metrics <- collect_metrics(base_tuned_rf_res)
+print(base_tune_rf_metrics)
 
-forest_tune_res %>%
-  collect_metrics()
+# Select best hyperparameters by R-squared
+base_tuned_rf_best <- select_best(base_tuned_rf_res, metric = "rsq")
 
-best_params <- select_best(forest_tune_res, metric = "rmse")
-best_params
+# Finalize model with best parameters
+base_final_rf_spec <- rand_forest(
+  mode = "regression",
+  mtry = base_tuned_rf_best$mtry,
+  min_n = base_tuned_rf_best$min_n,
+  trees = 500
+) %>%
+  set_engine("ranger", seed = rngseed)
 
-best_params$trees <- 500
+# Final workflow with tuned model
+base_final_rf_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_final_rf_spec)
+
+# Fit final model on training data
+base_final_rf_fit <- base_final_rf_wf %>%
+  fit(data = train_data)
+
+# Evaluate final model with rolling CV
+base_final_rf_res <- fit_resamples(
+  base_final_rf_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# View resampling performance
+collect_metrics(base_final_rf_res)
+
+# Predict on test data
+base_final_rf_test_preds <- predict(base_final_rf_fit, new_data = test_data)
+
+# Combine predictions with actuals
+base_final_rf_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = base_final_rf_test_preds$.pred)
+
+# Compute test metrics
+base_final_rf_test_metrics <- base_final_rf_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
+
+# View test set performance
+print(base_final_rf_test_metrics)
+
+```
+
+This code chunk fits and tunes a RF model with mobility predictors.
+```{r}
+# Create workflow with baseline recipe
+tuned_rf_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(forest_spec)
+
+# Tune the Random Forest model using time-blocked CV (folds)
+rf_tune_res <- tune_grid(
+  tuned_rf_wf,
+  resamples = rolling_folds,                     
+  grid = forest_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# View tuning results
+autoplot(rf_tune_res)
+rf_tune_metrics <- collect_metrics(rf_tune_res)
+print(rf_tune_metrics)
+
+# Select best hyperparameters by R-squared
+rf_best <- select_best(rf_tune_res, metric = "rsq")
+
+# Finalize model with best parameters
+final_rf_spec <- rand_forest(
+  mode = "regression",
+  mtry = rf_best$mtry,
+  min_n = rf_best$min_n,
+  trees = 500
+) %>%
+  set_engine("ranger", seed = rngseed)
+
+# Final workflow with tuned model
+final_rf_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(final_rf_spec)
+
+# Fit final model on training data
+final_rf_fit <- final_rf_wf %>%
+  fit(data = train_data)
+
+# Evaluate final model with rolling CV
+final_rf_res <- fit_resamples(
+  final_rf_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# View resampling performance
+collect_metrics(final_rf_res)
+
+# Predict on test data
+final_rf_test_preds <- predict(final_rf_fit, new_data = test_data)
+
+# Combine predictions with actuals
+final_rf_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = final_rf_test_preds$.pred)
+
+# Compute test metrics
+final_rf_test_metrics <- final_rf_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
+
+# View test set performance
+print(final_rf_test_metrics)
+```
+
+This code chunk fits and tunes a RF model with lagged mobility predictors.
+```{r}
+# Create lag workflow with baseline recipe
+lag_tuned_rf_wf <- workflow() %>%
+  add_recipe(recipe_lag) %>%
+  add_model(forest_spec)
+
+# Tune the Random Forest model using time-blocked CV (folds)
+lag_tuned_rf_res <- tune_grid(
+  lag_tuned_rf_wf,
+  resamples = rolling_folds,                     
+  grid = forest_grid,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# View tuning results
+autoplot(lag_tuned_rf_res)
+lag_tune_rf_metrics <- collect_metrics(lag_tuned_rf_res)
+print(lag_tune_rf_metrics)
+
+# Select best hyperparameters by R-squared
+lag_tuned_rf_best <- select_best(lag_tuned_rf_res, metric = "rsq")
+
+# Finalize model with best parameters
+lag_final_rf_spec <- rand_forest(
+  mode = "regression",
+  mtry = lag_tuned_rf_best$mtry,
+  min_n = lag_tuned_rf_best$min_n,
+  trees = 500
+) %>%
+  set_engine("ranger", seed = rngseed)
+
+# Final workflow with tuned model
+lag_final_rf_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(lag_final_rf_spec)
+
+# Fit final model on training data
+lag_final_rf_fit <- lag_final_rf_wf %>%
+  fit(data = train_data)
+
+# Evaluate final model with rolling CV
+lag_final_rf_res <- fit_resamples(
+  lag_final_rf_wf,
+  resamples = rolling_folds,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_resamples(save_pred = TRUE)
+)
+
+# View resampling performance
+collect_metrics(lag_final_rf_res)
+
+# Predict on test data
+lag_final_rf_test_preds <- predict(lag_final_rf_fit, new_data = test_data)
+
+# Combine predictions with actuals
+lag_final_rf_test_results <- test_data %>%
+  select(new_cases) %>%
+  mutate(predicted = lag_final_rf_test_preds$.pred)
+
+# Compute test metrics
+lag_final_rf_test_metrics <- lag_final_rf_test_results %>%
+  metrics(truth = new_cases, estimate = predicted)
+
+# View test set performance
+print(lag_final_rf_test_metrics)
+
 ```
@@
 ```{r}
-forest_spec <- rand_forest(
-  mode = "regression",
-  mtry = best_params$mtry,
-  trees = best_params$trees,
-  min_n = best_params$min_n
+# plot baseline rf model prediction vs actual
+plot_base_rf <- base_final_rf_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "darkgreen") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "Base RF",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot mobility rf model prediction vs actual
+plot_rf <- final_rf_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "steelblue") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "RF with Mobility",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot lagged rf model prediction vs actual
+plot_lag_rf <- lag_final_rf_test_results %>%
+  ggplot(aes(x = new_cases, y = predicted)) +
+  geom_point(alpha = 0.6, color = "purple") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "RF with Lagged Mobility",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+rf_graphs <- plot_base_rf + plot_rf + plot_lag_rf +
+  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
+  theme(
+    plot.margin = margin(10, 10, 10, 10),
+    plot.title = element_text(size = 14, hjust = 0.5),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+
+print(rf_graphs)
+
+figure_file = here("results", "figures", "rf_graphs.png")
+ggsave(
+  filename = figure_file,
+  plot = rf_graphs,
+  width = 16, height = 5, dpi = 300
+)
+```
+
+Third, XGBoost models are tuned, trained, and evaluated. Best models are chosen through R-squared.
+
+This code chunk fits and tunes an baseline XGBoost model with no mobility predictors.
+```{r}
+# Define the XGBoost model specification (baseline version)
+boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = tune(),
+  learn_rate = tune(),
+  loss_reduction = tune()
 ) %>%
-  set_engine("ranger", seed = 1234)
+  set_engine("xgboost") %>%
+  set_mode("regression")
 
-forest_fit_final <- forest_wf %>%
-  add_model(forest_spec) %>%
-  fit(data = data)
+# Build workflow with baseline recipe and model
+base_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(boost_spec)
 
-# Get final predictions
-forest_preds_final <- predict(forest_fit_final, new_data = data)
-forest_preds_final <- tibble(truth = data$new_cases, predicted = forest_preds_final$.pred)
+# Tune XGBoost model using time-blocked CV (folds)
+base_boost_tune_res <- tune_grid(
+  base_boost_wf,
+  resamples = rolling_folds,
+  grid = 30,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
 
-# Calculate performance metrics
-forest_rmse <- forest_preds_final %>%
+# View tuning results and print best model metrics
+base_boost_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+base_best_params <- select_best(base_boost_tune_res, metric = "rsq")
+print(base_best_params)
+
+# Final XGBoost model with best hyperparameters
+base_final_boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = base_best_params$tree_depth,
+  learn_rate = base_best_params$learn_rate,
+  loss_reduction = base_best_params$loss_reduction
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+# Final model workflow
+base_final_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(base_final_boost_spec)
+
+# Fit the final model
+base_final_boost_fit <- base_final_boost_wf %>%
+  fit(data = train_data)
+
+# Predict on test data and compute performance metrics
+base_final_boost_preds <- predict(base_final_boost_fit, new_data = test_data)
+
+base_final_boost_results <- tibble(
+  truth = test_data$new_cases,
+  predicted = base_final_boost_preds$.pred
+)
+
+base_final_boost_metrics <- base_final_boost_results %>%
   metrics(truth = truth, estimate = predicted)
 
-print(forest_rmse)
+# Print final performance metrics
+print(base_final_boost_metrics)
 ```
+
+This code chunk fits and tunes an XGBoost model with mobility predictors.
+```{r}
+# Build workflow with recipe and model
+boost_wf <- workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(boost_spec)
+
+# Tune XGBoost model using time-blocked CV (folds)
+boost_tune_res <- tune_grid(
+  boost_wf,
+  resamples = rolling_folds,
+  grid = 30,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# View tuning results and print best model metrics
+boost_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+best_params <- select_best(boost_tune_res, metric = "rsq")
+print(best_params)
+
+# Final XGBoost model with best hyperparameters
+final_boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = best_params$tree_depth,
+  learn_rate = best_params$learn_rate,
+  loss_reduction = best_params$loss_reduction
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+# Final model workflow
+final_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(final_boost_spec)
+
+# Fit the final model
+final_boost_fit <- final_boost_wf %>%
+  fit(data = train_data)
+
+# Predict on test data and compute performance metrics
+final_boost_preds <- predict(final_boost_fit, new_data = test_data)
+
+final_boost_results <- tibble(
+  truth = test_data$new_cases,
+  predicted = final_boost_preds$.pred
+)
+
+final_boost_metrics <- final_boost_results %>%
+  metrics(truth = truth, estimate = predicted)
+
+# Print final performance metrics
+print(final_boost_metrics)
+
+```
+
+This code chunk fits and tunes an XGBoost model with lagged mobility predictors.
+```{r}
+# Build workflow with baseline recipe and model
+lag_boost_wf <- workflow() %>%
+  add_recipe(recipe_lag) %>%
+  add_model(boost_spec)
+
+# Tune XGBoost model using time-blocked CV (folds)
+lag_boost_tune_res <- tune_grid(
+  lag_boost_wf,
+  resamples = rolling_folds,
+  grid = 30,
+  metrics = metric_set(rmse, rsq, mae),
+  control = control_grid(parallel_over = "everything")
+)
+
+# View tuning results and print best model metrics
+lag_boost_tune_res %>%
+  collect_metrics() %>%
+  print()
+
+lag_best_params <- select_best(lag_boost_tune_res, metric = "rsq")
+print(lag_best_params)
+
+# Final XGBoost model with best hyperparameters
+lag_final_boost_spec <- boost_tree(
+  trees = 500,
+  tree_depth = lag_best_params$tree_depth,
+  learn_rate = lag_best_params$learn_rate,
+  loss_reduction = lag_best_params$loss_reduction
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+# Final model workflow
+lag_final_boost_wf <- workflow() %>%
+  add_recipe(recipe_baseline) %>%
+  add_model(lag_final_boost_spec)
+
+# Fit the final model
+lag_final_boost_fit <- lag_final_boost_wf %>%
+  fit(data = train_data)
+
+# Predict on test data and compute performance metrics
+lag_final_boost_preds <- predict(lag_final_boost_fit, new_data = test_data)
+
+lag_final_boost_results <- tibble(
+  truth = test_data$new_cases,
+  predicted = lag_final_boost_preds$.pred
+)
+
+lag_final_boost_metrics <- lag_final_boost_results %>%
+  metrics(truth = truth, estimate = predicted)
+
+# Print final performance metrics
+print(lag_final_boost_metrics)
+```
+
+```{r}
+# plot baseline boost model prediction vs. actual
+plot_base_boost <- base_final_boost_results %>%
+  ggplot(aes(x = truth, y = predicted)) +
+  geom_point(alpha = 0.6, color = "darkgreen") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "Base Boost",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot mobility boost model prediction vs actual
+plot_boost <- final_boost_results %>%
+  ggplot(aes(x = truth, y = predicted)) +
+  geom_point(alpha = 0.6, color = "steelblue") +
+  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  labs(
+    title = "Boost with Mobility",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+# plot lagged boost model prediction vs actual
+plot_lag_boost <- lag_final_boost_results %>%
+  ggplot(aes(x = truth, y = predicted)) +
+  geom_point(alpha = 0.6, color = "purple") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(
+    title = "Boost with Lagged Mobility",
+    x = "Actual Cases", y = "Predicted Cases"
+  ) +
+  theme_bw()
+
+boost_graphs <- plot_base_boost + plot_boost + plot_lag_boost +
+  plot_layout(ncol = 3, widths = c(1.5, 1.5, 1.5)) &
+  theme(
+    plot.margin = margin(10, 10, 10, 10),
+    plot.title = element_text(size = 14, hjust = 0.5),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+
+print(boost_graphs)
+
+figure_file = here("results", "figures", "boost_graphs.png")
+ggsave(
+  filename = figure_file,
+  plot = boost_graphs,
+  width = 16, height = 5, dpi = 300
+)
+
+```
+
+This code chunk merges the results of the best models into a data frame.
+```{r}
+base_lasso_metrics <- base_final_lasso_test_metrics %>% 
+  mutate(model = "LASSO (Baseline)") %>%
+  rename(mean = .estimate)
+
+lasso_metrics <- final_lasso_test_metrics %>% 
+  mutate(model = "LASSO") %>%
+  rename(mean = .estimate)
+
+lag_lasso_metrics <- lag_final_lasso_test_metrics %>% 
+  mutate(model = "LASSO (Lagged Mobility)") %>%
+  rename(mean = .estimate)
+
+final_boost_metrics <- final_boost_metrics %>%
+  mutate(model = "XGBoost") %>%
+  rename(mean = .estimate)
+
+final_forest_metrics <- final_rf_test_metrics %>%
+  mutate(model = "Random Forest") %>%
+  rename(mean = .estimate)
+
+base_final_boost_metrics <- base_final_boost_metrics %>%
+  mutate(model = "XGBoost (Baseline)") %>%
+  rename(mean = .estimate)
+
+base_final_forest_metrics <- base_final_rf_test_metrics %>%
+  mutate(model = "Random Forest (Baseline)") %>%
+  rename(mean = .estimate)
+
+lag_boost_metrics <- lag_final_boost_metrics %>%
+  mutate(model = "XGBoost (Lagged Mobility)") %>%
+  rename(mean = .estimate)
+
+lag_forest_metrics <- lag_final_rf_test_metrics %>%
+  mutate(model = "Random Forest (Lagged Mobility)") %>%
+  rename(mean = .estimate)
+
+model_metrics <- bind_rows(base_lasso_metrics, lasso_metrics, lag_lasso_metrics, base_final_boost_metrics, final_boost_metrics, lag_boost_metrics, base_final_forest_metrics, final_forest_metrics, lag_forest_metrics)
+
+model_metrics <- model_metrics %>%
+  select(-.estimator) %>%
+  pivot_wider(names_from = '.metric', values_from = mean)
+```
+
+This code chunk creates a short summary table of the results from the model_metrics data frame.
+```{r}
+library(gt)
+
+model_table <- model_metrics %>%
+  gt() %>%
+  tab_header(
+    title = "Model Performance Metrics",
+    subtitle = "Best values are highlighted per metric"
+  ) %>%
+  fmt_number(columns = c(rmse, mae), decimals = 1) %>%
+  fmt_number(columns = rsq, decimals = 3) %>%
+  cols_label(
+    model = "Model",
+    rmse = "RMSE",
+    rsq = "R²",
+    mae = "MAE"
+  ) %>%
+
+  # Highlight lowest RMSE
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rmse,
+      rows = rmse == min(rmse, na.rm = TRUE)
+    )
+  ) %>%
+
+  # Highlight lowest MAE
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = mae,
+      rows = mae == min(mae, na.rm = TRUE)
+    )
+  ) %>%
+
+  # Highlight highest R²
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rsq,
+      rows = rsq == max(rsq, na.rm = TRUE)
+    )
+  ) %>%
+
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_options(
+    table.width = pct(100),
+    table.align = "left",
+    column_labels.padding = px(10),
+    data_row.padding = px(6)
+  )
+
+library(gtExtras)
+library(webshot2)
+
+print(model_table)
+
+gtsave(model_table, here("results", "figures", "models.png"))
+```

### 2025-04-21 products/manuscript/Manuscript.qmd
Revert "Work"
This reverts commit 63ab86524d9b8d1ab4357a823b03d789a625d851.

@@
     highlight-style: github
-bibliography: ../../assets/MADA.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+bibliography: ../../assets/references/MADA.bib
+csl: ../../assets/apa.csl
 ---
@@
 
+```{r}
+library(here)
+```
+
 # Summary/Abstract
 
-*Write a summary of your project.*
+**Background**: Human mobility patterns shifted significantly throughout the COVID-19 pandemic due to public health policies and individual risk perception.
+These changes, recorded in Google’s Community Mobility Report, may hold predictive value in understanding and forecasting transmission dynamics.
 
-{{< Examining the relationship between human mobility and infectious disease spread is essential for forecasting disease incidence and informing public health interventions.
-This analysis investigates community mobility patterns and their association with daily new COVID-19 case incidence in Georgia.
-Using county-level case counts and mobility trends from Google’s Community Mobility reports, machine learning models were constructed and evaluated; models included linear regression, Random Forest, and XGBoost.
-To assess the predictive power of mobility change metrics, two feature sets were utilized.
-The first set, referred to as “baseline”, included features known to affect disease incidence, such as lagged cases and population density.
-The second set included both the previous set’s features and metrics measuring changes in mobility (recorded as percent change from baseline).
-Models were trained and tuned using time-series blocked cross-validation and performance was evaluated R-squared, RMSE, and MAE metrics.
-Results indicated that models with mobility features could achieve modest improvements in RMSE and MAE, however, did not substantially increase R-squared, suggesting limited explanatory power.
-These findings suggest that while mobility data may slightly enhance short-term predictive accuracy, their overall predictive power in explaining COVID-19 incidence may be limited in this context.
->}}
+**Objective:** This analysis evaluated the additive predictive capabilities mobility dynamics may have on modeling daily COVID-19 case incidence at the county level in Georgia.
 
-# 2 Introduction
+**Methods:** Daily case data from John Hopkins University and mobility predictors from Google were used to model case incidence across 44 Georgia counties between March 14th 2020 and March 14th 2022.
+Three modeling approaches, LASSO regression, Random Forest, and XGBoost, were applied to assess the association.
+A model was trained on three sets of feature sets, a baseline set (lagged case counts and population density), a full set (adding on mobility variables), and a 7-day lagged version of the full set.
+Models were evaluated using RMSE, R-squared, and MAE metrics on a 30 day test set, with rolling window cross-validation used during training.
 
-## 2.1 COVID-19 Pandemic
+**Results:** The XGBoost model, with the lagged feature set, achieved the best overall performance (RMSE: 37.9; R²: 0.836; MAE: 23.5).
+This model’s improvement over other feature sets indicates significant additive predictive power of lagged mobility dynamics when forecasting case incidence.
+
+**Conclusion:** This analysis suggest that incorporation of time-lagged mobility data can improve the predictive performance of machine learning models for COVID-19 case incidence.
+The results highlight the potential in assessing behavioral data in forecasting infectious disease dynamics and support the integration of mobility metrics in public health surveillance.
+
+# Introduction
+
+## COVID-19 Pandemic
 
 The emergence of the novel corona-virus SARS-CoV-2 caused one of the most significant global health crises in modern history.
-First identified in Wuhan, China, in late 2019, the virus’s transmissibility and globalism led to its rapid spread worldwide, to leading the World Health Organization to declare it a pandemic on March 11, 2020.
-Just two days later, on March 13th, 2020, the Trump Administration declared a nationwide emergency due to the COVID-19 pandemic, marking the biggest virus outbreak since \_\_\_.
-In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lockdown procedures (CDC 2024)[@centersfordiseasecontrolandprevention_2024_covid19].
+First identified in late 2019, the virus rapidly spread across the world due to its high transmissiblity and global interconnectedness, leading the World Health Organization to declare it a pandemic on March 11, 2020.
+Two days later, on March 13th, 2020, the United States declared a nationwide emergency due to the COVID-19 pandemic, marking the biggest virus outbreak since the 1916 influenza pandemic.
+In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lock-down procedures [@centersfordiseasecontrolandpreventionCOVID19Timeline2024].
 
-## 2.2 Mobility Dynamics
+## Mobility Dynamics
 
-Human mobility, a key driver[of respiratory disease transmi](https://www.nature.com/articles/s41598-024-64230-1)ssion, shifted significantly during the pandemic in response to public health policies and disease risk perception.
-Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lockdowns.
-Lockdowns involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
-Looking at the Wikipedia page shows a lack of prior history implementing lockdowns.
+Human mobility, a key driver of respiratory disease transmission, shifted significantly during the pandemic in response to public health policies and disease risk perception [@paltraEffectMobilityReductions2024].
+Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lock-downs.
+Lock-downs involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
+Looking at the Wikipedia page shows a lack of prior history implementing lock-downs.
 Beyond these measures which lasted only a few weeks in Georgia, individual risk perception played into the compliance of other preventative behaviors (masking, social distancing, etc.).
-The[variability of individual risk perception](https://pmc.ncbi.nlm.nih.gov/articles/PMC9028425/#:~:text=The%20results%20indicate%20that%20a,%2C%20geographical%20factors%2C%20and%20timing.)has led to complex mobility patterns during the pandemic, for example, surges in mobility amidst large case outbreaks.
+The variability of risk individual perception has led to complex mobility patterns during the pandemic, for example, surges in mobility amidst large case outbreaks [@cipollettaRiskPerceptionCOVID192022].
 In order to analyze and quantify this relationship, machine learning models were adapted.
 
-\[insert stats about pandemic\]
+## Machine Learning
 
-## 2.3 Machine Learning
-
-[Machine Learning](https://pmc.ncbi.nlm.nih.gov/articles/PMC9949554/) refers to a class of data-driven algorithms that aim to analyze associations and relationships found in data.
+Machine learning refers to a class of data-driven algorithms that aim to analyze associations and relationships found in data.
 These techniques can be supervised or unsupervised; in supervised ML, models are given labeled inputs/features to derive and predict an output.
 These models aim to also approximate the relationship between outputs and inputs, quantify predictions, or approximate classification tasks.
-Beyond regular statistical analysis, machine learning methods offer powerful tools for forecasting infectious disease trends by identifying complex, nonlinear relationships between predictors and outcomes.
-S[ome common m](https://pmc.ncbi.nlm.nih.gov/articles/PMC8219638/)odels in disease forecasting include ARMA, ARIMA, LASSO, XGBoost, and various neural network techniques.
-Several autho[rs](https://pmc.ncbi.nlm.nih.gov/articles/PMC8731233/)have utilized case counts, estimations, demographics, and more to forecast disease trends.
+Beyond regular statistical analysis, machine learning methods offer powerful tools for forecasting infectious disease trends by identifying complex, nonlinear relationships between predictors and outcomes [@rashidiCommonStatisticalConcepts2023].
+Some common models in disease forecasting include ARMA, ARIMA, LASSO, XGBoost, and various neural network techniques [@alfredRolesMachineLearning2021].
+Several authors have utilized case counts, estimations, demographics, and more to forecast disease trends [@ogunjoPredictingCOVID19Cases2022].
 However, the predictive value of real-time mobility data, particularly in the context of an evolving pandemic, has not been assessed through modeling.
@@
 
-# 3 Data
+# Data
 
-## 3.1 Google’s Community Mobility Report
+## Google’s Community Mobility Report
 
-During the pandemic, Google began to collect aggregated, anonymized data from users utilizing Google products (apps, phones, etc.) to track changes in mobility.
+During the pandemic, Google began to collect aggregated, anonymized data from users utilizing Google products (apps, phones, etc.) to track changes in mobility [@googleCOVID19CommunityMobility2020].
 Within the data, Google measures mobility as a percent change difference from baseline measurements; for example, a -45% change in retail and recreation mobility indicates a 45% reduction in movement to those categorical locations.
@@
 
-## 3.2 John Hopkins University COVID-19 Case Data
+## John Hopkins University COVID-19 Case Data
 
 The COVID-19 Data Repository by the Center for Systems Science and Engineering at Johns Hopkins University is a comprehensive data set that tracks global COVID-19 cases, recoveries, and deaths.
-This data was recorded daily for several years across every county in the US.
+This data was recorded daily for several years across every county in the US [@cssegisanddataCOVID19DataRepository2022].
 
-## 3.3 Final Data-set
+## Final Data-set
 
@@
 Many of Georgia’s counties are rural and as such, lacked many mobility metrics.
-The data was filtered from March 14t 2020 to March 14 2022.
-This was selected specifically as March 14th marks the date of Governor Brian Kemp’s announcement of Georgia's Public Health State of Emergency.
+The data was filtered from March 14th 2020 to March 14th 2022.
+This was selected specifically as March 14th marks the date of Governor Brian Kemp’s announcement of Georgia's Public Health State of Emergency [@hartCOVID19PandemicGeorgia2025].
 
@@
 
-# 4 Methods
+# Methods
 
-## 4.1 Model Selection
+## Model Selection
 
-As informed by Alfred and Obit 2021, machine learning models have wide application in forecasting outbreaks and disease incidence.
+As informed by @alfredRolesMachineLearning2021, machine learning models have wide application in forecasting outbreaks and disease incidence.
 In their review, they outline several common applications of regression and classification models.
@@
 
-### Linear Regression
+**LASSO Regression**
 
-A statistical modeling method used to quantify the relationship between a continuous outcome variable and one or more predictors.
-In its simplest form, linear regression estimates, or fits, a linear equation explaining the relationship by minimizing the sum of squared differences between residuals.
-Linear regression assumes
+Also known as Least Absolute Shrinkage and Selection Operator, Lasso is a regularization technique used in regression modeling to prevent overfitting and improve model interpretability.
+Lasso adds a penalty term equal to the absolute value of the magnitude of coefficients to the loss function [@ibmLassoRegression2024].
+This penalty can shrink coefficients to zero, effectively performing variable selection.
+Lasso was applied to linear regression models in the analysis.
 
-### Random Forest
-
-(https://www.stat.berkeley.edu/\~breiman/randomforest2001.pdf)\
+**Random Forest**
 
@@
 A decision tree is a modeling technique that partitions  data into subsets based on the input features; it aims to minimize variance of the outcome variable in each subgroup.
-These splits form a tree-like structure whe -.
-Decision trees are powerful and can capture nonlinear complex relationships, however, can overfit training data.  Random forests are an average of these trees to produce a final prediction and can produce more stable/accurate models as a result.
+These splits form a tree-like structure where each internal node represents a decision based on a feature, each branch corresponds to an outcome of the decision, and each leaf node represents a predicted value.
+Decision trees are powerful and can capture nonlinear complex relationships, however, can overfit training data.  Random forests are an average of these trees to produce a final prediction and can produce more stable/accurate models as a result [@breimanRandomForests2001].
 
-### XGBoost
+**XGBoost**
 
-(https://arxiv.org/abs/1603.02754): Also known as Extreme Gradient Boosting, XGBoost is a machine learning algorithm based on decision tress, however, instead of averaging trees, it tries to correct errors made by previous ones.
+Also known as Extreme Gradient Boosting, XGBoost is a machine learning algorithm based on decision tress, however, instead of averaging trees, it tries to correct errors made by previous ones.
 This technique is known as boosting.
-In simple terms, XGBoost makes predictions with a single tree and looks at the errors and creates a second tree focused on minimizing the errors.
+In simple terms, XGBoost makes predictions with a single tree and looks at the errors and creates a second tree focused on minimizing the errors [@chenXGBoostScalableTree2016].
 
-Two sets of predictors were chosen to assess the effect mobility changes have on case incidence.
+**Feature Set**
+
+Three sets of predictors were chosen to assess the effect mobility changes have on case incidence.
 The first set, the baseline predictors, included population density and 3 spaced out lagged variables for case counts (1 day, 7 days, and 14 days).
-These were chosen because of  their [known influence](https://pmc.ncbi.nlm.nih.gov/articles/PMC8983058/#:~:text=This%20study%20has%20shown%20that,data%20for%20COVID%E2%80%9019%20cases.) on disease transmission.
-The second set, the mobility predictors, included population density, the 3 spaced out lagged variables, and mobility indicators from Google’s report (e.g., workplace, grocery, retail, transit).
+These were chosen because of  their known influence on disease transmission [@ogunjoPredictingCOVID19Cases2022].
+The second set, the mobility predictors, included population density, the 3 spaced out lagged case counts, and mobility indicators from Google’s report (e.g., workplace, grocery, retail, transit).
+Lastly, the third set included 7-day lagged versions of the mobiltiy predictors, population density, and the three spaced out lagged case counts.
 
-## 4.2 Cross Validation
+## Cross Validation
 
-To evaluate model performance while accounting for the temporal structure of the data, a time-blocked cross-validation strategy was employed.
-The dataset was partitioned into eight non-overlapping temporal folds, each representing a three-month interval spanning March 2020 to March 2022.
-Each observation (a day for a corresponding county) was assigned a fold based on its calendar month.
-Grouped k-fold cross-validation using these time-based identifiers were used to ensure data from the future could not leak into the training set.
+To evaluate model performance while accounting for the temporal structure of the data, a rolling origin cross-validation technique with non-cumulative, time-blocked splits was utilized.
+The last 30 days of the data set were set aside and used for a test set for final model evaluation.
+Model training and cross-validation were conducted utilizing the remaining data set.
 
-## 4.3 Model Evaluation and Metrics
+Cross-validation using a rolling window was implemented.
+Each fold contained a training window of four months (120 days per county) followed by a validation window of approximately one month (30 days).
+The window was advanced forward in 15-day increments which creates multiple sequential train-validation splits.
+Non-cumulative windows were enforced to ensure that training sets did not grow over time.
+This design was intended to reflect a realistic forecasting scenario in which models are periodically re-trained using a fixed window of recent data to predict outcomes in a short future period.
+Each validation period occurs strictly after its corresponding training window to prevent information leakage.\
+\
+After tuning and validation, the final model was re-trained on the full training set and evaluated on the 30 day test set to assess its predictive performance.
 
-### R-squared
+## Model Evaluation and Metrics
+
+**R-squared**
 
@@
 A higher R-squared value generally is indicative of a better-fitting model, although could be misleading when over fitting occurs.
+Criteria for final model selection focused on achieving the highest r-squared value.
 
-### RMSE
+**RMSE**
 
@@
 
-### MAE
+**MAE**
 
@@
 
-After the initial RF and XGBoost models were made, their hyperparameters were tuned to improve predictive performance.
+After the initial RF and XGBoost models were made, their hyper-parameters were tuned to improve predictive performance.
 
-## 4.4 Software
+## Software
 
 The analysis was conducted on RStudio 4.3.2 on Windows 11.
-The following R packages were used: [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html), [broom](https://cran.r-project.org/web/packages/broom/index.html), [here](https://cran.r-project.org/web/packages/here/index.html), [glmnet](https://cran.r-project.org/web/packages/glmnet/index.html), [MASS](https://cran.r-project.org/web/packages/MASS/index.html), [tidymodels](https://cran.r-project.org/web/packages/tidymodels/index.html), [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html), [rsample](https://cran.r-project.org/web/packages/rsample/index.html), [parsnip](https://cran.r-project.org/web/packages/parsnip/index.html), [future](https://cran.r-project.org/web/packages/future/index.html), [vip](https://cran.r-project.org/web/packages/vip/index.html), [zoo](https://cran.r-project.org/web/packages/zoo/index.html), [patchwork](https://cran.r-project.org/web/packages/patchwork/index.html), [e1071](https://cran.r-project.org/web/packages/e1071/index.html), [scales](https://cran.r-project.org/web/packages/scales/index.html), [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/index.html), [corrplot](https://cran.r-project.org/web/packages/corrplot/index.html), [reshape2](https://cran.r-project.org/web/packages/reshape2/index.html), [tidyr](https://cran.r-project.org/web/packages/tidyr/index.html), [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html), [readxl](https://cran.r-project.org/web/packages/readxl/index.html), [stringr](https://cran.r-project.org/web/packages/stringr/index.html), [skimr](https://cran.r-project.org/web/packages/skimr/index.html), [tigris](https://cran.r-project.org/web/packages/tigris/index.html), and [sf](https://cran.r-project.org/web/packages/sf/index.html).
+The following R packages were used: ggplot2 [@wickhamGgplot2CreateElegant2019], broom [@robinsonBroomConvertStatistical2022], here [@mullerHereSimplerWay2020], glmnet [@friedmanGlmnetLassoElasticNet2021], MASS [@ripleyMASSSupportFunctions2022], tidymodels [@kuhnTidymodelsEasilyInstall2023], dplyr [@wickhamDplyrGrammarData2020], rsample [@frickRsampleGeneralResampling2025], parsnip [@kuhnParsnipCommonAPI2025], future [@bengtssonFutureUnifiedParallel2025], vip [@greenwellVipVariableImportance2023], zoo [@zeileisZooS3Infrastructure2021], patchwork [@pedersenPatchworkComposerPlots2020], e1071 [@meyerE1071MiscFunctions2022], scales [@wickhamScalesScaleFunctions2022], RColorBrewer [@neuwirthRColorBrewerColorBrewerPalettes2022], corrplot [@weiCorrplotVisualizationCorrelation2017], reshape2 [@wickhamReshape2FlexiblyReshape2020], tidyr [@wickhamTidyrTidyMessy2020], lubridate [@spinuLubridateMakeDealing2022], readxl [@wickhamReadxlReadExcel2019], stringr [@wickhamStringrSimpleConsistent2019], skimr [@waringSkimrCompactFlexible2022], tigris [@walkerTigrisLoadCensus2025], and sf [@pebesmaSfSimpleFeatures2021]
 
@@
 
+To provide context for the modeling analysis, visualizations of Georgia's COVID-19 case incidence and mobility dynamics are presented.
+
 ```{r}
 #| label: cases_graph
-#| fig-cap: "Total COVID-19 Cases in Georgia over time"
+#| fig-cap: "Figure 5.1: Total COVID-19 Cases in Georgia over time"
 #| echo: FALSE
@@
 #| label: incidenceplot
-#| fig-cap: "Daily New COVID-19 Cases overtime in the GA (2020-2022)"
+#| fig-cap: "Figure 5.2: Daily New COVID-19 Cases overtime in the GA (2020-2022)"
 #| echo: FALSE
@@
 
+Figure 5.1 and 5.2 demonstrate distinct outbreaks of infection, with sharp peaks during the winter months and troughs in the summer, suggesting seasonal patterns and periods of increased transmission.
+
 ```{r}
 #| label: mobiltiygraph
-#| fig-cap: "Percent Changes in Mobility per week in GA"
+#| fig-cap: "Figure 5.3: Percent Changes in Mobility per week in GA"
 #| echo: FALSE
@@
 #| label: mobilitygrid
-#| fig-cap: "Percent Changes in Mobility per week in GA"
+#| fig-cap: "Figure 5.4: Percent Changes in Mobility per week in GA"
 #| echo: FALSE
@@
 
+Figure 5.3 and 5.4 demonstrate distinct changes in mobility dynamics throughout the pandemic, with sharp drops corresponding with outbreaks and the seasons.
+Workplace mobility tended to remain below the baseline level while residential mobility increased during this period.
+There were frequent fluctations in retail and grocery mobility, possibly a result of outbreaks, changing policies, or even holiday-related activity.
+Levels of mobility sharply declined and would remain below baseline until a year later.
+
 ```{r}
-#| label: mobilitymap
-#| fig-cap: "Average of mobility changes on 10/15/2022 mapped to the U.S."
+#| label: density
+#| fig-cap: "Figure 5.5: Density Plot of Daily New Cases"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","mobilitymap.png"))
+knitr::include_graphics(here("results","figures","density_cases.png"))
 ```
+
+```{r}
+#| label: density500
+#| fig-cap: "Figure 5.6: Density Plot of Daily New Cases (Over 500)"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","density_cases_500.png"))
+```
+
+Figure 5.5 and 5.6 shows the distribution of daily new COVID-19 cases in Georgia.
+The distribution is heavily right skewed, with most days clustered around lower incidence leels and fewer days where extremely high case transmission occurs.
+This skew suggests surges were relatively infrequent ocmapred to more moderate levels of daily incidence.
+This kurtosis can affect the performance of linear regrssion models and ARIMA where normality is an assumption whereas Random Forest and XGBoost are more robust to this skewing.
+
+```{r}
+#| label: popdensity
+#| fig-cap: "Figure 5.7: Distribution of Population Density among Counties included in Data"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","histo_pop_density.png"))
+```
+
+Population density across the included Georgia counties is also heavily right-skewed, with the majority of counties having lower densities.
+This imbalance may influence transmission dynamics and was incldued for modeling.
 
@@
 ```
+
 In addition to visual analysis, correlation analysis was done.
@@
 Analysis reveals that lagged case counts were strongly correlated with daily incidence while in contrast mobility variables demonstrated weak relationships.
-This is suggestive of lagged variables's strong predictive power and mobility metric’s weak power.
 
-## Full analysis
+## Model Analysis
 
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
+```{r}
+#| label: lassographs
+#| fig-cap: "Plot of Actual vs Predicted Cases in Lasso Regression Models"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","lasso_graphs.png"))
+```
+
+```{r}
+#| label: rfgraphs
+#| fig-cap: "Plot of Actual vs Predicted Cases in Random Forest Models"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","rf_graphs.png"))
+```
+
+```{r}
+#| label: boostgraphs
+#| fig-cap: "Plot of Actual vs Predicted Cases in XGBoost Models"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","boost_graphs.png"))
+```
+
+To assess the association between mobility dynamics and COVID-19 case incidence in Georgia, three modeling approaches were utilized, LASSO regression, Random Forest, and XGBoost.
+In addition to the three modeling approaches, three feature sets were utilized, a baseline recipe lacking mobility predictors, a full recipe containing mobility predictors, and a 7-day lagged version of the full recipe.
+Model performance was assessed using RMSE, R-squared, and MAE on a 30-day test set.\
+
+```{r}
+#| label: modeltable
+#| fig-cap: "Table 5.1: Model Performance"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","models.png"))
+```
 
@@
 
-*Summarize what you did, what you found and what it means.*
+This study aimed to evaluate the predictive value of mobility dynamis on COVID-19 case incidence at the county level in Georgia using three different machine learning methods, LASSO, Random Forest, and XGBoost.
+
+Among all models, XGBoost with lagged mobility predictors achieved the best overall performance, with an RMSE of 37.9, an R-squared of 0.836, and MAE of 23.5.
+This model outperformed all models in RMSE and had great relative performance in R-squared and MAE, suggesting the improvement of predictive ability through incorporating time-lagged mobility trends.
+The improvement in RMSE and MAE in relation to the other models indicates a meaningful reduction in both large and average prediction errors when lagged mobility data is included.\
+\
+The LASSO models also demonstrate strong predictive ability, with the mobility driven models having overall greater performance.
+As demonstrated in XGBoost, the lagged models have great performance, especially when compared to the baseline models.
+
+In contrast, the Random Forest models generally have the poorest performance in terms of RMSE, R-squared, and MAE.
+However, it should be noted that the lowest MAE in all models occurs in the Random Forest with lagged mobility variables.\
+\
+While mostly exploratory and preliminary, the results suggest that the inclusion of lagged mobility predictors can significantly improve model performance, specifically in XGboost.
+The lagged XGBoost model achieved great relative performance in RMSE and MAE while maintaining a high R-squared.
+This aligns with previous epidemiological findings suggesting that population movement precedes changes in infectious rates, often with a delay due to disease incubation and reporting lag.
+These findings highlight the importance of temporal feature engineering when using behavioral data such as mobility to predict case outcomes.
 
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
+This analysis has several limitations.
+First, the mobility data was incomplete, leading to the filtering of majority of Georgia counties out of the data set.
+This could be due to the methodology used for measuring which would skew the data towards more suburban and urban areas where phone usage is more common.
+This can also mask behaviors related to communities on a more individual scale.
+Additionally, this study did not incorporate many variables relating to policy changes or other events that could precede and explain changes in COVID-19 incidence.
+Lastly, machine learning models can be highly sensitive to parameter tuning and data pre-processing.
+
+A strength of this analysis is the inclusion of open source data, along with basic feature engineering, which allows for reproduction and cost-efficiency.
 
@@
 
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+This research establishes and explores the potential of incorporating mobility dynamics into public health surveillance, specifically, for COVID-19 case forecasting.
+Future research should explore the association further and consider the use of more individual behaviors, like adherence to masking or social distancing policies, to improve forecasting accuracy.
 

### 2025-04-21 code/analysis-code/statisitical-analysis.qmd
Meta data

@@
 ```
+
 ```{r}

### 2025-04-21 code/processing-code/processing-file.qmd
Meta data

---
title: "Cleaning Script"
author: "Vincent Nguyen"
date: "2025-02-21"
output: html_document
---

# Setup

Load needed packages. Make sure they are installed.

```{r}
# load needed packages
library(here) 
library(dplyr)
library(tidyr)
library(lubridate)
library(readxl)
library(stringr)
library(skimr)

library(tigris)
library(sf)
```

# Data loading

```{r}
#Google data is in three parts; load and join
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv")) %>%
  filter(sub_region_1 == "Georgia")
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))%>%
  filter(sub_region_1 == "Georgia")
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))%>%
  filter(sub_region_1 == "Georgia")

# Combine columns and remove irrelevant rows
combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)

# jhk covid time series cases
time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv")) %>%
  filter(Province_State == "Georgia")

# county population information
ch_pop_data <- read_excel(here::here("data", "raw-data", "co-est2023-pop.xlsx"), skip = 3)

```

# Cleaning and Wrangling

For this cleaning file, the workflow focused on preparing the data frames for joining and cleaning out columns as joins and mutations occurred.\

## County Population Information

The county population count data was first. Unnecessary rows were identified through the Rstudio view() feature and removed. Also, the data was converted into a long format for easy joining and analysis later.

```{r}
# Remove the second column
ch_pop_data <- ch_pop_data[, -2]

# Convert to a long format
ch_pop_data <- ch_pop_data %>%
  pivot_longer(
    cols = c(`2020`,`2021`, `2022`, `2023`),    
    names_to = "year",        
    values_to = "population_count"              
  )

# Fix the first column
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_remove(`...1`, "^\\."),            
    county = str_remove(county, "County"),     
    county = str_to_lower(county),                  
    state = str_extract(county, "(?<=,).*"),        
    county = str_remove(county, ",.*")              
  ) %>%
  mutate(
    state = str_to_lower(str_trim(state))           
  )

# Change year column to be double for joining later
ch_pop_data$year <- as.double(ch_pop_data$year)

# Clean names to prepare for joining
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_trim(str_to_lower(county)),  
    state = str_trim(str_to_lower(state))     
  )

# Only include Georgia counties
ch_pop_data <- ch_pop_data %>%
  filter(state == "georgia")

```

## Google Community Mobility Reports

Next, upon visual inspection in preparation for joining, it became apparent that the mobility report had many missing values. To quantify this, NA values were counted across columns and skim() was used.

After noticing a lack of data in the transit and parks category, they were removed. This does remove some important info, however, the completeness of the data was priortized instead. Additionally, not all incomplete rows were removed so completeness can be assessed later.

The data frame was also cleaned and prepared for joining.

```{r}
# Get a glimpse at what data is complete
colSums(is.na(combined_mobility_data))

skim(combined_mobility_data)

# Ok it is clear that some locations are absent from GA as a whole. Based on my own knowledge and this data, I should consider removing transit stations as public transit is not common in GA. Additionally, interestingly, parks are absent. Perhaps Google defines parks differently? For now, I will remove the category of mobility. I deal with completeness later in the code.
combined_mobility_data <- combined_mobility_data %>% 
  dplyr::select(-transit_stations_percent_change_from_baseline, -parks_percent_change_from_baseline)

# Prepare mobility report data for joining
mobility_df <- combined_mobility_data %>%
  rename(county = sub_region_2, state = sub_region_1) %>%
  mutate(county = tolower(county), 
         state = tolower(state),
         county = gsub(" county$", "", county),  # Remove "county" from county name
         date = as.Date(date, format = "%Y-%m-%d"))  # Correct format for "YYYY-MM-DD"
```

## John Hopkins Case Count Time Series

Lastly, the JHU data was prepared. After inspection, it is clear that the Time Series data is wide while the Mobility Report is long. For easier analysis and joining, the Time Series data was converted into a long format using pivot_longer().

```{r}

# Pivot the time series data and change names around
cases_long <- time_series_covid %>%
  pivot_longer(
    cols = starts_with("X"),  
    names_to = "date",
    values_to = "total_cases"
  ) %>%
  mutate(date = gsub("^X", "", date),  # Remove leading x for dates to join data sets in a bit
         date = gsub("\\.", "/", date),  # Replace dots with slashes
         date = as.Date(date, format = "%m/%d/%y"))  # Use "%y" instead of "%Y"

# Standardize county & state names (convert to lowercase to avoid mismatches)
cases_long <- cases_long %>%
  rename(county = Admin2, state = Province_State) %>%
  mutate(county = tolower(county), state = tolower(state))
```

## Filtering and Joining

After preparing the data sets and visual inspection, it was clear the dates for the data sets differed. The view() feature in Rstudio was used to find the overlap.

Once filtered, all three data sets were joined, an incidence per 10k column was added, and more column dropping occurred.

```{r}
# Filter cases_long and mobility_df based on the date range
cases_long_filtered <- cases_long %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

mobility_df_filtered <- mobility_df %>%
  filter(date >= "2020-02-15" & date <= "2022-10-15")

# Merge the data sets together
merged_df <- cases_long_filtered %>%
  inner_join(mobility_df_filtered, by = c("county", "state", "date"))

# Additionally, add a column about new cases for further analysis down the line.
merged_df<- merged_df %>%
  arrange(state, county, date) %>%  
  group_by(state, county) %>%  
  mutate(new_cases = c(NA, diff(total_cases))) %>%  
  ungroup()

# Join the county population count data
merged_df<- merged_df %>%
  mutate(year = year(date)) %>%
  left_join(
    ch_pop_data %>% 
      dplyr::select(year, population_count, county, state),
    by = c("state" = "state", "county" = "county", "year" = "year")
  )

# Create column for incidence rate per 10,000 (10,000 because numbers tend to be small when per 100k)
merged_df <- merged_df %>%
  mutate(incidence_rate = (new_cases / population_count) * 10000)

# Clean up and Removal of unnecessary columns
merged_df <- merged_df %>% dplyr::select(-UID, -iso2, -iso3, -code3, -country_region_code, -country_region, -iso_3166_2_code, -place_id, -census_fips_code, -metro_area, -FIPS, -Lat, -Long_, -Combined_Key)
```

## Implementation of SF Package for Population Density Calculation

To help explain transmission patterns, population density was calculated using the shape files found in the package, sf.

```{r}
library(tigris)
library(sf)

# Load county shapefile for GA
ga_counties <- counties(state = "GA", cb = TRUE, class = "sf")

# Make county column lowercase for merging
ga_counties <- ga_counties %>%
  mutate(NAME = tolower(NAME))

# Merge with processed population data (assumed already cleaned)
ga_counties <- ga_counties %>%
  left_join(ch_pop_data, by = c("NAME" = "county"))

# Calculate county area in square kilometers
ga_counties <- ga_counties %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6)  # Convert m² to km²

# Compute population density (people per km²) and remove unnecessary data
ga_counties <- ga_counties %>%
  mutate(pop_density = population_count / area_km2) %>%
  dplyr::select(pop_density, year, NAME)

# Add in population density column and remove the geometry from the df
merged_df <- merged_df %>%
  left_join(ga_counties, by = c("county" = "NAME", "year" = "year")) %>%
  as_tibble() %>%
  dplyr::select(-geometry)

```

## Additional Cleaning and Filtering

This code chunk focuses on cleaning some of the values to not be NA and some final filtering.

```{r}
# Clean negative and NA values
merged_df$new_cases[merged_df$new_cases < 0] <- 0
merged_df$new_cases[is.na(merged_df$new_cases)] <- 0
merged_df$incidence_rate[is.na(merged_df$incidence_rate)] <- 0

# Filter for 2 years (outside of this range generally is 0)
last_case_date <- as.Date("2022-03-14")
first_case_date <- as.Date("2020-03-14")

merged_df <- merged_df %>%
  filter(date >= first_case_date) %>%
  filter(date <= last_case_date) %>%
  group_by(county)

# Convert new_cases to integer
merged_df<- merged_df %>% 
  mutate(new_cases = as.integer(new_cases))

```

## Evaluation and Filtering for Completeness of Data

As a final step, missing values in the mobility data were addressed by evaluating the completeness of each county's observations; the focus was placed on identifying days in which all mobility metrics were reported. Counties with a high proportion of incomplete records were excluded, as these gaps could mask meaningful trends and introduce bias. To mitigate this issue, counties with less than 80% data completeness were removed to support more robust and reliable analyses.

```{r}
# check the completeness by county
completeness_by_county <- merged_df %>%
  dplyr::select(retail_and_recreation_percent_change_from_baseline,  
                grocery_and_pharmacy_percent_change_from_baseline,
                workplaces_percent_change_from_baseline,
                residential_percent_change_from_baseline) %>%
  group_by(county) %>%
  rowwise() %>%
  mutate(row_complete = all(!is.na(c_across(everything())))) %>%
  ungroup() %>%
  group_by(county) %>%
  summarise(
    total_rows = n(),
    complete_rows = sum(row_complete),  # Count of rows that are complete
    completeness_rate = complete_rows / total_rows  # Proportion of complete rows
  ) %>%
  arrange(desc(completeness_rate))

# Set the completeness threshold of 80%
completeness_threshold <- 0.80

# Filter counties based on the completeness rate
counties_to_include <- completeness_by_county %>%
  filter(completeness_rate >= completeness_threshold) %>%
  pull(county)

# Filter the dataset to include only counties with at least 80% completeness
merged_df_clean <- merged_df %>%
  filter(county %in% counties_to_include)
```

The cleaning and wrangling process is complete.

To summarize, the three data sets were joined, cleaned, filtered, and mutated to create columns (population density and incidence) for analysis later.

# Save data

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(merged_df_clean, file = save_data_location)

# Saved a more complete case data set for visualizations
save_data_location <- here::here("data","processed-data","processedcasedata.rds")
saveRDS(cases_long_filtered, file = save_data_location)
```

### 2025-04-21 products/manuscript/Manuscript.qmd
Meta data

@@
 The analysis was conducted on RStudio 4.3.2 on Windows 11.
-The following R packages were used: ggplot2 [@wickhamGgplot2CreateElegant2019], broom [@robinsonBroomConvertStatistical2022], here [@mullerHereSimplerWay2020], glmnet [@friedmanGlmnetLassoElasticNet2021], MASS [@ripleyMASSSupportFunctions2022], tidymodels [@kuhnTidymodelsEasilyInstall2023], dplyr [@wickhamDplyrGrammarData2020], rsample [@frickRsampleGeneralResampling2025], parsnip [@kuhnParsnipCommonAPI2025], future [@bengtssonFutureUnifiedParallel2025], vip [@greenwellVipVariableImportance2023], zoo [@zeileisZooS3Infrastructure2021], patchwork [@pedersenPatchworkComposerPlots2020], e1071 [@meyerE1071MiscFunctions2022], scales [@wickhamScalesScaleFunctions2022], RColorBrewer [@neuwirthRColorBrewerColorBrewerPalettes2022], corrplot [@weiCorrplotVisualizationCorrelation2017], reshape2 [@wickhamReshape2FlexiblyReshape2020], tidyr [@wickhamTidyrTidyMessy2020], lubridate [@spinuLubridateMakeDealing2022], readxl [@wickhamReadxlReadExcel2019], stringr [@wickhamStringrSimpleConsistent2019], skimr [@waringSkimrCompactFlexible2022], tigris [@walkerTigrisLoadCensus2025], and sf [@pebesmaSfSimpleFeatures2021]
+The following R packages were used: ggplot2 [@wickhamGgplot2CreateElegant2019], broom [@robinsonBroomConvertStatistical2022], here [@mullerHereSimplerWay2020], glmnet [@friedmanGlmnetLassoElasticNet2021], MASS [@ripleyMASSSupportFunctions2022], tidymodels [@kuhnTidymodelsEasilyInstall2023], dplyr [@wickhamDplyrGrammarData2020], rsample [@frickRsampleGeneralResampling2025], parsnip [@kuhnParsnipCommonAPI2025], future [@bengtssonFutureUnifiedParallel2025], vip [@greenwellVipVariableImportance2023], zoo [@zeileisZooS3Infrastructure2021], patchwork [@pedersenPatchworkComposerPlots2020], e1071 [@meyerE1071MiscFunctions2022], scales [@wickhamScalesScaleFunctions2022], RColorBrewer [@neuwirthRColorBrewerColorBrewerPalettes2022], corrplot [@weiCorrplotVisualizationCorrelation2017], reshape2 [@wickhamReshape2FlexiblyReshape2020], tidyr [@wickhamTidyrTidyMessy2020], lubridate [@spinuLubridateMakeDealing2022], readxl [@wickhamReadxlReadExcel2019], stringr [@wickhamStringrSimpleConsistent2019], skimr [@waringSkimrCompactFlexible2022], gtExtras [@mockGtExtrasExtendingGt2023], webshot2[@changWebshot2TakeScreenshots2023], tigris [@walkerTigrisLoadCensus2025], and sf [@pebesmaSfSimpleFeatures2021]
 

### 2025-04-21 code/analysis-code/statisitical-analysis.qmd
Color Corrections

@@
 library(patchwork)
+library(gtExtras)
+library(webshot2)
+library(gt)
 ```
@@
 
+# Print
 print(fold_sizes)
@@
   ggplot(aes(x = new_cases, y = predicted)) +
-  geom_point(alpha = 0.6, color = "darkgreen") +
-  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  geom_point(alpha = 0.6, color = "#69b3a2") +
+  geom_abline(slope = 1, intercept = 0, color = "#DBA2A2", linetype = "dashed") +
   labs(
@@
   ggplot(aes(x = new_cases, y = predicted)) +
-  geom_point(alpha = 0.6, color = "steelblue") +
-  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  geom_point(alpha = 0.6, color = "#BFA2DB") +
+  geom_abline(slope = 1, intercept = 0, color = "#DBA2A2", linetype = "dashed") +
   labs(
@@
   ggplot(aes(x = new_cases, y = predicted)) +
-  geom_point(alpha = 0.6, color = "purple") +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  geom_point(alpha = 0.6, color = "#F2C29B") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#DBA2A2") +
   labs(
@@
 
+# Combine + formatting
 lasso_graphs <- plot_base_lasso + plot_lasso + plot_lag_lasso +
@@
 
+# Print graph for inspection
 print(lasso_graphs)
 
+# Save file
 figure_file = here("results", "figures", "lasso_graphs.png")
@@
 
+# Combine + formatting
 rf_graphs <- plot_base_rf + plot_rf + plot_lag_rf +
@@
 
+# Print for visual inspection
 print(rf_graphs)
 
+# Save file
 figure_file = here("results", "figures", "rf_graphs.png")
@@
 
+# Create tibble with truth and predcitions
 base_final_boost_results <- tibble(
@@
 
+# Calculate metrics using previous tibble
 base_final_boost_metrics <- base_final_boost_results %>%
@@
 
+# Create tibble with truth and predcitions
 final_boost_results <- tibble(
@@
 
+# Calculate metrics using previous tibble
 final_boost_metrics <- final_boost_results %>%
@@
 
+# Create tibble with truth and predcitions
 lag_final_boost_results <- tibble(
@@
 
+# Calculate metrics using previous tibble
 lag_final_boost_metrics <- lag_final_boost_results %>%
@@
   ggplot(aes(x = truth, y = predicted)) +
-  geom_point(alpha = 0.6, color = "darkgreen") +
-  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  geom_point(alpha = 0.6, color = "#69b3a2") +
+  geom_abline(slope = 1, intercept = 0, color = "#DBA2A2", linetype = "dashed") +
   labs(
@@
   ggplot(aes(x = truth, y = predicted)) +
-  geom_point(alpha = 0.6, color = "steelblue") +
-  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
+  geom_point(alpha = 0.6, color = "#BFA2DB") +
+  geom_abline(slope = 1, intercept = 0, color = "#DBA2A2", linetype = "dashed") +
   labs(
@@
   ggplot(aes(x = truth, y = predicted)) +
-  geom_point(alpha = 0.6, color = "purple") +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  geom_point(alpha = 0.6, color = "#F2C29B") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#DBA2A2") +
   labs(
@@
 
+# Combine graphs + formatting
 boost_graphs <- plot_base_boost + plot_boost + plot_lag_boost +
@@
 
+# Print for visual inspection
 print(boost_graphs)
 
+# Save file
 figure_file = here("results", "figures", "boost_graphs.png")
@@
 ```{r}
+# Reformat these data frames for joining
+
 base_lasso_metrics <- base_final_lasso_test_metrics %>% 
@@
 
+# Join all metrics together
 model_metrics <- bind_rows(base_lasso_metrics, lasso_metrics, lag_lasso_metrics, base_final_boost_metrics, final_boost_metrics, lag_boost_metrics, base_final_forest_metrics, final_forest_metrics, lag_forest_metrics)
 
+# Pivot and remove unncessary columns
 model_metrics <- model_metrics %>%
@@
 ```{r}
-library(gt)
-
+# Create gt table
 model_table <- model_metrics %>%
@@
 
-library(gtExtras)
-library(webshot2)
-
 print(model_table)

### 2025-04-21 code/eda-code/eda.qmd
Color Corrections

@@
 ```{r}
-#load needed packages. make sure they are installed
+# load needed packages
 library(here) 
@@
 library(reshape2)
-
+library(lubridate)
+library(tidyr)
 ```
@@
 
+# Load full case data
 case_data <- readRDS(here("data", "processed-data", "processedcasedata.rds"))
 
+# Make incidence rate non negative
 data<- data %>%
@@
 
-DataVisualization
+Data Visualization
 
@@
 ```{r}
-
-density_cases <- data %>%
-  ggplot( aes(x=new_cases)) +
-    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
+# Plot density of case incidence
+density_cases <- ggplot(data, aes(x=new_cases)) +
+    geom_density(fill="#69b3a2", color="#69b3a2", alpha=0.8) +
   labs(title = "Distribution of Daily New Case Incidence", 
@@
 
+# print
 print(density_cases)
 
+# Save
 figure_file = here("results","figures","density_cases.png")
@@
 
+# plot density of case incidence over 500
 density_cases_500 <- data %>%
@@
   ggplot( aes(x=new_cases)) +
-    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
+    geom_density(fill="#69b3a2", color="#69b3a2", alpha=0.8) +
   labs(title = "Distribution of Daily New Case Incidence (Over 500)", 
@@
 
+# print
 print(density_cases_500)
 
+# save
 figure_file = here("results","figures","density_cases_500.png")
@@
 ```
+
 A histogram of population density was created.
@@
 ```{r}
+# average population density throughout all three years
 data_summary <- data %>%
@@
 
+# plot summary of population density
 histo_pop_density <- ggplot(data_summary, aes(x = pop_density)) + 
-  geom_histogram(bins = 30, fill = "#69b3a2", color = "black", alpha = 0.7) +
+  geom_histogram(bins = 30, fill = "#69b3a2", color = "#69b3a2", alpha = 0.7) +
   labs(title = "Histogram of Population Density (2-Year Average per County)", 
@@
 
+# print
 print(histo_pop_density)
 
+# save
 figure_file = here("results","figures","histo_pop_density.png")
@@
   ungroup() %>%
-  arrange(date) %>%  # Ensure dates are sorted
+  arrange(date) %>%
   mutate(
@@
     residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
+  )
+
+# Set custom colors
+custom_colors <- c(
+  "Grocery & Pharmacy" = "#69b3a2",
+  "Workplaces" = "#BFA2DB",
+  "Residential" = "#F2C29B",
+  "Retail" = "#DBA2A2"
 )
@@
   geom_line(aes(y = retail_mean_7day, color = "Retail")) +
+  scale_color_manual(values = custom_colors) +
   labs(
@@
 
+# print
 print(mobility_graph)
@@
 # Create one plot per mobility type
-p1 <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
-  geom_line(color = "steelblue") +
+grocery <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
+  geom_line(color = "#69b3a2") +
   labs(title = "Grocery & Pharmacy", x = "Date", y = "7-Day Avg % Change") +
@@
 
-p2 <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
-  geom_line(color = "darkred") +
+workplace <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
+  geom_line(color = "#BFA2DB") +
   labs(title = "Workplaces", x = "Date", y = "7-Day Avg % Change") +
@@
 
-p3 <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
-  geom_line(color = "forestgreen") +
+resident <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
+  geom_line(color = "#F2C29B") +
   labs(title = "Residential", x = "Date", y = "7-Day Avg % Change") +
@@
 
-p4 <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
-  geom_line(color = "purple") +
+retail <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
+  geom_line(color = "#DBA2A2") +
   labs(title = "Retail", x = "Date", y = "7-Day Avg % Change") +
@@
 
-# Arrange in a 2x2 grid
-mobility_grid <- (p1 | p2) / (p3 | p4)
+# Combine in 2x2
+mobility_grid <- (grocery | workplace) / (resident | retail)
 
-# Show it
+# print for visual inspection
 print(mobility_grid)
@@
 
+# Get first and last date of data
 first_date <- min(data$date)
@@
 
+# Create graph of total case count through the 3 years
 cases_graph <- case_data %>%
@@
    scale_x_date(
-      breaks = seq(from = first_date, to = last_date, by = "6 months"),  # Breaks every 3 months
-      labels = date_format("%b %Y")  # Format labels as month year (e.g., Jan 2020)
+      breaks = seq(from = first_date, to = last_date, by = "6 months"),
+      labels = date_format("%b %Y")
     ) +
@@
 
+# print
 print(cases_graph)
 
+# save
 figure_file = here("results","figures","cases_graph.png")
@@
 
+# plot incidence
 incidence_graph <- data %>%
@@
      scale_x_date(
-      date_breaks = "3 months",  # Automatically set the breaks to every 3 months
-      date_labels = "%b %Y"      # Format labels as month-year (e.g., Jan 2020)
+      date_breaks = "3 months",
+      date_labels = "%b %Y"
     ) +
@@
 
+# print
 print(incidence_graph)
 
+# save
 figure_file = here("results","figures","incidence_graph.png")
@@
 ```{r}
-
+# create df with total case count at the end of the year
 year_end_cases <- data %>%
@@
 
+# pick population density of the years
 pop_density_info <- data %>%
@@
 
+# combine dfs for ploting
 plot_df <- left_join(year_end_cases, pop_density_info, by = c("county", "year"))
 
+# plot
 year_end_cases_graph <- ggplot(plot_df, aes(x = pop_density, y = total_cases)) +
@@
 
+# print
 print(year_end_cases_graph)
 
+# save
 figure_file = here("results","figures","year_end_cases_graph.png")
@@
 ```{r}
-
+# create lagged variables
 data <- data %>%
@@
 
+# Drop NAs made as a result
 data <- data %>% drop_na()
@@
   geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
-  scale_color_gradient2(low = "#BFA2DB", high = "#92C7CF", mid = "white",
+  scale_color_gradient2(low = "#BFA2DB", high = "#69b3a2", mid = "white",
                         midpoint = 0, limit = c(-1, 1), name = "Correlation") +
@@
 
+# Print
 print(matrix)
 
+# save
 figure_file = here("results","figures","matrix.png")

### 2025-04-21 code/processing-code/processing-file.qmd
Color Corrections

@@
 library(skimr)
-
 library(tigris)

### 2025-04-25 review-zuber/project-review-template.qmd
Project part 5 review Mohammed Zuber

---
title: Project Review  
author: Mohammed Zuber
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Assessing the Added Predictive Power of Google’s Community Mobility Report for COVID-19 Forecasting on a County-Level

Name of project author(s): Vincent Nguyen

Name of project reviewer: Mohammed Zuber

# Score
Overall Score: 3 (Sufficient)
The project is complete, well-executed, and thoughtfully designed. Minor improvements in documentation and runtime optimization could be addressed, but the core analysis is sound and meets all expectations.

# Specific project content evaluation

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The manuscript provides a clear and well-structured background on the COVID-19 pandemic, the relevance of mobility dynamics, and the utility of machine learning in disease modeling. The rationale for exploring the predictive power of mobility data is strong, especially given the novelty of such data during the pandemic. The introduction thoroughly places the study in the context of existing literature and real-world relevance.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments
The objective of evaluating whether mobility dynamics add predictive value to COVID-19 forecasting is clearly stated. The alignment between the problem statement, data structure, and modeling goals is well articulated.

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Feedback and Comments:
The sources of the data (Google Mobility Reports and Johns Hopkins COVID-19 data) are clearly cited, and the construction of the final dataset is well explained, including the filtering criteria and time frame. However, the project folder appears to lack a codebook or variable dictionary, which would be important for clearly understanding the features used in modeling. Additionally, the README files in the processing-code, eda-code, and analysis-code folders should be updated to describe what each script does, in what order they should be run, and any dependencies required. This would greatly improve transparency and facilitate easier replication or adaptation of the analysis.

### Summary assessment
* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The manuscript shows thoughtful data wrangling, including filtering for completeness and temporal consistency. The exploratory analysis is comprehensive, with visualizations illustrating trends in mobility and case counts, as well as skewness and kurtosis in the data. These insights informed the choice of models that can handle non-normality.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The use of LASSO, Random Forest, and XGBoost is well justified given the data properties. The rationale for excluding ARIMA due to non-stationarity is appropriate. Cross-validation using a rolling origin design reflects real-world forecasting logic. Feature engineering with lagged predictors is methodologically sound.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The figures and tables are clear and appropriate for illustrating model performance and data trends.

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion correctly interprets findings in relation to the modeling results and existing literature. The strengths and limitations are explicitly addressed in a dedicated section, which adds to the completeness of the manuscript. It might help to more clearly tie model limitations (e.g., RF overfitting) to specific data properties for a deeper reflection.

### Summary assessment
* strong, complete and clear discussion

## Further comments

Running the full analysis-code file took more than 2 hours to complete.

### 2025-04-25 review-zuber/project-review-template.qmd

@@
 Running the full analysis-code file took more than 2 hours to complete.
+
+## Reproducibility
+Are all results fully reproducible?
+Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention?
+Are you able to re-run the whole analysis without having to do manual interventions/edits?
+
+### Feedback and Comments
+
+The project is fully reproducible, with all code executing smoothly and generating the expected outputs, including figures and tables.
+The analysis runs without any need for manual intervention, and all necessary data files and scripts are included in the repository.
+This reflects strong adherence to reproducible research standards.
+
+### Summary assessment
+* fully reproducible without issues
+
+## Thoroughness
+How thorough was the overall study?
+Were alternatives (e.g.
+different ways of processing the data or different models) considered?
+Were alternatives discussed?
+Were the questions/hypotheses fully and thoroughly addressed?
+
+### Feedback and Comments
+
+The project demonstrates a strong level of thoroughness. Multiple modeling approaches were considered.
+
+### Summary assessment
+* strong level of thorougness
+
+## Further comments
+
+Overall, Great Job!

### 2025-05-01 code/analysis-code/statisitical-analysis.qmd
Part 6

@@
 range(test_data$date)
-
-```
-
-Recipe setting for modeling
+```
+
+Recipe setting for modeling. Also rename the variable names.
 ```{r}
 # Recipe for baseline model
-recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data)
+recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data) %>% step_rename(
+  Population_Density = pop_density,
+    Lag_1_Day = lag_1,
+    Lag_7_Days = lag_7,
+    Lag_14_Days = lag_14
+)
 
@@
                    residential_percent_change_from_baseline +
-                     lag_1 + lag_7 + lag_14, data = data)
+                   lag_1 + lag_7 + lag_14, data = data) %>%
+  step_rename(
+    Population_Density = pop_density,
+    Retail_Mobility = retail_and_recreation_percent_change_from_baseline,
+    Grocery_Mobility = grocery_and_pharmacy_percent_change_from_baseline,
+    Workplace_Mobility = workplaces_percent_change_from_baseline,
+    Residential_Mobility = residential_percent_change_from_baseline,
+    Lag_1_Day = lag_1,
+    Lag_7_Days = lag_7,
+    Lag_14_Days = lag_14
+  )
 
 # Recipe with lagged versions of mobility
-recipe_lag <- recipe(new_cases ~ pop_density + retail_lag7 + grocery_lag7 + residential_lag7 + workplace_lag7 + lag_1 + lag_7 + lag_14, data = data)
+recipe_lag <- recipe(new_cases ~ pop_density + retail_lag7 + grocery_lag7 + residential_lag7 + workplace_lag7 + lag_1 + lag_7 + lag_14, data = data) %>%
+  step_rename(
+    Population_Density = pop_density,
+    Retail_Mobility_Lag7 = retail_lag7,
+    Grocery_Mobility_Lag7 = grocery_lag7,
+    Residential_Mobility_Lag7 = residential_lag7,
+    Workplace_Mobility_Lag7 = workplace_lag7,
+    Lag_1_Day = lag_1,
+    Lag_7_Days = lag_7,
+    Lag_14_Days = lag_14
+  )
 ```
@@
 print(base_final_lasso_test_metrics)
+```
+
+Collect Metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+base_final_lasso_cv_metrics <- collect_metrics(base_final_lasso_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare Test metrics to match
+base_final_lasso_test_metrics_fixed <- base_final_lasso_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_base_lasso_metrics <- bind_rows(
+  base_final_lasso_cv_metrics,
+  base_final_lasso_test_metrics_fixed
+)
+
+# View the final combined metrics
+print(all_base_lasso_metrics)
 
 # View variable importance for the Lasso model
-vip(base_final_lasso_fit$fit$fit)
-
+vip_base_lasso <- vip(base_final_lasso_fit$fit$fit)
+
+# Alter aesthetics
+vip_base_lasso <- vip_base_lasso + labs(
+  title = "Baseline LASSO Variable Importance",
+  x = "Importance",
+  y = "Feature"
+) + theme_bw()
+
+figure_file = here("results", "figures", "vip_base_lasso.png")
+ggsave(filename = figure_file, plot = vip_base_lasso, width = 8, height = 6, dpi = 300)
 ```
@@
 print(final_lasso_test_metrics)
-
-# View variable importance
-vip(final_lasso_fit$fit$fit)
+```
+
+Collect Metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+final_lasso_cv_metrics <- collect_metrics(final_lasso_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare Test metrics to match
+final_lasso_test_metrics_fixed <- final_lasso_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_lasso_metrics <- bind_rows(
+  final_lasso_cv_metrics,
+  final_lasso_test_metrics_fixed
+)
+
+# View the final combined metrics
+print(all_lasso_metrics)
+
+# Plot VIP
+vip_full_lasso <- vip(final_lasso_fit$fit$fit)
+
+# Alter aesthetics
+vip_full_lasso <- vip_full_lasso + 
+  labs(
+    title = "LASSO with Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Define save location
+figure_file <- here("results", "figures", "vip_full_lasso.png")
+
+# Save plot
+ggsave(filename = figure_file, plot = vip_full_lasso, width = 8, height = 6, dpi = 300)
 ```
@@
 print(lag_final_lasso_test_metrics)
-
-# View variable importance
-vip(lag_final_lasso_fit$fit$fit)
+```
+
+Collect Metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+lag_final_lasso_cv_metrics <- collect_metrics(lag_final_lasso_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare Test metrics to match
+lag_final_lasso_test_metrics_fixed <- lag_final_lasso_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_lag_lasso_metrics <- bind_rows(
+  lag_final_lasso_cv_metrics,
+  lag_final_lasso_test_metrics_fixed
+)
+
+# View the final combined metrics
+print(all_lag_lasso_metrics)
+
+# Plot VIP
+vip_lag_lasso <- vip(lag_final_lasso_fit$fit$fit)
+
+# Alter aesthetics
+vip_lag_lasso <- vip_lag_lasso + 
+  labs(
+    title = "LASSO with Lagged Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Define save location
+figure_file <- here("results", "figures", "vip_lag_lasso.png")
+
+# Save plot
+ggsave(filename = figure_file, plot = vip_lag_lasso, width = 8, height = 6, dpi = 300)
 
@@
     plot.margin = margin(10, 10, 10, 10),
-    plot.title = element_text(size = 14, hjust = 0.5),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10)
+    plot.title = element_text(size = 20, hjust = 0.5),
+    axis.title = element_text(size = 18),
+    axis.text = element_text(size = 16)
   )
@@
   plot = lasso_graphs,
-  width = 16, height = 5, dpi = 300
-)
-
+  width = 18, height = 6, dpi = 400
+)
 ```
@@
 ) %>%
-  set_engine("ranger", seed = rngseed)
+  set_engine("ranger", seed = rngseed, importance = "impurity")
 
@@
 print(base_final_rf_test_metrics)
-
+```
+
+Collect metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+base_final_rf_cv_metrics <- collect_metrics(base_final_rf_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+base_final_rf_test_metrics_fixed <- base_final_rf_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_base_rf_metrics <- bind_rows(
+  base_final_rf_cv_metrics,
+  base_final_rf_test_metrics_fixed
+)
+
+# Print metrics
+print(all_base_rf_metrics)
+
+# Create vip
+vip_base_rf <- vip(base_final_rf_fit)
+
+# Alter aesthetics
+vip_base_rf <- vip_base_rf + 
+  labs(
+    title = "Baseline Random Forest Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Save vip
+figure_file <- here("results", "figures", "vip_base_rf.png")
+ggsave(filename = figure_file, plot = vip_base_rf, width = 8, height = 6, dpi = 300)
 ```
@@
 final_rf_wf <- workflow() %>%
-  add_recipe(recipe_baseline) %>%
+  add_recipe(recipe) %>%
   add_model(final_rf_spec)
@@
 print(final_rf_test_metrics)
+```
+
+Collect metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+final_rf_cv_metrics <- collect_metrics(final_rf_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+final_rf_test_metrics_fixed <- final_rf_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_rf_metrics <- bind_rows(
+  final_rf_cv_metrics,
+  final_rf_test_metrics_fixed
+)
+
+# Print metrics
+print(all_rf_metrics)
+
+# Create vip
+vip_full_rf <- vip(final_rf_fit)
+
+# Alter aesthetics
+vip_full_rf <- vip_full_rf + 
+  labs(
+    title = "Random Forest with Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Save vip
+figure_file <- here("results", "figures", "vip_full_rf.png")
+ggsave(filename = figure_file, plot = vip_full_rf, width = 8, height = 6, dpi = 300)
 ```
@@
 lag_final_rf_wf <- workflow() %>%
-  add_recipe(recipe_baseline) %>%
+  add_recipe(recipe_lag) %>%
   add_model(lag_final_rf_spec)
@@
 print(lag_final_rf_test_metrics)
-
+```
+
+Collect metrics and CV
+```{r}
+# Collect CV metrics and drop extra columns
+lag_final_rf_cv_metrics <- collect_metrics(lag_final_rf_res) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+lag_final_rf_test_metrics_fixed <- lag_final_rf_test_metrics %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_lag_rf_metrics <- bind_rows(
+  lag_final_rf_cv_metrics,
+  lag_final_rf_test_metrics_fixed
+)
+
+# Print metrics
+print(all_lag_rf_metrics)
+
+# Create vip
+vip_lag_rf <- vip(lag_final_rf_fit)
+
+# Alter aesthetics
+vip_lag_rf <- vip_lag_rf + 
+  labs(
+    title = "Random Forest with Lagged Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Save vip
+figure_file <- here("results", "figures", "vip_lag_rf.png")
+ggsave(filename = figure_file, plot = vip_lag_rf, width = 8, height = 6, dpi = 300)
 ```
@@
   plot = rf_graphs,
-  width = 16, height = 5, dpi = 300
+  width = 18, height = 6, dpi = 400
 )
@@
 
+autoplot(base_boost_tune_res)
+
 base_best_params <- select_best(base_boost_tune_res, metric = "rsq")
@@
 print(base_final_boost_metrics)
+
+# Extract Model
+base_boost_model <- base_final_boost_fit$fit$fit
+```
+
+Collect metrics and VIP
+```{r}
+
+# select metrics from best model
+base_boost_cv_metrics <- base_boost_tune_res %>%
+  collect_metrics() %>%
+  inner_join(base_best_params, by = c("tree_depth", "learn_rate", "loss_reduction")) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+base_final_boost_test_metrics <- base_final_boost_results %>%
+  metrics(truth = truth, estimate = predicted) %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_base_boost_metrics <- bind_rows(
+  base_boost_cv_metrics,
+  base_final_boost_test_metrics
+)
+
+# Print metrics
+print(all_base_boost_metrics)
+
+# Create vip
+vip_base_boost <- vip(base_boost_model)
+
+# Alter aesthetics
+vip_base_boost <- vip_base_boost + 
+  labs(
+    title = "Baseline XGBoost Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Print
+print(vip_base_boost)
+
+# Save vip
+figure_file <- here("results", "figures", "vip_base_boost.png")
+ggsave(filename = figure_file, plot = vip_base_boost, width = 8, height = 6, dpi = 300)
 ```
@@
   print()
+autoplot(boost_tune_res)
 
@@
 final_boost_wf <- workflow() %>%
-  add_recipe(recipe_baseline) %>%
+  add_recipe(recipe) %>%
   add_model(final_boost_spec)
@@
 
+# Extract Model
+full_boost_model <- final_boost_fit$fit$fit
+```
+
+Collect metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+boost_cv_metrics <- boost_tune_res %>%
+  collect_metrics() %>%
+  inner_join(best_params, by = c("tree_depth", "learn_rate", "loss_reduction")) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+final_boost_test_metrics <- final_boost_results %>%
+  metrics(truth = truth, estimate = predicted) %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_boost_metrics <- bind_rows(
+  boost_cv_metrics,
+  final_boost_test_metrics
+)
+
+# Print metrics
+print(all_boost_metrics)
+
+# Create vip
+vip_full_boost <- vip(full_boost_model)
+
+# Alter aesthetics
+vip_full_boost <- vip_full_boost + 
+  labs(
+    title = "XGBoost with Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Save vip
+figure_file <- here("results", "figures", "vip_full_boost.png")
+ggsave(filename = figure_file, plot = vip_full_boost, width = 8, height = 6, dpi = 300)
 ```
@@
 lag_final_boost_wf <- workflow() %>%
-  add_recipe(recipe_baseline) %>%
+  add_recipe(recipe_lag) %>%
   add_model(lag_final_boost_spec)
@@
 print(lag_final_boost_metrics)
+
+# Extract Model
+lag_boost_model <- lag_final_boost_fit$fit$fit
+```
+
+Collect metrics and VIP
+```{r}
+# Collect CV metrics and drop extra columns
+lag_boost_cv_metrics <- lag_boost_tune_res %>%
+  collect_metrics() %>%
+  inner_join(lag_best_params, by = c("tree_depth", "learn_rate", "loss_reduction")) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Train")
+
+# Prepare test metrics to match format
+lag_final_boost_test_metrics <- lag_final_boost_results %>%
+  metrics(truth = truth, estimate = predicted) %>%
+  mutate(mean = .estimate) %>%
+  select(.metric, .estimator, mean) %>%
+  mutate(Data = "Test")
+
+# Combine CV and Test metrics
+all_lag_boost_metrics <- bind_rows(
+  lag_boost_cv_metrics,
+  lag_final_boost_test_metrics
+)
+
+# Print metrics
+print(all_lag_boost_metrics)
+
+# Create vip
+vip_lag_boost <- vip(lag_boost_model)
+
+# Alter aesthetics
+vip_lag_boost <- vip_lag_boost + 
+  labs(
+    title = "XGBoost with Lagged Mobility Variable Importance",
+    x = "Importance",
+    y = "Feature"
+  ) + 
+  theme_bw()
+
+# Print
+print(vip_lag_boost)
+
+# Save vip
+figure_file <- here("results", "figures", "vip_lag_boost.png")
+ggsave(filename = figure_file, plot = vip_lag_boost, width = 8, height = 6, dpi = 300)
 ```
@@
   plot = boost_graphs,
-  width = 16, height = 5, dpi = 300
+  width = 18, height = 6, dpi = 400
 )
@@
 
-base_lasso_metrics <- base_final_lasso_test_metrics %>% 
-  mutate(model = "LASSO (Baseline)") %>%
-  rename(mean = .estimate)
-
-lasso_metrics <- final_lasso_test_metrics %>% 
-  mutate(model = "LASSO") %>%
-  rename(mean = .estimate)
-
-lag_lasso_metrics <- lag_final_lasso_test_metrics %>% 
-  mutate(model = "LASSO (Lagged Mobility)") %>%
-  rename(mean = .estimate)
-
-final_boost_metrics <- final_boost_metrics %>%
-  mutate(model = "XGBoost") %>%
-  rename(mean = .estimate)
-
-final_forest_metrics <- final_rf_test_metrics %>%
-  mutate(model = "Random Forest") %>%
-  rename(mean = .estimate)
-
-base_final_boost_metrics <- base_final_boost_metrics %>%
-  mutate(model = "XGBoost (Baseline)") %>%
-  rename(mean = .estimate)
-
-base_final_forest_metrics <- base_final_rf_test_metrics %>%
-  mutate(model = "Random Forest (Baseline)") %>%
-  rename(mean = .estimate)
-
-lag_boost_metrics <- lag_final_boost_metrics %>%
-  mutate(model = "XGBoost (Lagged Mobility)") %>%
-  rename(mean = .estimate)
-
-lag_forest_metrics <- lag_final_rf_test_metrics %>%
-  mutate(model = "Random Forest (Lagged Mobility)") %>%
-  rename(mean = .estimate)
+base_lasso_metrics <- all_base_lasso_metrics %>% 
+  mutate(model = "LASSO (Baseline)")
+
+lasso_metrics <- all_lasso_metrics %>% 
+  mutate(model = "LASSO")
+
+lag_lasso_metrics <- all_lag_lasso_metrics %>% 
+  mutate(model = "LASSO (Lagged Mobility)")
+
+base_final_boost_metrics <- all_base_boost_metrics %>%
+  mutate(model = "XGBoost (Baseline)")
+
+final_boost_metrics <- all_boost_metrics %>%
+  mutate(model = "XGBoost")
+
+lag_boost_metrics <- all_lag_boost_metrics %>%
+  mutate(model = "XGBoost (Lagged Mobility)")
+
+base_final_forest_metrics <- all_base_rf_metrics %>%
+  mutate(model = "Random Forest (Baseline)")
+
+final_forest_metrics <- all_rf_metrics %>%
+  mutate(model = "Random Forest")
+
+lag_forest_metrics <- all_lag_rf_metrics %>%
+  mutate(model = "Random Forest (Lagged Mobility)")
 
@@
   select(-.estimator) %>%
-  pivot_wider(names_from = '.metric', values_from = mean)
+  pivot_wider(names_from = c('.metric', 'Data'), values_from = mean)
+```
+
+```{r}
+model_table <- model_metrics %>%
+  select(model, rmse_Train, mae_Train, rsq_Train, rmse_Test, mae_Test, rsq_Test) %>%
+  gt() %>%
+  tab_header(
+    title = "Model Performance Metrics",
+    subtitle = "Train and Test Results | Best Performance Highlighted in Green"
+  ) %>%
+  fmt_number(columns = everything(), decimals = 3) %>%
+  cols_label(
+    model = "Model",
+    rmse_Train = "RMSE",
+    mae_Train = "MAE",
+    rsq_Train = "R²",
+    rmse_Test = "RMSE",
+    mae_Test = "MAE",
+    rsq_Test = "R²"
+  ) %>%
+  tab_spanner(
+    label = "Train Metrics",
+    columns = c(rmse_Train, mae_Train, rsq_Train)
+  ) %>%
+  tab_spanner(
+    label = "Test Metrics",
+    columns = c(rmse_Test, mae_Test, rsq_Test)
+  ) %>%
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rmse_Test,
+      rows = rmse_Test == min(rmse_Test, na.rm = TRUE)
+    )
+  ) %>%
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = mae_Test,
+      rows = mae_Test == min(mae_Test, na.rm = TRUE)
+    )
+  ) %>%
+  tab_style(
+    style = cell_fill(color = "lightgreen"),
+    locations = cells_body(
+      columns = rsq_Test,
+      rows = rsq_Test == max(rsq_Test, na.rm = TRUE)
+    )
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_options(
+    table.width = pct(100),
+    table.align = "left",
+    column_labels.padding = px(10),
+    data_row.padding = px(6)
+  )
+
+print(model_table)
+
+gtsave(model_table, here("results", "figures", "models.png"))
 ```
@@
   tab_header(
-    title = "Model Performance Metrics",
+    title = "Model Performance Metrics on Data",
     subtitle = "Best values are highlighted per metric"

### 2025-05-01 products/manuscript/Manuscript.qmd
Part 6

--- 2025-04-21 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Assessing the Added Predictive Power of Google's Community Mobility Report for COVID-19 Forecasting on a County-Level"
+title: "Assessing the Explanatory and Predictive Power of Google's Community Mobility Report for COVID-19 Forecasting on a County-Level"
 author: "Vincent Nguyen"
@@
 
-```{r}
+```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
 library(here)
@@
 **Background**: Human mobility patterns shifted significantly throughout the COVID-19 pandemic due to public health policies and individual risk perception.
-These changes, recorded in Google’s Community Mobility Report, may hold predictive value in understanding and forecasting transmission dynamics.
+These changes, recorded in Google’s Community Mobility Report, may hold value in understanding transmission and forecasting cases.
 
-**Objective:** This analysis evaluated the additive predictive capabilities mobility dynamics may have on modeling daily COVID-19 case incidence at the county level in Georgia.
+**Objective:** This analysis evaluated the additive explanatory and predictive performance mobility dynamics may have on modeling daily COVID-19 case incidence at the county level in Georgia.
 
-**Methods:** Daily case data from John Hopkins University and mobility predictors from Google were used to model case incidence across 44 Georgia counties between March 14th 2020 and March 14th 2022.
-Three modeling approaches, LASSO regression, Random Forest, and XGBoost, were applied to assess the association.
-A model was trained on three sets of feature sets, a baseline set (lagged case counts and population density), a full set (adding on mobility variables), and a 7-day lagged version of the full set.
-Models were evaluated using RMSE, R-squared, and MAE metrics on a 30 day test set, with rolling window cross-validation used during training.
+**Methods:** Daily case data from John Hopkins University and mobility predictors from Google's Community Mobility Report (CMR) were used to develop machine learning models with the objective of explaining case incidence across 44 Georgia counties between March 14th 2020 and March 14th 2022.
+Three modeling approaches, LASSO regression, Random Forest, and XGBoost, were applied to investigate the association.
+Models were trained on three feature sets, a baseline set (lagged case counts and population density), a set including the mobility predictors, and a set including 7-day lagged versions of the mobility predictors.
+Models were evaluated using RMSE, R², and MAE metrics on a 30 day test set, with rolling window cross-validation used during training.
 
-**Results:** The XGBoost model, with the lagged feature set, achieved the best overall performance (RMSE: 37.9; R²: 0.836; MAE: 23.5).
-This model’s improvement over other feature sets indicates significant additive predictive power of lagged mobility dynamics when forecasting case incidence.
+**Results:** In LASSO and Random Forest models, the inclusion of mobility and lagged mobility predictors resulted in significant reductions of prediction error metrics (MAE and RMSE).
+LASSO with mobility predictors achieved the highest R² (0.843) and identified mobility trends as key drivers of case incidence.
+Explanatory power, R², did not improve significantly in Random Forest and XGBoost, suggesting that mobility alone does not fully explain variation in daily new case incidence.
 
-**Conclusion:** This analysis suggest that incorporation of time-lagged mobility data can improve the predictive performance of machine learning models for COVID-19 case incidence.
-The results highlight the potential in assessing behavioral data in forecasting infectious disease dynamics and support the integration of mobility metrics in public health surveillance.
+**Conclusion:** This analysis suggests that the incorporation of time-lagged mobility data can improve the predictive performance of some machine learning models for COVID-19 case incidence, particularly in reducing prediction error metrics like MAE or RMSE.
+While the improvement in explanatory power was limited, the results highlight the potential in utilizing behavioral data in forecasting disease transmission dynamics and the integration of mobility metrics in public health surveillance.
 
@@
 
-The emergence of the novel corona-virus SARS-CoV-2 caused one of the most significant global health crises in modern history.
-First identified in late 2019, the virus rapidly spread across the world due to its high transmissiblity and global interconnectedness, leading the World Health Organization to declare it a pandemic on March 11, 2020.
-Two days later, on March 13th, 2020, the United States declared a nationwide emergency due to the COVID-19 pandemic, marking the biggest virus outbreak since the 1916 influenza pandemic.
-In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lock-down procedures [@centersfordiseasecontrolandpreventionCOVID19Timeline2024].
+The emergence of the novel corona-virus SARS-CoV-2 brought about the onset of the one of the most significant global health crises in modern history.
+First identified in late 2019, the virus rapidly spread across the world due to its high transmissibility and global interconnectedness, leading the World Health Organization to declare it a pandemic on March 11, 2020.
+Two days later, on March 13th, 2020, the United States declared the COVID-19 pandemic a national emergency, marking the biggest virus outbreak since the 1916 influenza pandemic.
+In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lock-down procedures [@centersfordiseasecontrolandpreventionCOVID19Timeline2024e].
 
@@
 
-Human mobility, a key driver of respiratory disease transmission, shifted significantly during the pandemic in response to public health policies and disease risk perception [@paltraEffectMobilityReductions2024].
+Human mobility, a key driver of respiratory disease transmission, shifted significantly during the pandemic in response to public health policies and disease risk perception [@paltraEffectMobilityReductions2024e].
 Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lock-downs.
 Lock-downs involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings.
-Looking at the Wikipedia page shows a lack of prior history implementing lock-downs.
 Beyond these measures which lasted only a few weeks in Georgia, individual risk perception played into the compliance of other preventative behaviors (masking, social distancing, etc.).
-The variability of risk individual perception has led to complex mobility patterns during the pandemic, for example, surges in mobility amidst large case outbreaks [@cipollettaRiskPerceptionCOVID192022].
-In order to analyze and quantify this relationship, machine learning models were adapted.
+The variability of risk individual perception has led to complex mobility patterns during the pandemic, for example, surges in mobility amidst large case outbreaks [@cipollettaRiskPerceptionCOVID192022e].
+Previous analysis has suggested that mobility patterns are correlated with decreases in COVID-19 case growth rates [@badrAssociationMobilityPatterns2020b].
+In order to further explore and quantify this relationship, machine learning models were adapted.
 
@@
 These models aim to also approximate the relationship between outputs and inputs, quantify predictions, or approximate classification tasks.
-Beyond regular statistical analysis, machine learning methods offer powerful tools for forecasting infectious disease trends by identifying complex, nonlinear relationships between predictors and outcomes [@rashidiCommonStatisticalConcepts2023].
-Some common models in disease forecasting include ARMA, ARIMA, LASSO, XGBoost, and various neural network techniques [@alfredRolesMachineLearning2021].
-Several authors have utilized case counts, estimations, demographics, and more to forecast disease trends [@ogunjoPredictingCOVID19Cases2022].
-However, the predictive value of real-time mobility data, particularly in the context of an evolving pandemic, has not been assessed through modeling.
-This study aims to explore whether mobility dynamics, form Google’s Community Mobility Reports, can enhance predictive performance when modeling COVID-19 transmission at the county level.
+Beyond regular statistical analysis, machine learning methods offer powerful tools for forecasting infectious disease trends by identifying complex, nonlinear relationships between predictors and outcomes [@rashidiCommonStatisticalConcepts2023e].
+Some common models in disease forecasting include ARMA, ARIMA, LASSO, XGBoost, and various neural network techniques [@alfredRolesMachineLearning2021e].
+Several authors have utilized case counts, estimations, demographics, and more to forecast disease trends [@ogunjoPredictingCOVID19Cases2022e].
+
+## Previous Literature
+
+The Google CMR have been widely utilized in machine learning research.
+As noted by @iriniAssociationsMobilityPatterns2021b, Google's CMR have been integrated into diverse machine learning algorithms to predict a range of outcomes, such as case incidence, outcomes, drug development, and more.
+Studies have demonstrated associations between mobility (as measured by Google) and case incidence in areas such as Jakarta, as well as COVID-19 death in Ireland [[@nandaCommunityMobilityCOVID192022b @iriniAssociationsMobilityPatterns2021b]].
+@paezUsingGoogleCommunity2020b also employed multivariate regression to investigate the report's potential in COVID-19 incidence.
+Overall, the CMR data has been validated across several international contexts and in various regression tasks.
+However, to our knowledge, this data has yet to be applied in a county-level American context and while utilizing modeling methods like Random Forest and XGBoost.
+
+In addition to the predictors found in the CMR, lagged mobility predictors will also be created, given their previously demonstrated relevance [@sulyokCommunityMovementCOVID192020b].
 
@@
 
-During the pandemic, Google began to collect aggregated, anonymized data from users utilizing Google products (apps, phones, etc.) to track changes in mobility [@googleCOVID19CommunityMobility2020].
+During the pandemic, Google began to collect aggregated, anonymized data from users utilizing Google products (apps, phones, etc.) to track changes in mobility [@googleCOVID19CommunityMobility2020e].
 Within the data, Google measures mobility as a percent change difference from baseline measurements; for example, a -45% change in retail and recreation mobility indicates a 45% reduction in movement to those categorical locations.
@@
 Additionally, two mobility metrics, transit stations and recreational parks, were omitted in this analysis due to incompleteness.
+According to Google, recreational parks are specifically intended to mean official national parks [@googleOverviewCommunityMobility2020b].
+Georgia, specifically, only has 11 national parks  [@nationalparkserviceGeorgiaUSNational2017b] .
 
@@
 The COVID-19 Data Repository by the Center for Systems Science and Engineering at Johns Hopkins University is a comprehensive data set that tracks global COVID-19 cases, recoveries, and deaths.
-This data was recorded daily for several years across every county in the US [@cssegisanddataCOVID19DataRepository2022].
+This data was recorded daily for several years across every county in the US [@cssegisanddataCOVID19DataRepository2022e].
 
@@
 The data was filtered from March 14th 2020 to March 14th 2022.
-This was selected specifically as March 14th marks the date of Governor Brian Kemp’s announcement of Georgia's Public Health State of Emergency [@hartCOVID19PandemicGeorgia2025].
+This was selected specifically as March 14th marks the date of Governor Brian Kemp’s announcement of Georgia's Public Health State of Emergency [@hartCOVID19PandemicGeorgia2025e].
+Imputation strategies were not considered as data was not missing at random; missing data tended to come from specific counties rather than being randomly distributed across the data set.
+Imputing these values could obscure important patterns or misrepresent the complexitiy of the pandemic.
+Exclusion of these counties was chosen instead.
 
@@
 
-As informed by @alfredRolesMachineLearning2021, machine learning models have wide application in forecasting outbreaks and disease incidence.
+As informed by @alfredRolesMachineLearning2021e, machine learning models have wide application in forecasting outbreaks and disease incidence.
 In their review, they outline several common applications of regression and classification models.
@@
 Previous iterations of the analysis indicated that the data was non-normal and non-stationary which violated assumptions in common time series models like ARIMA.
-Linear regression was chosen as a standard model while Random Forest and XGBoost were chosen for their lack of assumptions required for analysis.
+LASSO regression was chosen for its previously demonstrated relevance [@nandaCommunityMobilityCOVID192022b] Random Forest and XGBoost were chosen for their lack of assumptions required for analysis.
 
@@
 Also known as Least Absolute Shrinkage and Selection Operator, Lasso is a regularization technique used in regression modeling to prevent overfitting and improve model interpretability.
-Lasso adds a penalty term equal to the absolute value of the magnitude of coefficients to the loss function [@ibmLassoRegression2024].
+Lasso adds a penalty term equal to the absolute value of the magnitude of coefficients to the loss function [@ibmLassoRegression2024d].
 This penalty can shrink coefficients to zero, effectively performing variable selection.
@@
 These splits form a tree-like structure where each internal node represents a decision based on a feature, each branch corresponds to an outcome of the decision, and each leaf node represents a predicted value.
-Decision trees are powerful and can capture nonlinear complex relationships, however, can overfit training data.  Random forests are an average of these trees to produce a final prediction and can produce more stable/accurate models as a result [@breimanRandomForests2001].
+Decision trees are powerful and can capture nonlinear complex relationships, however, can overfit training data.  Random forests are an average of these trees to produce a final prediction and can produce more stable/accurate models as a result [@breimanRandomForests2001e].
 
@@
 This technique is known as boosting.
-In simple terms, XGBoost makes predictions with a single tree and looks at the errors and creates a second tree focused on minimizing the errors [@chenXGBoostScalableTree2016].
+In simple terms, XGBoost makes predictions with a single tree and looks at the errors and creates a second tree focused on minimizing the errors [@chenXGBoostScalableTree2016e].
 
@@
 The first set, the baseline predictors, included population density and 3 spaced out lagged variables for case counts (1 day, 7 days, and 14 days).
-These were chosen because of  their known influence on disease transmission [@ogunjoPredictingCOVID19Cases2022].
+These were chosen because of  their known influence on disease transmission [@ogunjoPredictingCOVID19Cases2022e].
 The second set, the mobility predictors, included population density, the 3 spaced out lagged case counts, and mobility indicators from Google’s report (e.g., workplace, grocery, retail, transit).
-Lastly, the third set included 7-day lagged versions of the mobiltiy predictors, population density, and the three spaced out lagged case counts.
+Lastly, the third set included 7-day lagged versions of the mobility predictors, population density, and the three spaced out lagged case counts.
 
@@
 Each fold contained a training window of four months (120 days per county) followed by a validation window of approximately one month (30 days).
+The 120 days of training and 30 days of validation is intended to mimic the 4:1 ratio of general machine learning splits [@sivakumarTradeoffTrainingTesting2024b].
 The window was advanced forward in 15-day increments which creates multiple sequential train-validation splits.
@@
 
-**R-squared**
+**R²**
 
 A statistical measure that explains the proportion of the data variation that can be explained by the model.
-A higher R-squared value generally is indicative of a better-fitting model, although could be misleading when over fitting occurs.
-Criteria for final model selection focused on achieving the highest r-squared value.
+A higher R² value generally is indicative of a better-fitting model, although could be misleading when over fitting occurs.
+Criteria for final model selection focused on achieving the highest R² value.
+R² was chosen as the final selector for best model.
+Previous literature has used R² as a measure of forecasting performance [@nandaCommunityMobilityCOVID192022b].
 
@@
 
-After the initial RF and XGBoost models were made, their hyper-parameters were tuned to improve predictive performance.
+After the models were made, their hyper-parameters were tuned to improve predictive performance.
 
@@
 The analysis was conducted on RStudio 4.3.2 on Windows 11.
-The following R packages were used: ggplot2 [@wickhamGgplot2CreateElegant2019], broom [@robinsonBroomConvertStatistical2022], here [@mullerHereSimplerWay2020], glmnet [@friedmanGlmnetLassoElasticNet2021], MASS [@ripleyMASSSupportFunctions2022], tidymodels [@kuhnTidymodelsEasilyInstall2023], dplyr [@wickhamDplyrGrammarData2020], rsample [@frickRsampleGeneralResampling2025], parsnip [@kuhnParsnipCommonAPI2025], future [@bengtssonFutureUnifiedParallel2025], vip [@greenwellVipVariableImportance2023], zoo [@zeileisZooS3Infrastructure2021], patchwork [@pedersenPatchworkComposerPlots2020], e1071 [@meyerE1071MiscFunctions2022], scales [@wickhamScalesScaleFunctions2022], RColorBrewer [@neuwirthRColorBrewerColorBrewerPalettes2022], corrplot [@weiCorrplotVisualizationCorrelation2017], reshape2 [@wickhamReshape2FlexiblyReshape2020], tidyr [@wickhamTidyrTidyMessy2020], lubridate [@spinuLubridateMakeDealing2022], readxl [@wickhamReadxlReadExcel2019], stringr [@wickhamStringrSimpleConsistent2019], skimr [@waringSkimrCompactFlexible2022], tigris [@walkerTigrisLoadCensus2025], and sf [@pebesmaSfSimpleFeatures2021]
+The following R packages were used: ggplot2 [@wickhamGgplot2CreateElegant2019e], broom [@robinsonBroomConvertStatistical2022e], here [@mullerHereSimplerWay2020e], glmnet [@friedmanGlmnetLassoElasticNet2021e], MASS [@ripleyMASSSupportFunctions2022e], tidymodels [@kuhnTidymodelsEasilyInstall2023e], dplyr [@wickhamDplyrGrammarData2020e], rsample [@frickRsampleGeneralResampling2025e], parsnip [@kuhnParsnipCommonAPI2025e], future [@bengtssonFutureUnifiedParallel2025e], vip [@greenwellVipVariableImportance2023e], zoo [@zeileisZooS3Infrastructure2021e], patchwork [@pedersenPatchworkComposerPlots2020e], e1071 [@meyerE1071MiscFunctions2022e], scales [@wickhamScalesScaleFunctions2022e], RColorBrewer [@neuwirthRColorBrewerColorBrewerPalettes2022e], corrplot [@weiCorrplotVisualizationCorrelation2017e], reshape2 [@wickhamReshape2FlexiblyReshape2020e], tidyr [@wickhamTidyrTidyMessy2020e], lubridate [@spinuLubridateMakeDealing2022e], readxl [@wickhamReadxlReadExcel2019e], stringr [@wickhamStringrSimpleConsistent2019e], skimr [@waringSkimrCompactFlexible2022e], gtExtras [@mockExtendingGtBeautiful2023b], webshot2[@changTakeScreenshotsWeb2023b], tigris [@walkerTigrisLoadCensus2025e], and sf [@pebesmaSfSimpleFeatures2021e]
 
@@
 
-## 5.1 Exploratory/Descriptive analysis
+## Exploratory/Descriptive analysis
 
@@
 Workplace mobility tended to remain below the baseline level while residential mobility increased during this period.
-There were frequent fluctations in retail and grocery mobility, possibly a result of outbreaks, changing policies, or even holiday-related activity.
+There were frequent fluctuations in retail and grocery mobility, possibly a result of outbreaks, changing policies, or even holiday-related activity.
 Levels of mobility sharply declined and would remain below baseline until a year later.
+This seasonality violates the assumption of stationarity required for many time-series modeling techniques.
 
@@
 The distribution is heavily right skewed, with most days clustered around lower incidence leels and fewer days where extremely high case transmission occurs.
-This skew suggests surges were relatively infrequent ocmapred to more moderate levels of daily incidence.
-This kurtosis can affect the performance of linear regrssion models and ARIMA where normality is an assumption whereas Random Forest and XGBoost are more robust to this skewing.
+This skew suggests surges were relatively infrequent compared to more moderate levels of daily incidence.
+This kurtosis can affect the performance of linear regression models and ARIMA where normality is an assumption whereas Random Forest and XGBoost are more robust to this skewing.
 
@@
 Population density across the included Georgia counties is also heavily right-skewed, with the majority of counties having lower densities.
-This imbalance may influence transmission dynamics and was incldued for modeling.
+This imbalance may influence transmission dynamics and was included for modeling.
 
@@
 #| label: correlation
-#| fig-cap: "Correlation Matrix of Predictors using Spearman's"
+#| fig-cap: "Figure 5.8: Correlation Matrix of Predictors using Spearman's"
 #| echo: FALSE
@@
 
+To evaluate the explanatory and predictive value of mobility trends on new COVID-19 case incidence, three modeling techniques were utilized, LASSO, Random Forest, and XGBoost.
+For each model, three variants were created based on the different set of predictors: baseline predictors, mobility predictors, and lagged mobility predictors.
+This allowed for analysis of how mobility and its delated effects can influence COVID-19 case predictions.
+
+All three models were tuned and utilized rolling origin cross-validation to respect the temporal structure of the data and prevent data leakage from the future.
+The best-performing hyperparameters were selected based on R².
+Predictions were made on the test data set and final model performance was evaluated using RMSE, R², and MAE.
+
+```{r}
+#| label: modeltable
+#| fig-cap: "Table 5.1: Model Performance"
+#| out.width: "120%"  
+#| out.height: "120%" 
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","models.png"))
+```
+The LASSO regression model with non-lagged mobility predictors had the strongest R² while XGBoost's inclusion of lagged mobility predictors had the lowest MAE and RMSE.
+
+Actual vs. predicted plots were created to visually assess model fit and diagnostic accuracy. These plots can help identify under- or over- prediction patterns.
+
 ```{r}
 #| label: lassographs
-#| fig-cap: "Plot of Actual vs Predicted Cases in Lasso Regression Models"
+#| fig-cap: "Figure 5.9: Plot of Actual vs Predicted Cases in Lasso Regression Models"
+#| out.width: "120%"  
+#| out.height: "120%" 
 #| echo: FALSE
@@
 #| label: rfgraphs
-#| fig-cap: "Plot of Actual vs Predicted Cases in Random Forest Models"
+#| fig-cap: "Figure 5.10: Plot of Actual vs Predicted Cases in Random Forest Models"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","rf_graphs.png"))
+#| out.width: "120%"  
+#| out.height: "120%" 
+knitr::include_graphics(here("results", "figures", "rf_graphs.png"))
+
 ```
@@
 #| label: boostgraphs
-#| fig-cap: "Plot of Actual vs Predicted Cases in XGBoost Models"
+#| fig-cap: "Figure 5.11: Plot of Actual vs Predicted Cases in XGBoost Models"
+#| out.width: "120%"  
+#| out.height: "120%" 
 #| echo: FALSE
@@
 
-To assess the association between mobility dynamics and COVID-19 case incidence in Georgia, three modeling approaches were utilized, LASSO regression, Random Forest, and XGBoost.
-In addition to the three modeling approaches, three feature sets were utilized, a baseline recipe lacking mobility predictors, a full recipe containing mobility predictors, and a 7-day lagged version of the full recipe.
-Model performance was assessed using RMSE, R-squared, and MAE on a 30-day test set.\
+In all three figures, the largest residuals are associated with over-predictions by the models. This is most likely due to the large prevalence of low case count days.
+
+To assess the importance of the predictors, variable importance plots were generated for each model type.
+These plots can provide insight into how mobility predictors contributed to the predictive power of each model.
 
 ```{r}
-#| label: modeltable
-#| fig-cap: "Table 5.1: Model Performance"
+#| label: lassovip
+#| fig-cap: "Figure 5.12: Variable Importance Plot of Mobility Included LASSO Model"
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","models.png"))
+knitr::include_graphics(here("results","figures","vip_full_lasso.png"))
 ```
+
+```{r}
+#| label: rfvip
+#| fig-cap: "Figure 5.13: Variable Importance Plot of Mobility Included Random Forest Model"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","vip_full_rf.png"))
+```
+
+```{r}
+#| label: xgbvip
+#| fig-cap: "Figure 5.14: Variable Importance Plot of Mobility Included XGBoost Model"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","vip_full_boost.png"))
+```
+
+Across the three models, LASSO regression identified the mobility predictors as highly important, indicating a strong relationship between the mobility dynamics and case incidence.
+Random Forest indicated moderate importance to these predictors while XGBoost found the predictors to be insignificant.
+
+Additional VIP plots for the other predictor sets can be found in the supplementary materials for the analysis.
 
@@
 
-## Summary and Interpretation
+## Model Performance
 
-This study aimed to evaluate the predictive value of mobility dynamis on COVID-19 case incidence at the county level in Georgia using three different machine learning methods, LASSO, Random Forest, and XGBoost.
+This study aimed to investigate the extent to which mobility dynamics predict variation in COVID-19 case incidence at the county level in Georgia using three different machine learning methods, LASSO, Random Forest, and XGBoost.
+To assess practical predictive performance, RMSE and MAE were also reported
 
-Among all models, XGBoost with lagged mobility predictors achieved the best overall performance, with an RMSE of 37.9, an R-squared of 0.836, and MAE of 23.5.
-This model outperformed all models in RMSE and had great relative performance in R-squared and MAE, suggesting the improvement of predictive ability through incorporating time-lagged mobility trends.
-The improvement in RMSE and MAE in relation to the other models indicates a meaningful reduction in both large and average prediction errors when lagged mobility data is included.\
+Among all models, LASSO with mobility predictors achieved the best overall performance, with an RMSE of 46.5, an R² of 0.843, and MAE of 26.9.
+The inclusion of mobility predictors were able to improve the RMSE, MAE, and R² from the baseline model.
+In addition to improvements among the baseline LASSO model, this model outperformed all models in R² and had great relative performance in RMSE and MAE, suggesting that mobility predictors contributed meaningfully to the model's capacity to explain variance in COVID-19 cases while also improving predictive accuracy.
+LASSO identified mobility variables as key predictors, suggesting a linear relationship between mobility trends and COVID-19 case incidence.
+This is consistent with previous literature where linear regression models incorporating mobility data have been effective in case forecasting.
+
+Considerable improvement in MAE and RMSE were observed in mobility included models for Random Forest.
+Compared to baseline, the mobility included predictors had a MAE reduction of 21.72% and 10% reduction in RMSE.
+The lagged predictor set performed even better, with a 38.87% reduction in MAE and 26.5% reduction in RMSE.
+These predictors were assigned moderate importance, indicating that that both non-lagged and lagged mobility predictors contributed useful predictive value.
+However, the inclusion of mobility predictors resulted in only a slight improvement in R², suggesting that while these features reduced prediction errors, they did not substantially enhance the model's ability to explain the overall variance in COVID-19 case incidence.
+This may imply that the model's explanatory power mostly comes from the other variables present rather than mobility.
+Additionally, unlike in LASSO, mobility predictors were not dominant drivers of case trends.
+
+In contrast, XGBoost had strong baseline predictive performance but saw modest or negative improvement gains with the inclusion of mobility predictors.
+Additionally, investigating variable importance plots further, XGBoost tended to rate the mobility predictors as unimportant.
+XGBoost's use of "greedy" optimization can cause the modeling technique to overlook weak or redudant predictors, like mobility, especially if their effect is context-dependent or overshadowed by stronger features.
+The noisiness or redundancy found in the mobility  predictors may lead to its low importance.
 \
-The LASSO models also demonstrate strong predictive ability, with the mobility driven models having overall greater performance.
-As demonstrated in XGBoost, the lagged models have great performance, especially when compared to the baseline models.
 
-In contrast, the Random Forest models generally have the poorest performance in terms of RMSE, R-squared, and MAE.
-However, it should be noted that the lowest MAE in all models occurs in the Random Forest with lagged mobility variables.\
-\
-While mostly exploratory and preliminary, the results suggest that the inclusion of lagged mobility predictors can significantly improve model performance, specifically in XGboost.
-The lagged XGBoost model achieved great relative performance in RMSE and MAE while maintaining a high R-squared.
-This aligns with previous epidemiological findings suggesting that population movement precedes changes in infectious rates, often with a delay due to disease incubation and reporting lag.
-These findings highlight the importance of temporal feature engineering when using behavioral data such as mobility to predict case outcomes.
+While mostly exploratory, preliminary, and limited to this study's context, the results suggest that the inclusion of lagged mobility predictors can significantly reduce prediction error metrics such as MAE and RMSE, specifically in LASSO and Random Forest models.
+These reductions indicate improved short-term predictive accuracy, even when overall explanatory power does not significantly improve.
+With this increase in short term forecasting accuracy, public health agencies can track mobility shifts to anticipate potential case surges and take more proactive approaches for prevention, such as medical resource allocation or adjusting guidelines of interventions.
+However, the inclusion of mobility predictors seemingly did not significantly improve the R² of models, suggesting that these predictors do not improve the explanatory power of models.
+While mobility trends can be a useful input for predictive surveillance systems, they are not sufficient alone to explain complex patterns of COVID-19 transmission, reaffirming the need to include additional structural or behavioral predictors to achieve robust model performance.
+
+## Mobility Dynamics and Individualism
+
+The CMR's explanatory capacity is seemingly greater in non-American contexts.
+For example, in Jakarta, the CMR was able to explain 52% of the variation in case counts with just the mobility predictors [@nandaCommunityMobilityCOVID192022b].
+This could be the result of complex social and political contexts persisting in the US.
+Mobility trends may lack explanatory power in the US if they do not consistently follow or correlate with transmission.
+In other words, inconsistency in American behavior could limit the CMR's explanatory power.
+
+It is posited that the US has a "cultural default" of being more focused on individualism in which a person is prioritized over the needs of the group while some cultures, like those found in East Asia, value collectivism [@markusCulturalDefaultsTime2024b].
+@bazziRuggedIndividualismCollective2021b, has demonstrated that the unique sense of individualism, coined "rugged individualism" is a prominent feature of American culture that has led to hampered pandemic response, specifically, worse adherence to social distancing guidelines, mask usage, and a weaker local government effort to control the virus spread.
+This sense of individualism may have led to noisy and complex mobility patterns independent of COVID-19 case incidence.
+For instance, large spikes in cases may not affect mobility patterns due to this disregard of the severity of the pandemic.
+
+Investment into other data collection strategies in the midset of a pandemic may prove to be more effective in forecasting disease incidence.
+Additionally, the demonstration of American individualism during the pandemic may be an indication for public health agencies to begin focusing on developing prevention control methods that cater to U.S.
+culture.
 
@@
 First, the mobility data was incomplete, leading to the filtering of majority of Georgia counties out of the data set.
-This could be due to the methodology used for measuring which would skew the data towards more suburban and urban areas where phone usage is more common.
+This could be due to the methodology used for measuring which would skew the data towards more suburban and urban areas where Google product usage is more common.
 This can also mask behaviors related to communities on a more individual scale.
@@
 Lastly, machine learning models can be highly sensitive to parameter tuning and data pre-processing.
+In some instances, models like XGBoost and Random Forest are at risk of overfitting [@jhaPopularMachineLearning2024b].
 
@@
 
-This research establishes and explores the potential of incorporating mobility dynamics into public health surveillance, specifically, for COVID-19 case forecasting.
-Future research should explore the association further and consider the use of more individual behaviors, like adherence to masking or social distancing policies, to improve forecasting accuracy.
+This research explores the potential of incorporating mobility dynamics into public health surveillance, specifically, for COVID-19 case forecasting.
+The improvement of models, in the metrics of interest, suggest the added predictive value of mobility predictors, indicating that these variables can improve forecasting accuracy for COVID-19 case incidence.
+Future research should explore the integration of additional behavioral predictors, examine these findings in different geographic and sociopolitical contexts, and assess how mobility patterns interact with other public health interventions to influence disease transmission dynamics.
 

### 2025-05-01 project-review-files/project-review-Yang_files/project-review-Yang.qmd
Fixes to Part 6

---
title: Project Review Overall Score - 3
author: Guozheng Yang
date: 04/21/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Assessing the Added Predictive Power of Google’s Community Mobility Report for COVID-19 Forecasting on a County-Level

Name of project author(s): Vincent Nguyen

Name of project reviewer: Guozheng Yang

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

The background section is clearly written and well-referenced, providing a solid explanation of the COVID-19 timeline and the role of human mobility in infectious disease dynamics. The inclusion of public health response measures (e.g., lockdowns and behavioral shifts) sets the context well. The introduction of machine learning methods is appropriate and demonstrates a thoughtful application to the research question. I like how the author structured the introduction section into three sub-sections. The logic flow is smooth.

The project has a strong motivation, with a well-demonstrated rationale for integrating mobility data into prediction models. A stronger emphasis on what makes this work novel (e.g., the county-level focus in Georgia or comparisons across models) would enhance the framing.

### Summary assessment

* strong contextualization and motivation

## Question description

### Feedback and Comments

The research questions are clearly stated. The author aims to assess the additive value of mobility data for predicting COVID-19 incidence using machine learning models. The hypotheses are testable and directly tied to the contents taught in this class.

One small improvement would be to more explicitly define how the author identify the "best" model by assessing different model performance metrics. If the author wants to comprehensively judge the advantages and disadvantages of each model, it is sufficient to state it clearly in one or two sentences.

### Summary assessment

* question/hypotheses fully clear

## Data description

### Feedback and Comments

The data sources are described clearly, and relevant metadata (e.g., source names, time coverage, and inclusion criteria) are provided. It is good that the author also appropriately acknowledges limitations in the Google Mobility data set (e.g., lack of rural data). The process of filtering and constructing the final data set (44 counties with >=80% completeness) is described concisely.

One question: do the excluded 115 counties have geographical clustering patterns? Were they excluded just because they are in the rural area? It is totally fine if the author did not check that, I'm just curious.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis

### Feedback and Comments

The data wrangling and cleaning process is sufficiently explained, and the author has appropriately handled missingness by filtering out incomplete counties. The exploratory visualizations (e.g., time-series plots, density plots, mobility trends) are helpful and well-integrated with the narrative. The correlation analysis supports model feature engineering.

One minor improvement would be to explain the mobility indicators excluded due to missing data in more detail, and whether any imputation strategies were considered. Also, it would be great if the author could demonstrate why the rolling windows were set to 120 days / 30 days. Is this validated by previous studies or just suitable for this data set?

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments

The modeling approach is well thought out, using LASSO, Random Forest, and XGBoost. The justification for avoiding traditional time series models (due to non-normality and non-stationarity) is sound. The use of cross-validation with rolling windows is a strength and reflects good understanding of time-series prediction challenges.

The evaluation metrics (RMSE, MAE, R-square) are appropriate, and the author shows good command of tuning and validation. One small suggestion: the author may want to comment on why RMSE was prioritized over MAE for final model ranking, or present both equally.

### Summary assessment

* strong and reasonable analysis

## Presentation

### Feedback and Comments

Figures and tables are generally clean and well-labeled. The performance comparison is clearly presented. However, the author could add more details in figure captions for standalone interpretation. Also, figures in the 5.2 and 5.3 sub-sections do not have the indexing number (e.g., Figure X.X).

Overall, visual materials support the narrative well.

### Summary assessment

* results are very well presented

## Discussion/Conclusions

### Feedback and Comments

The discussion effectively summarizes results and reflects on model performance across the different feature sets and algorithms. The interpretation of lagged mobility predictors improving forecasting accuracy is logical and linked to incubation periods and reporting lag. Limitations are well acknowledged as well, especially regarding data completeness and potential confounding. The suggestion for future work (e.g., integrating policy or behavioral variables) is on point.

A minor suggestion is to refer to existing literature to discuss the consistent and inconsistent findings of this study. The author mentioned "This aligns with previous epidemiological findings suggesting that population movement precedes changes in infectious rates, often with a delay due to disease incubation and reporting lag." but no paper is cited. The author could also elaborate slightly more on public health implications (e.g., how mobility data could support real-time response or alert systems).

### Summary assessment

* strong, complete and clear discussion

## Further comments

Excellent work overall. The cross-validation design and feature engineering with time-lagged mobility predictors are well executed. The author clearly demonstrates an understanding of the modeling process and thoughtfully discusses limitations. But small improvements are needed. For example, a code chunk accidentally appears in the manuscript.

# Overall project content evaluation

## Structure

### Feedback and Comments

The folder and file structure appear straightforward. File names are intuitive, and no unnecessary files are included. I also like the detailed instructions provided in each README file. I could easily follow the workflow in this project.

It would be better if the author could provide more instructions in the README file in the analysis-code folder.

### Summary assessment

* well structured

## Documentation

### Feedback and Comments

The author retained the structural framework of the template folders. Overall, the project folder is well-organized: clean, concise, and free of redundant files. Each executable file contains a necessary README file in the same folder, which greatly enhances usability for readers. Additionally, the executable files are sufficiently annotated to explain each step, further demonstrating the author's attention to clarity and reproducibility.

One minor suggestion: the poster folder seems redundant.

### Summary assessment

* fully and well documented

## Reproducibility

### Feedback and Comments

The author put great effort in coding. The code for Random Forest got my laptop frozen, so I decided to shrink the range of for parameter tuning to quickly check the reproducibility. Overall, the analysis appears fully reproducible. Great job!

### Summary assessment

* fully reproducible without issues

## Thoroughness

### Feedback and Comments

The study demonstrates a strong level of thoroughness in modeling and data analysis. I like it that the author examines feature sets thoughtfully and integrates epidemiological reasoning into the modeling choices. The limitations and future directions are also clearly laid out in the discussion section. It is a very complete course project.

### Summary assessment

* strong level of thorougness

## Further comments

This is a well-managed course project. The analysis is presented in a clear and professional manner. The comparisons across models are insightful, and covers enough contents from this course.

### 2025-05-02 code/analysis-code/statisitical-analysis.qmd
Edits

@@
 # Reformat these data frames for joining
-
 base_lasso_metrics <- all_base_lasso_metrics %>% 
@@
 
-```{r}
+This code chunk creates a short summary table of the results from the model_metrics data frame.
+
+```{r}
+# Create table and also reorder data frame columns
 model_table <- model_metrics %>%
@@
 ```
-
-This code chunk creates a short summary table of the results from the model_metrics data frame.
-```{r}
-# Create gt table
-model_table <- model_metrics %>%
-  gt() %>%
-  tab_header(
-    title = "Model Performance Metrics on Data",
-    subtitle = "Best values are highlighted per metric"
-  ) %>%
-  fmt_number(columns = c(rmse, mae), decimals = 1) %>%
-  fmt_number(columns = rsq, decimals = 3) %>%
-  cols_label(
-    model = "Model",
-    rmse = "RMSE",
-    rsq = "R²",
-    mae = "MAE"
-  ) %>%
-
-  # Highlight lowest RMSE
-  tab_style(
-    style = cell_fill(color = "lightgreen"),
-    locations = cells_body(
-      columns = rmse,
-      rows = rmse == min(rmse, na.rm = TRUE)
-    )
-  ) %>%
-
-  # Highlight lowest MAE
-  tab_style(
-    style = cell_fill(color = "lightgreen"),
-    locations = cells_body(
-      columns = mae,
-      rows = mae == min(mae, na.rm = TRUE)
-    )
-  ) %>%
-
-  # Highlight highest R²
-  tab_style(
-    style = cell_fill(color = "lightgreen"),
-    locations = cells_body(
-      columns = rsq,
-      rows = rsq == max(rsq, na.rm = TRUE)
-    )
-  ) %>%
-
-  tab_style(
-    style = cell_text(weight = "bold"),
-    locations = cells_column_labels(everything())
-  ) %>%
-  tab_options(
-    table.width = pct(100),
-    table.align = "left",
-    column_labels.padding = px(10),
-    data_row.padding = px(6)
-  )
-
-print(model_table)
-
-gtsave(model_table, here("results", "figures", "models.png"))
-```
